---
title: "Model Tuning and Evaluation"
author: "Group 5"
date: today
format:
  html:
    toc: true
    code-fold: show
    df-print: kable
    embed-resources: true
  pdf:
    toc: true
    df-print: kable
execute:
  warning: false
  message: false
---

# Initial Setup

## Run this if you need the packages

```{r}
#| label: Install Packages
#| eval: false
# install.packages(c("tidyverse", "survey", "rpart", "rpart.plot", "ranger", "pROC", "caret"))
```

## Loading the packages we need

```{r}
#| label: Package Setup

library(tidyverse)  
library(survey)     # survey-weighted regressions, likelihood-ratio test
library(rpart)      
library(rpart.plot) 
library(ranger)     # random forest
library(pROC)       # ROC curves, AUC with confidence intervals, DeLong test
library(caret)
```

## File paths

```{r}
#| label: File-Path Declarations

analytical_directory <- "../Datasets/Analytical/"
model_directory <- "../Models/"
```

# 1. Evaluating Models

## 1.1 What is being compared

| Question | Models | Task |
|------------------------|------------------------|------------------------|
| Employment | Survey-weighted logistic regression, decision tree, random forest, and a demographics-only baseline | Binary classification of employed |
| Income | Survey-weighted linear regression, random forest, and a demographics-only baseline | Regression of log_salary |

The **baseline models** are what we create in this file. The Milestone 2 success criteria required the models to be compared with a demographics only baseline (a model that knows a person's age, sex, population group, province and settlement type), but nothing about their education, digital access or household. If the full models do not predict better than the baseline we create in this section, education and digital access add little predictive information beyond who and where a person is.

## 1.2 Metrics and why they were chosen

We can't predict based off of a single metric, therefore each task gets several that measure different things.

**Employment (classification).** The models predict the probability of being **Employed**, so Employed is the positive class. About 35% of the labour force is unemployed, so a model that always predicts Employed would already be 65% accurate. When you take this into account you realise accuracy on its own is therefore misleading.

| Metric | What it measures | Why it is used |
|------------------------|------------------------|------------------------|
| ROC-AUC | How well the predicted probabilities rank employed people compared to unemployed people. | It is the main tuning and comparison metric: it does not depend on the threshold or on the class balance. |
| Balanced accuracy | The average of sensitivity and specificity. | Treats both classes as equally important, unlike accuracy. |
| Sensitivity (recall) | Share of employed people predicted Employed. | To correctly identify the number of employed people. |
| Specificity | Share of unemployed people that get predicted as unemployed. | To correctly identify the number of unemployed people. |
| Precision, F1 | Precision is the share of Employed predictions that are correct. F1 is the balance of precision and sensitivity. | These are used as additional metrics to support our decision making efforts. |
| Accuracy | Share of all predictions that are correct. | Reported for completeness, next to the 65% of the "always Employed" rule. |
| Brier score | Mean squared difference between predicted probability and the actual outcome (0 = perfect). | The Shiny app will show probabilities, so they must be accurate, not only well ranked. |

**Income (regression).** Errors are measured on the log scale, where the model was fitted:

| Metric | What it measures                                          |
|--------|-----------------------------------------------------------|
| RMSE   | Typical size of an error, with large errors counting more |
| MAE    | Average size of an error                                  |
| R²     | Share of the variation in `log_salary` the model explains |

The residuals of the linear regression are also plotted, because its standard errors rely on them being roughly symmetric with constant spread.

## 1.3 Helper functions

These small functions calculate the metrics above, thus they are written once here so that the cross-validation, the test set and the subgroup checks all use exactly the same calculation.

```{r}
#| label: Helper Functions

# We use k-fold cross-validation, the data is split into k separate, roughly equal parts, and each part is called a fold. Every row belongs to exactly one fold
cross_validate <- function(data, fold_id, fit_and_predict) {
  predictions <- rep(NA_real_, nrow(data))
  
  for (k in sort(unique(fold_id))) {
    held_out <- fold_id == k
    predictions[held_out] <- fit_and_predict(data[!held_out, ], data[held_out, ])
  }
  
  predictions
}

# ROC curve and AUC with Employed as the positive class
roc_curve <- function(truth, probability) {
  roc(truth, 
      probability, 
      levels = c("Unemployed", "Employed"), 
      direction = "<", quiet = TRUE)
}

auc_value <- function(truth, probability) {
  as.numeric(
    auc(
      roc_curve(truth, probability)
      )
    )
}

# AUC for each cross-validation
fold_auc <- function(truth, probability, fold_id) {
  sapply(sort(unique(fold_id)), 
         function(k) auc_value(truth[fold_id == k], probability[fold_id == k])
         )
}

# Population weighted AUC
weighted_auc <- function(truth, probability, weight) {
  employed <- truth == "Employed"
  
  wins <- outer(probability[employed], probability[!employed], ">") +
    0.5 * outer(probability[employed], probability[!employed], "==")
  
  sum(weight[employed] * (wins %*% weight[!employed])) /
    (sum(weight[employed]) * sum(weight[!employed]))
}

# Threshold based classification metrics
classification_metrics <- function(truth, probability, threshold, weight = rep(1, length(truth))) {
  predicted_employed <- probability >= threshold
  actual_employed <- truth == "Employed"
  
  tp <- sum(weight[predicted_employed & actual_employed])
  tn <- sum(weight[!predicted_employed & !actual_employed])
  fp <- sum(weight[predicted_employed & !actual_employed])
  fn <- sum(weight[!predicted_employed & actual_employed])
  
  sensitivity <- tp / (tp + fn)
  specificity <- tn / (tn + fp)
  precision <- tp / (tp + fp)
  
  tibble(
    accuracy = (tp + tn) / (tp + tn + fp + fn),
    balanced_accuracy = (sensitivity + specificity) / 2,
    sensitivity = sensitivity,
    specificity = specificity,
    precision = precision,
    f1 = 2 * precision * sensitivity / (precision + sensitivity),
    brier = sum(weight * (probability - actual_employed)^2) / sum(weight)
  )
}

# Regression metrics on the log salary scale
regression_metrics <- function(actual, predicted, weight = rep(1, length(actual))) {
  error <- actual - predicted
  tibble(
    rmse = sqrt(sum(weight * error^2) / sum(weight)),
    mae = sum(weight * abs(error)) / sum(weight),
    r_squared = 1 - sum(weight * error^2) /
      sum(weight * (actual - weighted.mean(actual, weight))^2)
  )
}
```

# 2. Training data

The training files are read through the same preparation functions as in the modelling notebooks (`Scripts/R/model_preparation.R`), so the rows, columns and factor levels are identical to those the saved models were fitted on.

```{r}
#| label: Read Training Data

source("R/model_preparation.R")

model_train <- prepare_employment_data(file.path(analytical_directory, "model_train.csv"))
income_train <- prepare_income_data(file.path(analytical_directory, "income_train.csv"))

tibble(
  dataset = c("model_train", "income_train"),
  rows = c(nrow(model_train), nrow(income_train)),
  households = c(
    n_distinct(model_train$household_id), 
    n_distinct(income_train$household_id)
    )
)
```

# 3. Household-grouped cross-validation folds

## 3.1 Why the folds are grouped by household

Tuning a model requires data it was not fitted on, but the test set has to stay untouched until the final evaluation. Cross-validation solves this using only the training data, where the training rows are split into five folds (subsets or partitions of our data) and each model is fitted five times.

The folds are grouped by household for the same reason as the Milestone 2 train/test split. People in the same household share every household level predictor (internet access, computer, smartphones, household size). If housemates were in different folds, the model would be evaluated on people almost identical to ones it was trained on, and the cross-validated performance would look better than it really is. `caret::groupKFold()` prevents this by putting every member of a household in the same fold.

```{r}
#| label: Create Folds

set.seed(67) 


employment_folds <- caret::groupKFold(model_train$household_id, k = 5)

employment_fold_id <- rep(NA_integer_, nrow(model_train))
for (k in seq_along(employment_folds)) {
  employment_fold_id[-employment_folds[[k]]] <- k
}

income_folds <- caret::groupKFold(income_train$household_id, k = 5)

income_fold_id <- rep(NA_integer_, nrow(income_train))
for (k in seq_along(income_folds)) {
  income_fold_id[-income_folds[[k]]] <- k
}
```

## 3.2 Checking the folds

```{r}
#| label: Fold Verification

tibble(
  check = c(
    "Employment: rows without a fold",
    "Employment: households split across folds",
    "Income: rows without a fold",
    "Income: households split across folds"
  ),
  value = c(
    sum(is.na(employment_fold_id)),
    tibble(
      household_id = model_train$household_id, 
      fold = employment_fold_id
      ) |>
      group_by(household_id) |> 
      summarise(folds = n_distinct(fold)) |> 
      filter(folds > 1) |> 
      nrow(),
    sum(is.na(income_fold_id)),
    tibble(
      household_id = income_train$household_id, 
      fold = income_fold_id) |>
      group_by(household_id) |> 
      
      summarise(
        folds = n_distinct(fold)) |> 
      filter(folds > 1) |> 
      nrow()
  )
)

model_train |>
  mutate(fold = employment_fold_id) |>
  group_by(fold) |>
  summarise(
    people = n(), 
    households = n_distinct(household_id),
    share_employed = round(mean(employed == "Employed"), 3)
    )
```

Every row is in exactly one fold, no household is split across folds and the share of employed people is similar in every fold, so each fold is a fair small-scale copy of the training data.

# 4. Employment models: cross-validation and tuning

## 4.1 Logistic regression and the baseline

```{r}
#| label: Baseline Formula

# Demographics-only baseline
employment_baseline_formula <- employed ~ age_centred + age_centred_squared + female + population_group + province + settlement_type
```

```{r}
#| label: Cross-Validate Logistic and Baseline

# oof stands for "out of fold" and it is used for detecting overfitting, optimizing model hyperparameters identifying base models for ensemble learning and ensuring the reliability and accuracy of machine learning models


logistic_oof <- cross_validate(
  model_train, 
  employment_fold_id, 
  function(training, held_out) {
  design <- svydesign(
    ids = ~household_id, weights = ~person_weight, data = training
    )
  
  model <- svyglm(
    employment_regression_formula, 
    design = design, 
    family = quasibinomial()
    )
  
  as.numeric(
    predict(model, held_out, type = "response")
    )
  }
)

baseline_oof <- cross_validate(
  model_train, 
  employment_fold_id, 
  function(training, held_out) {
  design <- svydesign(
    ids = ~household_id, weights = ~person_weight, data = training
    )
  
  model <- svyglm(
    employment_baseline_formula, 
    design = design, 
    family = quasibinomial()
    )
  
  as.numeric(
    predict(model, held_out, type = "response")
    )
  }
)

tibble(
  model = c("Baseline (demographics only)", "Logistic regression"),
  mean_cv_auc = c(
    mean(fold_auc(
      model_train$employed, 
      baseline_oof, 
      employment_fold_id
      )
    ),
    mean(fold_auc(
      model_train$employed, 
      logistic_oof, 
      employment_fold_id
      )
    )
  ),
  sd_cv_auc = c(
    sd(fold_auc(
      model_train$employed, 
      baseline_oof, 
      employment_fold_id
        )
      ),
  sd(fold_auc(
    model_train$employed,
    logistic_oof,
    employment_fold_id
      )
    )
  )
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 4.2 Decision tree tuning

| Parameter   | Values tried                |
|-------------|-----------------------------|
| `maxdepth`  | 3, 4, 5                     |
| `minbucket` | 50, 100, 200                |
| `cp`        | 0.0005, 0.001, 0.002, 0.005 |

```{r}
#| label: Tune Decision Tree

tree_grid <- expand_grid(
  maxdepth = c(3, 4, 5),
  minbucket = c(50, 100, 200),
  cp = c(0.0005, 0.001, 0.002, 0.005)
)

tree_results <- tree_grid |> 
  mutate(
    mean_cv_auc = NA_real_, 
    sd_cv_auc = NA_real_, 
    leaves = NA_integer_
    )

tree_oof <- list()

for (i in seq_len(nrow(tree_grid))) {
  settings <- tree_grid[i, ]
  
  tree_control <- rpart.control(
    maxdepth = settings$maxdepth, 
    minbucket = settings$minbucket,
    cp = settings$cp, xval = 0
    )

  predictions <- cross_validate(
    model_train, 
    employment_fold_id, 
    function(training, held_out) {
      tree <- rpart(
        employment_tree_formula, 
        data = training, 
        method = "class", 
        control = tree_control
        )
   
      predict(tree, held_out, type = "prob")[, "Employed"]
    }
  )

  aucs <- fold_auc(
    model_train$employed, 
    predictions, 
    employment_fold_id
    )
  
  tree_results$mean_cv_auc[i] <- mean(aucs)
  tree_results$sd_cv_auc[i] <- sd(aucs)

  full_tree <- rpart(
    employment_tree_formula,
    data = model_train, 
    method = "class", 
    control = tree_control
    )
  
  tree_results$leaves[i] <- sum(full_tree$frame$var == "<leaf>")
  tree_oof[[i]] <- predictions
}

tree_results |>
  arrange(desc(mean_cv_auc)) |>
  mutate(across(c(mean_cv_auc, sd_cv_auc), \(x) round(x, 4))) |>
  head(10)
```

**Selection rule: the one-standard-error rule.** Many settings give almost the same AUC, and the differences between them are smaller than the variation between folds. Instead of taking the single highest AUC, the tree with the fewest leaves is chosen among all settings whose mean AUC is within one standard error of the best.

```{r}
#| label: Select Tree Settings

best_tree_row <- which.max(tree_results$mean_cv_auc)
one_se_limit <- tree_results$mean_cv_auc[best_tree_row] -
  tree_results$sd_cv_auc[best_tree_row] / sqrt(5)

tree_selected <- tree_results |>
  mutate(row = row_number()) |>
  filter(mean_cv_auc >= one_se_limit) |>
  arrange(leaves, desc(mean_cv_auc)) |>
  slice(1)

tibble(
  measure = c(
    "Best mean CV AUC",
    "One-standard-error limit", 
    "Selected maxdepth",
    "Selected minbucket", 
    "Selected cp", 
    "Selected mean CV AUC", 
    "Leaves in selected tree"
    ),
  
  value = as.character(
    c(
      round(tree_results$mean_cv_auc[best_tree_row], 4),
      round(one_se_limit, 4),
      tree_selected$maxdepth, 
      tree_selected$minbucket, 
      tree_selected$cp,
      round(tree_selected$mean_cv_auc, 4), tree_selected$leaves
      )
    )
)

tree_selected_oof <- tree_oof[[tree_selected$row]]
```

Every tree's AUC lies in a narrow band, and the best trees all use the maximum depth of 5, so the depth limit set for readability does cost some accuracy.

## 4.3 Random forest tuning

Random Forest's main parameters are "mtry" (how many predictors each split may choose from) and "min.node.size" (the smallest leaf in each tree). "num.trees" is not tuned: more trees only make the predictions more stable.

| Parameter | Values tried | Effect |
|------------------------|------------------------|------------------------|
| `mtry` | 2, 3, 4, 6 | Small values make the trees more different from each other;, while large values let each split use the strongest predictors. |
| `min.node.size` | 10, 30, 100, 200, 400 | Larger leaves give smoother, often better-calibrated probabilities, while smaller leaves fit more detail. |

Random Forest is chosen by the highest mean cross-validated AUC. The one-standard-error rule is not used here, because it is not meant to be readable.

```{r}
#| label: Tune Random Forest

forest_grid <- expand_grid(
  mtry = c(2, 3, 4, 6),
  min.node.size = c(10, 30, 100, 200, 400)
)

forest_results <- forest_grid |> 
  mutate(mean_cv_auc = NA_real_, sd_cv_auc = NA_real_)

forest_oof <- list()

for (i in seq_len(nrow(forest_grid))) {
  settings <- forest_grid[i, ]

  predictions <- cross_validate(
    model_train, 
    employment_fold_id,
    function(training, held_out) {
    forest <- ranger(
      employment_tree_formula, 
      data = training,
      num.trees = 300, 
      mtry = settings$mtry, 
      min.node.size = settings$min.node.size,
      probability = TRUE, 
      respect.unordered.factors = "order", 
      seed = 67
    )
    
    predict(forest, held_out)$predictions[, "Employed"]
  })

  aucs <- fold_auc(
    model_train$employed, 
    predictions, 
    employment_fold_id
    
    )
  forest_results$mean_cv_auc[i] <- mean(aucs)
  forest_results$sd_cv_auc[i] <- sd(aucs)
  forest_oof[[i]] <- predictions
}

forest_results |>
  arrange(desc(mean_cv_auc)) |>
  mutate(across(c(mean_cv_auc, sd_cv_auc), \(x) round(x, 4)))

forest_selected <- forest_results |>
  mutate(row = row_number()) |>
  slice_max(mean_cv_auc, n = 1, with_ties = FALSE)

forest_selected_oof <- forest_oof[[forest_selected$row]]
```

## 4.4 Choosing the classification thresholds

A model's probabilities become Employed/Unemployed predictions through a threshold. The default of 0.5 is only right when both kinds of error are equally costly *and* the classes are balanced. Here 65% of people are employed, so a 0.5 threshold labels most people Employed and misses many of the unemployed.

The threshold for each model is chosen from its **out-of-fold** predictions, as the threshold that maximises **balanced accuracy**. This gives the Employed and Unemployed classes equal importance (miinimises skewness).

```{r}
#| label: Choose Thresholds

youden_threshold <- function(truth, probability) {
  coords(
    roc_curve(truth, probability), 
    "best", 
    best.method = "youden",
    ret = "threshold", transpose = FALSE
  )$threshold[1]
}

employment_thresholds <- tibble(
  model = c("Baseline", "Logistic regression", "Decision tree", "Random forest"),
  threshold = c(
    youden_threshold(model_train$employed, baseline_oof),
    youden_threshold(model_train$employed, logistic_oof),
    youden_threshold(model_train$employed, tree_selected_oof),
    youden_threshold(model_train$employed, forest_selected_oof)
  )
)

employment_thresholds |> mutate(threshold = round(threshold, 3))
```

The effect of the tuned threshold compared with 0.5, on the out-of-fold predictions:

```{r}
#| label: Threshold Comparison

oof_predictions <- list(
  "Baseline" = baseline_oof,
  "Logistic regression" = logistic_oof,
  "Decision tree" = tree_selected_oof,
  "Random forest" = forest_selected_oof
)

bind_rows(
  lapply(
    names(oof_predictions), 
    function(m) {
      bind_rows(
        classification_metrics(model_train$employed, oof_predictions[[m]], 0.5) |>
        mutate(model = m, threshold = "0.5", .before = 1),
        
        classification_metrics(
          model_train$employed, 
          oof_predictions[[m]],
          employment_thresholds$threshold[employment_thresholds$model == m]
          ) |>
        mutate(model = m, threshold = "Tuned", .before = 1)
      )
    }
  )
) |>
  select(model, threshold, balanced_accuracy, sensitivity, specificity, accuracy) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

At 0.5, every model identifies fewer than half of the unemployed people (specificity below 0.5) while labelling most employed people correctly. The tuned thresholds trade some of that sensitivity for much higher specificity, which raises balanced accuracy for every model. Accuracy falls slightly, because the larger class (Employed) is now predicted less often, which was expected and is why accuracy is not used to choose the threshold.

# 5. Income models: cross-validation and tuning

## 5.1 Linear regression and the baselines

```{r}
#| label: Cross-Validate Income Regressions

income_baseline_formula <- log_salary ~ age_centred + age_centred_squared + female +
  population_group + province + settlement_type

linear_oof <- cross_validate(
  income_train, 
  income_fold_id, 
  function(training, held_out) {
    design <- svydesign(
      ids = ~household_id, weights = ~person_weight, data = training
      )
    
    model <- svyglm(
      income_regression_formula, design = design, family = gaussian()
      )
    
    as.numeric(predict(model, held_out))
    }
  )

income_baseline_oof <- cross_validate(
  income_train, 
  income_fold_id, 
  function(training, held_out) {
    design <- svydesign(
      ids = ~household_id, weights = ~person_weight, data = training
      )
    
    model <- svyglm(
      income_baseline_formula, design = design, family = gaussian()
      )
    
    as.numeric(predict(model, held_out))
    }
  )

mean_oof <- cross_validate(
  income_train, 
  income_fold_id,
  function(training, held_out) {
    rep(
      weighted.mean(training$log_salary, training$person_weight), 
      nrow(held_out)
      )
    }
  )

bind_rows(
  regression_metrics(income_train$log_salary, mean_oof) |> 
    mutate(model = "Training average", .before = 1),
  
  regression_metrics(income_train$log_salary, income_baseline_oof) |> 
    mutate(model = "Baseline (demographics only)", .before = 1),
  
  regression_metrics(income_train$log_salary, linear_oof) |> 
    mutate(model = "Linear regression", .before = 1)
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 5.2 Random forest tuning

| Parameter       | Values tried        |
|-----------------|---------------------|
| `mtry`          | 1, 2, 3, 4, 6, 8    |
| `min.node.size` | 5, 20, 50, 100, 200 |

The setting with the lowest mean cross-validated RMSE is chosen.

```{r}
#| label: Tune Income Forest

income_forest_grid <- expand_grid(
  mtry = c(1, 2, 3, 4, 6, 8),
  min.node.size = c(5, 20, 50, 100, 200)
)

income_forest_results <- income_forest_grid |> 
  mutate(mean_cv_rmse = NA_real_, sd_cv_rmse = NA_real_)

income_forest_oof <- list()

for (i in seq_len(nrow(income_forest_grid))) {
  settings <- income_forest_grid[i, ]

  predictions <- cross_validate(
    income_train, 
    income_fold_id, 
    function(training, held_out) {
      forest <- ranger(
        income_tree_formula, 
        data = training,
        num.trees = 300, mtry = settings$mtry, 
        min.node.size = settings$min.node.size,
        respect.unordered.factors = "order", 
        seed = 67
      )
      
      predict(forest, held_out)$predictions
      }
    )

  fold_rmse <- sapply(1:5, function(k) {
    regression_metrics(
      income_train$log_salary[income_fold_id == k], 
      predictions[income_fold_id == k]
      )$rmse
  })
  
  income_forest_results$mean_cv_rmse[i] <- mean(fold_rmse)
  income_forest_results$sd_cv_rmse[i] <- sd(fold_rmse)
  income_forest_oof[[i]] <- predictions
}

income_forest_results |>
  arrange(mean_cv_rmse) |>
  mutate(across(c(mean_cv_rmse, sd_cv_rmse), \(x) round(x, 4))) |>
  head(10)

income_forest_selected <- income_forest_results |>
  mutate(row = row_number()) |>
  slice_min(mean_cv_rmse, n = 1, with_ties = FALSE)

income_forest_selected_oof <- income_forest_oof[[income_forest_selected$row]]
```

# 6. Refitting and saving the tuned models

The chosen settings are now used to refit each tuned model on the full training file.

```{r}
#| label: Refit Tuned Models

employment_tree_tuned <- rpart(
  employment_tree_formula,
  data = model_train,
  method = "class",
  control = rpart.control(
    maxdepth = tree_selected$maxdepth, 
    minbucket = tree_selected$minbucket,
    cp = tree_selected$cp, 
    xval = 0
    )
)

employment_forest_tuned <- ranger(
  employment_tree_formula, 
  data = model_train,
  num.trees = 500, 
  mtry = forest_selected$mtry, 
  min.node.size = forest_selected$min.node.size,
  probability = TRUE, 
  respect.unordered.factors = "order", 
  importance = "permutation", 
  seed = 67
)

income_forest_tuned <- ranger(
  income_tree_formula, 
  data = income_train,
  num.trees = 500,
  mtry = income_forest_selected$mtry, 
  min.node.size = income_forest_selected$min.node.size,
  respect.unordered.factors = "order", 
  importance = "permutation",
  seed = 67
)

employment_baseline <- svyglm(
  employment_baseline_formula,
  design = svydesign(
    ids = ~household_id, weights = ~person_weight, data = model_train
    ),
  family = quasibinomial()
)

income_baseline <- svyglm(
  income_baseline_formula,
  design = svydesign(
    ids = ~household_id, weights = ~person_weight, data = income_train
    ),
  family = gaussian()
)

# Models that were not tuned
employment_logistic <- readRDS(file.path(model_directory, "employment_logistic.rds"))

income_linear <- readRDS(file.path(model_directory, "income_linear.rds"))
```

```{r}
#| label: Save Tuned Models

saveRDS(employment_tree_tuned, file.path(model_directory, "employment_tree_tuned.rds"))
saveRDS(employment_forest_tuned, file.path(model_directory, "employment_forest_tuned.rds"))
saveRDS(employment_baseline, file.path(model_directory, "employment_baseline.rds"))
saveRDS(income_forest_tuned, file.path(model_directory, "income_forest_tuned.rds"))
saveRDS(income_baseline, file.path(model_directory, "income_baseline.rds"))

# The thresholds are part of each classification model and are needed by the Shiny app
write_csv(employment_thresholds, file.path(model_directory, "employment_thresholds.csv"))

tibble(
  setting = c(
    "Tree maxdepth",
    "Tree minbucket",
    "Tree cp",
    "Employment forest mtry", 
    "Employment forest min.node.size",
    "Income forest mtry", "Income forest min.node.size"
    ),
  
  starting_value = as.character(
    c(5, 100, 0.001, 3, 10, 2, 5)
    ),
  
  tuned_value = as.character(
    c(
      tree_selected$maxdepth, 
      tree_selected$minbucket,
      tree_selected$cp,
      forest_selected$mtry, 
      forest_selected$min.node.size,
      income_forest_selected$mtry, 
      income_forest_selected$min.node.size
      )
    )
)
```

The tuned tree:

```{r}
#| label: Plot Tuned Tree
#| fig-width: 12
#| fig-height: 7

rpart.plot(
  employment_tree_tuned,
  extra = 106,
  box.palette = "RdGn",
  branch.lty = 3,
  tweak = 1.1,
  main = "Tuned decision tree: probability of being employed"
)
```

The tuned tree splits only on age, smartphone access, computer ownership, household size and population group. It never splits on "matric_plus" or "any_home_internet", so it cannot show an education × internet interaction.

# 7. Test-set evaluation

## 7.1 Reading the test sets

The test sets are read with the same preparation functions, and the checks confirm they share no households with the training data.

```{r}
#| label: Read Test Data

model_test <- prepare_employment_data(file.path(analytical_directory, "model_test.csv"))
income_test <- prepare_income_data(file.path(analytical_directory, "income_test.csv"))

tibble(
  check = c(
    "Employment test rows",
    "Employment test households",
    "Employment households also in training", 
    "Income test rows",
    "Income test households", 
    "Income households also in training"
    ),
  
  value = c(
    nrow(model_test), 
    n_distinct(model_test$household_id),
    length(
      intersect(
        model_test$household_id, 
        model_train$household_id
        )
      ),
    nrow(income_test), 
    n_distinct(income_test$household_id),
    length(
      intersect(
        income_test$household_id, 
        income_train$household_id
        )
      )
    )
)

model_test |> 
  count(employed) |> 
  mutate(share = round(n / sum(n), 3))
```

## 7.2 Employment: test-set predictions and metrics

```{r}
#| label: Employment Test Predictions

employment_test_predictions <- model_test |>
  select(
    household_id,
    person_weight,
    employed, 
    population_group, 
    female, 
    settlement_type, 
    province
    ) |>
  mutate(
    "Baseline" = as.numeric(
      predict(employment_baseline, model_test, type = "response")
      ),
    
    "Logistic regression" = as.numeric(
      predict(employment_logistic, model_test, type = "response")
      ),
    
    "Decision tree" = 
      predict(employment_tree_tuned, model_test, type = "prob")[, "Employed"],
    
    "Random forest" = 
      predict(employment_forest_tuned, model_test)$predictions[, "Employed"]
  )

employment_models <- c("Baseline", "Logistic regression", "Decision tree", "Random forest")
```

```{r}
#| label: Employment Test Metrics

employment_test_metrics <- bind_rows(
  lapply(
    employment_models,
    function(m) {
      probability <- employment_test_predictions[[m]]
      
      threshold <- employment_thresholds$threshold[employment_thresholds$model == m]
      
      auc_ci <- ci.auc(roc_curve(employment_test_predictions$employed, probability))
  
      bind_rows(
        classification_metrics(
          employment_test_predictions$employed, 
          probability,
          threshold
          ) |>
        mutate(
          model = m, 
          weighting = "Unweighted",
          auc = auc_ci[2],
          auc_lower = auc_ci[1],
          auc_upper = auc_ci[3], 
          .before = 1
          ),
    
        classification_metrics(
          employment_test_predictions$employed, 
          probability, 
          threshold,
          employment_test_predictions$person_weight
          ) |>
          mutate(
            model = m,
            weighting = "Population-weighted",
            auc = weighted_auc(
              employment_test_predictions$employed, 
              probability,
              employment_test_predictions$person_weight
              ),
            auc_lower = NA, 
            auc_upper = NA, 
            .before = 1
            )
        )
      }
    )
  )

employment_test_metrics |>
  filter(weighting == "Unweighted") |>
  select(-weighting) |>
  pivot_longer(-model, names_to = "metric") |>
  pivot_wider(names_from = model, values_from = value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

The test results are close to the cross-validated results we previously evaluated, so the cross-validation gave a realistic estimate and the models are not overfitted to the training data. Every model clears the AUC ≥ 0.70 threshold, including the demographics-only baseline, so that threshold on its own does not separate the models. The comparison with the baseline is what shows whether education, digital access and the household predictors add anything.

The same metrics weighted by "person_weight", which estimate performance for the population rather than the sample:

```{r}
#| label: Employment Weighted Test Metrics

employment_test_metrics |>
  filter(weighting == "Population-weighted") |>
  select(-weighting, -auc_lower, -auc_upper) |>
  pivot_longer(-model, names_to = "metric") |>
  pivot_wider(names_from = model, values_from = value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 7.3 Confusion matrices

Each model's test predictions at its tuned threshold, as counts of people:

```{r}
#| label: Confusion Matrices

bind_rows(
  lapply(
    employment_models,
    function(m) {
      threshold <- employment_thresholds$threshold[employment_thresholds$model == m]
  
      tibble(
        model = m,
        actual = employment_test_predictions$employed,
        predicted = if_else(
          employment_test_predictions[[m]] >= threshold, "Employed", "Unemployed"
          )
        )
      }
    )
  ) |>
  count(model, actual, predicted) |>
  pivot_wider(
    names_from = predicted, 
    values_from = n, 
    names_prefix = "predicted_"
    ) |>
  arrange(
    factor(model, levels = employment_models), 
    actual
    )
```

## 7.4 ROC curves and statistical comparison

```{r}
#| label: ROC Curves
#| fig-width: 7
#| fig-height: 6

employment_rocs <- lapply(
  employment_models, 
  function(m) {
    roc_curve(
      employment_test_predictions$employed, 
      employment_test_predictions[[m]]
      )
    }
  )

names(employment_rocs) <- employment_models

ggroc(employment_rocs, legacy.axes = TRUE) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  labs(
    title = "ROC curves on the test set",
    x = "1 − specificity",
    y = "Sensitivity",
    colour = "Model"
    ) +
  coord_equal() +
  theme_minimal()
```

The DeLong test checks whether the difference in AUC between two models are significant enough to cause problems. Each model is compared with the logistic regression, the main explanatory model:

```{r}
#| label: DeLong Tests

#This is a function for later use for AUC/ROC testing
delong_results <- bind_rows(
  lapply(
    setdiff(employment_models, "Logistic regression"), 
    function(m) {
      test <- roc.test(
        employment_rocs[["Logistic regression"]],
        employment_rocs[[m]],
        method = "delong", 
        paired = TRUE
        )
      
      tibble(
        comparison = paste("Logistic regression vs", m),
        auc_logistic = as.numeric(auc(employment_rocs[["Logistic regression"]])),
        auc_other = as.numeric(auc(employment_rocs[[m]])),
        difference = auc_logistic - auc_other,
        p_value = test$p.value
        )
      }
    )
  )

delong_results |>
  mutate(
    across(
      auc_logistic:difference, \(x) round(x, 3)
      ),
    p_value = format.pval(p_value, digits = 3, eps = 0.0001)
    )
```

The DeLong test treats people as independent, however household members are not fully independent, meaning these p-values are to small so we need to take precaution when using them

## 7.5 Calibration

```{r}
#| label: Calibration Plot
#| fig-width: 8
#| fig-height: 6

employment_test_predictions |>
  pivot_longer(
    all_of(employment_models), 
    names_to = "model", 
    values_to = "probability"
    ) |>
  group_by(model) |>
  mutate(bin = ntile(probability, 10)) |>
  group_by(model, bin) |>
  summarise(
    mean_predicted = mean(probability),
    share_employed = mean(employed == "Employed"),
    .groups = "drop"
    ) |>
  ggplot(
    aes(x = mean_predicted, y = share_employed, colour = model)
    ) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_line() +
  geom_point() +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "Calibration on the test set", 
    x = "Average predicted probability of employment",
    y = "Actual share employed", colour = "Model"
    ) +
  theme_minimal()
```

All four models lie close to the diagonal across the whole range, so their probabilities can be read at face value: none of them are overconfident or overcautious.

## 7.6 Performance for different groups

A model can perform well overall but worse for particular groups. AUC and the threshold metrics are calculated separately by population group, sex and settlement type for the logistic regression and the random forest and get used as input for a fairness evaluation

```{r}
#| label: Subgroup Performance

subgroup_metrics <- function(group_column) {
  bind_rows(
    lapply(
      c("Logistic regression", "Random forest"), 
      function(m) {
      threshold <- employment_thresholds$threshold[employment_thresholds$model == m]
    
      employment_test_predictions |>
      mutate(
        group = as.character(.data[[group_column]]), 
        probability = .data[[m]]
        ) |>
      group_by(group) |>
      summarise(
        people = n(),
        share_employed = mean(employed == "Employed"),
        auc = auc_value(employed, probability),
        classification_metrics(employed, probability, threshold),
        .groups = "drop"
      ) |>
      mutate(
        model = m,
        grouping = group_column, 
        .before = 1
        )
      }
      )
    )
}

bind_rows(
  subgroup_metrics("population_group"),
  subgroup_metrics("female"),
  subgroup_metrics("settlement_type")
) |>
  select(
    grouping,
    group, 
    model, 
    people, 
    share_employed,
    auc, 
    balanced_accuracy,
    sensitivity, 
    specificity
    ) |>
  mutate(across(share_employed:specificity, \(x) round(x, 3)))
```

For `female`, 1 means female and 0 means male.

The AUC is similar across most groups, so both models rank people about equally well within each group. The threshold-based metrics differ much more, because one threshold is applied to groups with very different employment rates:

- For the **Indian/Asian and White** groups, where more than 85% of the labour force is employed, both models almost never predict Unemployed: specificity is 0.2 or lower. Unemployed people in these groups are nearly always misclassified as employed. These groups are also small in the test set, so the estimates are uncertain.
- In **traditional areas**, where only about half the labour force is employed, the logistic regression misses more than half of the employed people (sensitivity about 0.45).

This is not a bug in the models, but a consequence of a single threshold meeting different base rates. It means the classifications should not be used to make decisions about individuals and it is carried into the fairness evaluation.

## 7.7 Income: test-set metrics

```{r}
#| label: Income Test Metrics

income_test_predictions <- income_test |>
  select(
    household_id,
    person_weight,
    log_salary) |>
  mutate(
    "Training average" = weighted.mean(
      income_train$log_salary, 
      income_train$person_weight
      ),
    
    "Baseline" = as.numeric(
      predict(income_baseline, income_test)
      ),
    
    "Linear regression" = as.numeric(
      predict(income_linear, income_test)
      ),
    
    "Random forest" = predict(income_forest_tuned, income_test)$predictions
  )

income_models <- c(
  "Training average",
  "Baseline",
  "Linear regression",
  "Random forest"
  )

income_test_metrics <- bind_rows(
  lapply(
    income_models,
    function(m) {
      bind_rows(
        regression_metrics(
          income_test_predictions$log_salary, income_test_predictions[[m]]
          ) |>
          mutate(
            model = m, 
            weighting = "Unweighted", 
            .before = 1
            ),
        regression_metrics(
          income_test_predictions$log_salary, 
          income_test_predictions[[m]],
          income_test_predictions$person_weight) |>
          mutate(
            model = m,
            weighting = "Population-weighted", 
            .before = 1
            )
        )
      }
    )
  )

income_test_metrics |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

Both the linear regression and the forest clearly beat the two baselines and the test results match the cross-validated ones. Even so, both explain only about 30% of the variation in log salary: education, internet access, age, sex, population group and location leave most of the differences in salary unexplained. The income models are useful for estimating average differences between groups, not for predicting an individual's salary.

## 7.8 Residuals of the linear regression

```{r}
#| label: Residual Plots
#| fig-width: 10
#| fig-height: 4.5

residual_data <- income_test_predictions |>
  transmute(
    fitted = `Linear regression`, 
    residual = log_salary - `Linear regression`
    )

residual_plot <- 
  ggplot(
    residual_data, aes(x = fitted, y = residual)
    ) +
  geom_point(alpha = 0.15) +
  geom_hline(yintercept = 0, colour = "red") +
  geom_smooth(se = FALSE, colour = "steelblue") +
  labs(
    title = "Residuals vs fitted (test set)", 
    x = "Fitted log salary",
    y = "Residual"
    ) +
  theme_minimal()

qq_plot <- ggplot(residual_data, aes(sample = residual)) +
  stat_qq(alpha = 0.3) +
  stat_qq_line(colour = "red") +
  labs(
    title = "Normal Q-Q plot of residuals", 
    x = "Theoretical quantiles",
    y = "Residual quantiles"
    ) +
  theme_minimal()

residual_plot
qq_plot
```

The residuals are centred on zero across the whole range of fitted values (the smoothed blue line stays on the red line), with roughly constant spread, so the linear form and the constant-variance assumption hold reasonably well. The Q-Q plot shows heavier tails than a normal distribution: a small number of people earn far more/less than the model predicts.

## 7.9 How closely the models agree

The metrics above show how well each model predicts, but not whether the models make the same predictions, such as when models with similar AUCs still disagree about many individual people. Correlating the models test set predictions with each other shows how much the more flexible models add beyond the simpler ones: if a flexible model's predictions are almost the same as the regression's, it has found little the regression missed.

The heatmaps show the Pearson correlation between each pair of models predictions.

```{r}
#| label: Prediction Correlations

employment_prediction_correlations <- employment_test_predictions |>
  select(all_of(employment_models)) |>
  cor()

income_prediction_correlations <- income_test_predictions |>
  select(
    Baseline, 
    `Linear regression`,
    `Random forest`, 
    `Actual log salary` = log_salary
    ) |>
  cor()

# Heatmap of a correlation matrix (light for weaker, Dark for stronger correlations)
correlation_heatmap <- function(correlations, title) {
  as_tibble(
    correlations, 
    rownames = "model_1"
    ) |>
    pivot_longer(-model_1, names_to = "model_2", values_to = "correlation") |>
    mutate(
      model_1 = factor(model_1, levels = rownames(correlations)),
      model_2 = factor(model_2, levels = rev(colnames(correlations))),
      text_colour = if_else(correlation >= 0.75, "white", "black")
    ) |>
    ggplot(
      aes(x = model_1, y = model_2, fill = correlation)
      ) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(
      aes(label = sprintf("%.2f", correlation), colour = text_colour), size = 4.5
      ) +
    scale_colour_identity() +
    scale_fill_gradientn(
      colours = c(
        "#cde2fb",
        "#86b6ef", 
        "#3987e5",
        "#1c5cab",
        "#0d366b"
        ),
      limits = c(0.4, 1)
      ) +
    coord_equal() +
    labs(
      title = title,
      x = NULL, 
      y = NULL, 
      fill = "Correlation"
      ) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 30, hjust = 1)
      )
}
```

```{r}
#| label: Employment Correlation Heatmap
#| fig-width: 7
#| fig-height: 5.5

correlation_heatmap(
  employment_prediction_correlations,
  "Employment: correlation of predicted probabilities (test set)"
  )
```

```{r}
#| label: Income Correlation Heatmap
#| fig-width: 7
#| fig-height: 5.5

correlation_heatmap(
  income_prediction_correlations,
  "Income: correlation of predicted log salary (test set)"
  )
```

The same comparison as a table, for every pair of employment models.

```{r}
#| label: Model Agreement Table

employment_classes <- sapply(
  employment_models, 
  function(m) {
    employment_test_predictions[[m]] >=
    employment_thresholds$threshold[employment_thresholds$model == m]
    }
  )

bind_rows(
  lapply(
    combn(employment_models, 2, simplify = FALSE), 
    function(pair) {
      class_1 <- employment_classes[, pair[1]]
      class_2 <- employment_classes[, pair[2]]
   
      observed_agreement <- mean(class_1 == class_2)
      
      chance_agreement <- 
        mean(class_1) * mean(class_2) + (1 - mean(class_1)) * (1 - mean(class_2))
      
      tibble(
        model_1 = pair[1],
        model_2 = pair[2],
        pearson = cor(
          employment_test_predictions[[pair[1]]],
          employment_test_predictions[[pair[2]]]
          ),
        spearman = cor(
          employment_test_predictions[[pair[1]]],
          employment_test_predictions[[pair[2]]],
          method = "spearman"
          ),
        same_class = observed_agreement,
        kappa = (observed_agreement - chance_agreement) / (1 - chance_agreement)
        )
      }
    )
  ) |>
  arrange(desc(pearson)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

- **The selected regressions and the forests mostly make the same predictions.**

- **The decision tree agrees least with the other models**, and least of all with the baseline.

- **The baseline differs clearly from the full models**.

- **The income models agree with each other much more than with the actual salaries**.

# 8. The education × internet interaction

## 8.1 Likelihood-ratio tests

```{r}
#| label: Interaction LRT

employment_lrt <- regTermTest(
  employment_logistic,
  ~ matric_plus:any_home_internet,
  method = "LRT"
  )
income_lrt <- regTermTest(
  income_linear,
  ~ matric_plus:any_home_internet,
  method = "LRT"
  )

tibble(
  model = c(
    "Employment (logistic regression)",
    "Income (linear regression)"
    ),
  interaction_effect = c(
    paste0(
      "Odds ratio ", 
      round(exp(coef(employment_logistic)["matric_plus:any_home_internet"]), 3)
      ),
    paste0(
      round(
        100 * (exp(coef(income_linear)["matric_plus:any_home_internet"]) - 1), 1
            ), 
      "% salary difference"
      )
    ),
  lrt_p_value = c(employment_lrt$p, income_lrt$p),
  significant_at_0.05 = lrt_p_value < 0.05
  ) |>
  mutate(lrt_p_value = format.pval(lrt_p_value, digits = 3, eps = 0.0001))
```

## 8.2 Predicted probabilities for each group, from every model

```{r}
#| label: Group Probabilities

group_average <- function(predict_function, data) {
  expand_grid(matric_plus = 0:1, any_home_internet = 0:1) |>
    rowwise() |>
    mutate(probability = {
      scenario <- data
      scenario$matric_plus <- matric_plus
      scenario$any_home_internet <- any_home_internet
      weighted.mean(predict_function(scenario), data$person_weight)
      }
    ) |>
    ungroup()
  }

group_probabilities <- bind_rows(
  group_average(
    \(d) as.numeric(
      predict(employment_logistic, d, type = "response")
      ), 
    model_train
    ) |>
    mutate(model = "Logistic regression"),
  group_average(
    \(d) 
    predict(employment_tree_tuned, d, type = "prob")[, "Employed"], 
    model_train
    ) |>
    mutate(model = "Decision tree"),
  group_average(
    \(d) 
    predict(employment_forest_tuned, d)$predictions[, "Employed"], 
    model_train
    ) |>
    mutate(model = "Random forest")
) |>
  mutate(group = case_when(
    matric_plus == 0 & any_home_internet == 0 ~ "Neither",
    matric_plus == 1 & any_home_internet == 0 ~ "Matric only",
    matric_plus == 0 & any_home_internet == 1 ~ "Internet only",
    TRUE ~ "Both"
    )
  )

group_probability_table <- group_probabilities |>
  select(model, group, probability) |>
  pivot_wider(
    names_from = group,
    values_from = probability
    ) |>
  mutate(
    interaction_on_probability_scale = 
      (Both - `Internet only`) - (`Matric only` - Neither)
    )

group_probability_table |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

```{r}
#| label: Group Probability Plot
#| fig-width: 8
#| fig-height: 4.5

group_probabilities |>
  mutate(
    Matric = if_else(matric_plus == 1, "Matric or higher", "Below matric"),
    Internet = if_else(any_home_internet == 1, "Home internet", "No home internet")
  ) |>
  ggplot(
    aes(x = Internet, y = probability, colour = Matric, group = Matric)
    ) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~model) +
  labs(
    title = "Average predicted probability of employment by education and internet 
    access",
    x = NULL, 
    y = "Predicted probability of employment",
    colour = NULL
    ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Non-parallel lines mean the internet gap differs by education level, which is the interaction.

The logistic regression and the random forest agree on the **direction**: in both, people with matric gain more from home internet than people without matric, so the two go together as complements, in comparison random forest's interaction was smaller. Random forest was not told about the interaction and found the same pattern on its own, but its estimate is also shrunk by averaging many trees that often do not split on these two predictors. The decision tree gives the same probability to all four groups, because it never splits on education or internet access.

## 8.3 Do the tree-based models use education and internet access?

```{r}
#| label: Tree and Forest Use of Key Predictors

tree_variables <- unique(
  as.character(
    employment_tree_tuned$frame$var[employment_tree_tuned$frame$var != "<leaf>"]
    )
  )

forest_importance <- tibble(
  predictor = names(importance(employment_forest_tuned)),
  permutation_importance = importance(employment_forest_tuned)
  ) |>
  arrange(desc(permutation_importance)) |>
  mutate(rank = row_number())

tibble(
  check = c(
    "Tuned tree splits on matric_plus", 
    "Tuned tree splits on any_home_internet",
    "Forest importance rank of matric_plus (of 15)",
    "Forest importance rank of any_home_internet (of 15)"
    ),
  
  value = c(
    as.character("matric_plus" %in% tree_variables),
    as.character("any_home_internet" %in% tree_variables),
    as.character(
      forest_importance$rank[forest_importance$predictor == "matric_plus"]
      ),
    as.character(
      forest_importance$rank[forest_importance$predictor == "any_home_internet"]
      )
    )
  )

forest_importance |>
  mutate(permutation_importance = signif(permutation_importance, 3))
```

Age, household size and the device predictors drive the tree-based models' predictions far more than education or home internet. Two reasons, which are about the data rather than a fault in the models:

- **Little variation in internet access.** About 85% of the labour force lives in a household with internet access, so splitting on it separates only a small group.
- **Overlap with the devices.** Almost everyone with home internet also has a smartphone in the household, so once a tree has split on "smartphone_access", "any_home_internet" adds little.

Low importance does not mean internet access is unrelated to employment. It means that, for predicting employment, other predictors carry the same information. Whether internet access is *associated* with employment, holding the other predictors fixed, is what the regression's coefficients and the likelihood-ratio test answer.

## 8.4 Sensitivity check: without the device predictors

"has_computer" and "smartphone_access" are strongly linked to home internet: almost everyone with home internet also has a smartphone in the household. They are in the main model as controls for other kinds of digital access and for household wealth, but they may also absorb part of the internet effect and employment can pay for devices (reverse direction). The logistic regression is therefore refitted without them to see whether the conclusions depend on this choice.

```{r}
#| label: Sensitivity Without Devices

employment_logistic_no_devices <- svyglm(
  update(
    employment_regression_formula, 
    . ~ . - has_computer - smartphone_access
    ),
  design = svydesign(
    ids = ~household_id,
    weights = ~person_weight,
    data = model_train
    ),
  family = quasibinomial()
)

key_terms <- c(
  "matric_plus",
  "any_home_internet",
  "matric_plus:any_home_internet"
  )

odds_ratio_table <- function(model, label) {
  estimates <- coef(model)[key_terms]
  limits <- confint(model)[key_terms, ]
  tibble(
    specification = label,
    term = key_terms, 
    odds_ratio = exp(estimates),
    ci_lower = exp(limits[, 1]), ci_upper = exp(limits[, 2]))
}

bind_rows(
  odds_ratio_table(
    employment_logistic,
    "Main model (with devices)"
    ),
  odds_ratio_table(
    employment_logistic_no_devices,
    "Without has_computer and smartphone_access"
    )
) |>
  mutate(across(odds_ratio:ci_upper, \(x) round(x, 3)))

tibble(
  specification = c(
    "Main model (with devices)",
    "Without has_computer and smartphone_access"
    ),
  interaction_lrt_p_value = c(
    employment_lrt$p,
    regTermTest(
      employment_logistic_no_devices, 
      ~ matric_plus:any_home_internet,
      method = "LRT"
      )$p
    )
  ) |>
  mutate(interaction_lrt_p_value = format.pval(
    interaction_lrt_p_value, 
    digits = 3,
    eps = 0.0001)
    )
```

The **interaction** hardly changes (odds ratio about 1.29 with the devices and 1.28 without) and it is significant at α = 0.05 in both, so the complements finding does not depend on whether the devices are controlled for.

The **main effects** do change: without the devices, both matric and home internet are clearly associated with higher odds of employment on their own. With the devices, the internet main effect is small and not significant, because "smartphone_access" carries most of the same information.

# 9. Model comparison and selection

## 9.1 Employment comparison matrix

```{r}
#| label: Employment Comparison Matrix

employment_cv_auc <- tibble(
  model = employment_models,
  cv_auc = c(
    mean(fold_auc(model_train$employed, baseline_oof, employment_fold_id)),
    mean(fold_auc(model_train$employed, logistic_oof, employment_fold_id)),
    tree_selected$mean_cv_auc,
    forest_selected$mean_cv_auc
  )
)

employment_comparison <- employment_test_metrics |>
  filter(weighting == "Unweighted") |>
  select(
    model, 
    auc, 
    auc_lower,
    auc_upper, 
    balanced_accuracy,
    sensitivity,
    specificity,
    f1, 
    brier
    ) |>
  left_join(employment_cv_auc, by = "model") |>
  mutate(
    auc_gain_over_baseline = auc - auc[model == "Baseline"],
    meets_auc_0.70 = if_else(auc >= 0.70, "Yes", "No"),
    test_auc_95_ci = paste0(round(auc_lower, 3), "–", round(auc_upper, 3))
  ) |>
  left_join(
    tibble(
      model = employment_models,
      interpretability = c(
        "High (odds ratios)", 
        "High (odds ratios)",
        "High (if-then rules)",
        "Low (black box)"
        ),
      tests_interaction = c(
        "No (no education or internet predictors)",
        "Yes (odds ratio and LRT)",
        if_else(
          "matric_plus" %in% tree_variables 
          & "any_home_internet" %in% tree_variables,
          "Only through its splits", "No (does not split on both)"
          ),
        "Indirectly (group averages)"
        ),
      uses_population_group_to_predict = if_else(
        c(TRUE, TRUE, "population_group" %in% tree_variables, TRUE), "Yes", "No"
      )
    ),
    by = "model"
  )

employment_comparison |>
  select(
    model,
    cv_auc, 
    test_auc = auc,
    test_auc_95_ci,
    auc_gain_over_baseline,
    meets_auc_0.70,
    balanced_accuracy, 
    sensitivity, 
    specificity, 
    f1, 
    brier,
    interpretability,
    tests_interaction,
    uses_population_group_to_predict
    ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 9.2 Income comparison matrix

```{r}
#| label: Income Comparison Matrix

income_cv_rmse <- tibble(
  model = income_models,
  cv_rmse = c(
    regression_metrics(income_train$log_salary, mean_oof)$rmse,
    regression_metrics(income_train$log_salary, income_baseline_oof)$rmse,
    regression_metrics(income_train$log_salary, linear_oof)$rmse,
    income_forest_selected$mean_cv_rmse
  )
)

income_test_metrics |>
  filter(weighting == "Unweighted") |>
  select(
    model,
    test_rmse = rmse,
    test_mae = mae,
    test_r_squared =
      r_squared
    ) |>
  left_join(income_cv_rmse, by = "model") |>
  left_join(
    tibble(
      model = income_models,
      interpretability = c(
        "Trivial", 
        "High (% differences)", 
        "High (% differences)", 
        "Low (black box)"
        ),
      tests_interaction = c(
        "No", 
        "No",
        "Yes (coefficient, LRT)", 
        "Indirectly"
        )
      ),
    by = "model"
  ) |>
  select(
    model, 
    cv_rmse, 
    test_rmse, 
    test_mae, 
    test_r_squared, 
    interpretability, 
    tests_interaction
    ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## 9.3 Model selection

```{r}
#| label: Selection Figures
#| echo: false

# Short names for the figures quoted in the text below
test_auc <- function(m) round(
  employment_comparison$auc[employment_comparison$model == m], 3
  )

test_ba <- function(m) round(
  employment_comparison$balanced_accuracy[employment_comparison$model == m], 3
  )

test_brier <- function(m) sprintf(
  "%.3f", employment_comparison$brier[employment_comparison$model == m]
  )

delong_p <- function(m) format.pval(
  delong_results$p_value[delong_results$comparison == paste("Logistic regression vs", m)],
  digits = 2, eps = 0.0001
  )

income_rmse <- function(m) round(
  income_test_metrics$rmse[income_test_metrics$model == m & income_test_metrics$weighting == "Unweighted"]
  , 3
  )

income_r2 <- function(m) round(
  income_test_metrics$r_squared[income_test_metrics$model == m & income_test_metrics$weighting == "Unweighted"]
  , 3
  )
```

The brief's key principle is that the highest-scoring model is not automatically the best one, therefore the selection below weighs the evidence in sections 7 and 8 against what each model has to deliver.

**Employment: the survey-weighted logistic regression is selected.**

- **It meets the performance criterion.** Its AUC test is above the 0.70 threshold and clearly above the demographics only baseline. Education, digital access and the household predictors therefore add real predictive information beyond who and where a person is.
- **It gives up little against the best-scoring model.** The random forest has the highest test AUC.
- **It is the only model that answers the business question directly.** It estimates the education and internet interaction as an odds ratiowith a likelihood-ratio test and the survey weights make that estimate describe the South African labour force. Random forest can only show the interaction indirectly, through group averages.
- **It is easier to deploy and explain.** Every prediction can be traced to odds ratios that a policy stakeholder can read and a regression is small and fast to score in a Shiny app.

The **random forest** is kept as the predictive benchmark: it shows that a flexible model gains only a small AUC result over the logistic regression and its predictions correlate with the logistic regression's, which supports using the simpler model.

The **decision tree** is not selected. Its test AUC is below the demographics-only baseline, and it never splits on education or internet access, so it cannot serve as the readable check of the interaction it was chosen for. That role is filled by the random forest instead.

**Income: the survey-weighted linear regression is selected.** The random forest predicts slightly better, a difference of about 1% in RMSE, however it is too small to outweigh the linear regression's advantages: it reports the salary differences as percentages with confidence intervals, it uses the survey weights and its residuals behave acceptably. Its answer to the research question is that the interaction is **not** significant: among people who report a salary, there is no evidence that home internet changes the salary premium for matric. Both models explain only about 30% of the variation in log salary, so the income results remain exploratory.

# 10. Reproducibility

Every random step uses seed 67: the cross-validation folds and every random forest. Rerunning this file therefore gives the same folds, the same tuned settings and the same test results. With the seed set, `ranger` gives identical forests whatever number of processor threads it uses, so the results do not depend on the computer's number of cores.

The files this notebook produces:

| File | Contents |
|------------------------------------|------------------------------------|
| `Models/employment_tree_tuned.rds` | Tuned decision tree |
| `Models/employment_forest_tuned.rds` | Tuned employment random forest |
| `Models/employment_baseline.rds` | Demographics-only employment baseline |
| `Models/income_forest_tuned.rds` | Tuned income random forest |
| `Models/income_baseline.rds` | Demographics-only income baseline |
| `Models/employment_thresholds.csv` | The classification threshold of each employment model |

The selected models, `employment_logistic.rds` and `income_linear.rds`, were saved by the modelling notebooks and are not changed here.
