---
title: "Data Preprocessing For Datasets"
author: "Group 5"
date: today
format:
  html:
    toc: true #Gives a table of contents
    code-fold: show
    df-print: kable
    embed-resources: true #Puts the plots inside the HTML so it can be moved on its own
  pdf:
    toc: true
    df-print: kable
execute:
  warning: false #Makes warnings not appear to keep text clean in the file will appear in console when run
  message: false
---

## Loading the packages we need:

```{r}
#| label: Package Setup

library(readr)
library(tidyverse)
library(janitor)
```

## File paths for the dataset folders:

```{r}
#| label: File-Path Declarations

directory <- "../Datasets/"
cleaned_directory <- "../Datasets/Cleaned/" #Cleaned Datasets get stored here
dir.create(cleaned_directory, showWarnings = FALSE)
```

# 1. Data Import, Structure and Selection

## 1.1 Import with explicit column types

**Problem:** `HouseholdID` is an 18-digit number. R stores numbers as doubles, which are only exact up to about 15–16 digits, so letting `read_csv()` guess the type rounds the IDs (Milestone 1, section 8.3). `PersonNumber` is stored as "01", "02", … and must have the same type in every file for joins to match. `EducationAttendanceCode` and `MinutesToWork` contain text placeholders mixed into numeric codes.

**Treatment:** read every file with explicit types: IDs as text (`col_character()`) in all four files, and the two contaminated columns as text so no value is silently lost. `clean_names()` converts all column names to snake_case. The datasets are named D01, D02, D03 and D08, as in the ERD and the EDA, and are cleaned step by step under the same names. A summary of each uncleaned dataset is recorded in 1.2, before anything changes, for the before/after comparisons.

```{r}
#| label: Import Datasets

D01 <- read_csv(file.path(directory, "D01_Population_Profile.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character(), PersonNumber = col_character())) |> clean_names()

D02 <- read_csv(file.path(directory, "D02_Learning_Profile.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character(), PersonNumber = col_character(), EducationAttendanceCode = col_character())) |> clean_names()

D03 <- read_csv(file.path(directory, "D03_Economic_Participation.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character(), PersonNumber = col_character(),MinutesToWork = col_character())) |> clean_names()

D08 <- read_csv(file.path(directory, "D08_Connectivity_Access.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
```

**Verification:** no parsing problems, IDs stored as text with all 18 digits intact.

```{r}
#| label: Verify Imports

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  parsing_problems = c(nrow(problems(D01)), nrow(problems(D02)), nrow(problems(D03)), nrow(problems(D08))),
  
  household_id_type = c(class(D01$household_id), class(D02$household_id), class(D03$household_id), class(D08$household_id)),
  
  shortest_id = c(min(nchar(D01$household_id), na.rm = TRUE), min(nchar(D02$household_id)), min(nchar(D03$household_id)), min(nchar(D08$household_id))),
  
  longest_id = c(max(nchar(D01$household_id), na.rm = TRUE), max(nchar(D02$household_id)), max(nchar(D03$household_id)), max(nchar(D08$household_id))),
  
  distinct_households = c(n_distinct(D01$household_id, na.rm = TRUE), n_distinct(D02$household_id), n_distinct(D03$household_id), n_distinct(D08$household_id)),
  
  distinct_if_stored_as_number = c(n_distinct(as.numeric(D01$household_id), na.rm = TRUE), n_distinct(as.numeric(D02$household_id)), n_distinct(as.numeric(D03$household_id)), n_distinct(as.numeric(D08$household_id)))
  
)
```

**Effect:** the last column shows what would have happened without this treatment. Stored as numbers, D08's `r format(n_distinct(D08$household_id), big.mark = ",")` distinct households would collapse to `r format(n_distinct(as.numeric(D08$household_id)), big.mark = ",")`, merging different households together. This is exactly the count behind the EDA's 98.77% household match, so that figure was caused by the rounding problem rather than by missing households. As text, every ID keeps all 18 digits.

## 1.2 Uncleaned "before" summary

**Problem:** the rubric requires before-and-after evidence for every major preprocessing decision.

**Treatment:** record the size, missing values, key uniqueness and duplicates of each uncleaned dataset in `uncleaned_summary`, so the later sections can compare the cleaned data against the uncleaned data.

```{r}
#| label: Uncleaned Datasets Summary

uncleaned_summary <- tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  cols = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08)),
  
  missing_values = c(sum(is.na(D01)), sum(is.na(D02)), sum(is.na(D03)), sum(is.na(D08))),
  
  missing_household_id = c(sum(is.na(D01$household_id)), sum(is.na(D02$household_id)),sum(is.na(D03$household_id)), sum(is.na(D08$household_id))),
  
  distinct_keys = c(nrow(distinct(D01, household_id, person_number)), nrow(distinct(D02, household_id, person_number)),
                    nrow(distinct(D03, household_id, person_number)), n_distinct(D08$household_id)),
  
  exact_duplicate_rows = c(nrow(D01) - nrow(distinct(D01)), nrow(D02) - nrow(distinct(D02)),
                           nrow(D03) - nrow(distinct(D03)), nrow(D08) - nrow(distinct(D08)))
  
)
uncleaned_summary
```

Columns with missing values in each uncleaned dataset:

```{r}
#| label: Missing Values in Datasets


bind_rows(
  D01 = enframe(colSums(is.na(D01)), name = "column", value = "missing"),
  D02 = enframe(colSums(is.na(D02)), name = "column", value = "missing"),
  D03 = enframe(colSums(is.na(D03)), name = "column", value = "missing"),
  D08 = enframe(colSums(is.na(D08)), name = "column", value = "missing"),
  .id = "dataset"
) |>
  filter(missing > 0) |> #Only show columns that have missing values
  left_join(select(uncleaned_summary, dataset, rows), by = "dataset") |>
  mutate(percent_of_rows = round(missing / rows * 100, 2))
```

## 1.3 Grain and key checks

**Problem:** before any join, the unit of observation (grain) and the key of every dataset must be confirmed, and it must be shown that the datasets can legitimately be linked.

**Treatment:** compare row counts with distinct keys, and check how many households each selected dataset shares with D01.

```{r}
#| label: Grain Check/Verification

D01_repeated_keys <- D01 |> filter(!is.na(household_id)) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D02_repeated_keys <- D02 |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D03_repeated_keys <- D03 |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D08_repeated_keys <- D08 |> count(household_id) |> filter(n > 1) |> nrow()

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  grain = c("person", "person", "person", "household"),
  
  key = c("household_id + person_number", "household_id + person_number", "household_id + person_number", "household_id"),
  
  rows = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  distinct_keys = c(nrow(distinct(filter(D01, !is.na(household_id)), household_id, person_number)), nrow(distinct(D02, household_id, person_number)), nrow(distinct(D03, household_id, person_number)), n_distinct(D08$household_id)),
  
  keys_appearing_more_than_once = c(D01_repeated_keys, D02_repeated_keys, D03_repeated_keys, D08_repeated_keys),
  
  rows_missing_household_id = c(sum(is.na(D01$household_id)), sum(is.na(D02$household_id)), sum(is.na(D03$household_id)), sum(is.na(D08$household_id)))
  
)
```

```{r}
#| label: Shared Households Check

D01_households <- unique(na.omit(D01$household_id)) #Distinct household IDs in D01

tibble(

  compared_with = c("D02", "D03", "D08"),

  households_in_dataset = c(n_distinct(D02$household_id), n_distinct(D03$household_id), n_distinct(D08$household_id)),

  shared_with_D01 = c(length(intersect(D01_households, D02$household_id)), length(intersect(D01_households, D03$household_id)), length(intersect(D01_households, D08$household_id))),
  
  percent_of_D01 = round(shared_with_D01 / length(D01_households) * 100, 2)
  
)
```

```{r}
#| label: People per Household Check

D01 |>
  filter(!is.na(household_id)) |>
  count(household_id, name = "persons") |>
  summarise(households = n(), mean_persons = round(mean(persons), 2), median_persons = median(persons), max_persons = max(persons))
```

**Effect / Verification:** D01, D02 and D03 are one row per person and D08 is exactly one row per household (`r format(nrow(D08), big.mark = ",")` rows, `r format(n_distinct(D08$household_id), big.mark = ",")` distinct IDs). All `r format(length(D01_households), big.mark = ",")` D01 households appear in D02, D03 and D08 (100%). This corrects the 98.77% reported in the EDA: its count of `r format(n_distinct(as.numeric(D08$household_id)), big.mark = ",")` is the number of distinct IDs left after rounding (step 1.1). Household IDs repeat in the person files because a household contains several people. These are legitimate repeats in a one-to-many relationship, not duplicates. The keys that appear more than once in D02 and D03, and the missing IDs in D01, are treated in 2.1 and 2.2.

## 1.4 Datasets not selected

**Problem:** eleven datasets are available, but a dataset should only be used if it measures a link in the research pathway (digital access → education → employment) and can be linked to the same people at the chosen grain: one row per person, identified by `household_id` + `person_number`. Using a dataset that fails either test adds variables with no role in the analysis, or attaches records to the wrong people.

The seven datasets not selected in Task 1 are read here only to document why they are not used:

```{r}
#| label: Datasets Not Selected

D04 <- read_csv(file.path(directory, "D04_Dwelling_Conditions.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

D05 <- read_csv(file.path(directory, "D05_Basic_Services.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

D06 <- read_csv(file.path(directory, "D06_Household_Energy.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

D07 <- read_csv(file.path(directory, "D07_Household_Wellbeing.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

D09 <- read_csv(file.path(directory, "D09_Household_Resources.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

D10 <- read_csv(file.path(directory, "D10_Mobility_Access.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()

#D11 is large, so only its keys and quarter are read in full; its column names are read from the header
D11 <- read_csv(file.path(directory, "D11_Labour_Market.csv"), show_col_types = FALSE, col_select = c(HouseholdID, PersonNumber, Quarter),
                col_types = cols(.default = col_character())) |> clean_names()

D11_columns <- read_csv(file.path(directory, "D11_Labour_Market.csv"), show_col_types = FALSE, n_max = 0) |> clean_names() |> names()

tibble(

  dataset = c("D04", "D05", "D06", "D07", "D09", "D10", "D11"),

  measures = c("Dwelling type, materials and condition", "Water and sanitation", "Electricity and energy use",
               "Food security and household income", "Income sources and household assets",
               "Public transport and access to health facilities", "Employment, education and occupation"),

  grain = c("household", "household", "household", "household", "household", "household", "person per quarter"),

  rows = c(nrow(D04), nrow(D05), nrow(D06), nrow(D07), nrow(D09), nrow(D10), nrow(D11)),

  distinct_households = c(n_distinct(D04$household_id), n_distinct(D05$household_id), n_distinct(D06$household_id),
                          n_distinct(D07$household_id), n_distinct(D09$household_id), n_distinct(D10$household_id),
                          n_distinct(D11$household_id)),

  households_shared_with_D01 = c(length(intersect(D01_households, D04$household_id)),
                                 length(intersect(D01_households, D05$household_id)),
                                 length(intersect(D01_households, D06$household_id)),
                                 length(intersect(D01_households, D07$household_id)),
                                 length(intersect(D01_households, D09$household_id)),
                                 length(intersect(D01_households, D10$household_id)),
                                 length(intersect(D01_households, D11$household_id))),

  reason_not_selected = c("Not a link in the pathway", "Not a link in the pathway", "Not a link in the pathway",
                          "Not in the pathway; household income partly reveals the employment outcome",
                          "Salary income reveals the employment outcome; computer_asset_code duplicates D08",
                          "Not a link in the pathway",
                          "Different sample: cannot be linked to any person in D01")
)
```

D09's only digital-access column is a copy of the one already selected from D08:

```{r}
#| label: D09 Computer Column Check

D09 |>
  distinct(household_id, computer_asset_code_d09 = computer_asset_code) |>
  inner_join(select(D08, household_id, computer_asset_code), by = "household_id") |>
  summarise(households_compared = n(), values_that_differ_from_D08 = sum(computer_asset_code_d09 != computer_asset_code))
```

D11 is the only unselected dataset that measures the employment outcome, so the reasons it cannot be joined are checked directly:

```{r}
#| label: D11 Linkage Check

tibble(
  check = c("rows in D11",
            "distinct people in D11",
            "people recorded in more than one quarter",
            "households shared with D01",
            "PSUs (sampling areas) shared with D01",
            "digital-access columns in D11"),

  result = c(nrow(D11),
             nrow(distinct(D11, household_id, person_number)),
             sum(count(D11, household_id, person_number)$n > 1),
             length(intersect(D11$household_id, D01_households)),
             length(intersect(str_sub(D11$household_id, 1, 11), str_sub(D01_households, 1, 11))), #First 11 digits of an ID = PSU (2.1)
             length(str_subset(D11_columns, "internet|phone|computer")))
)

#Levels of geography at which D11 could be linked to the four selected datasets
selected_columns <- c(names(D01), names(D02), names(D03), names(D08))

tibble(
  geography_level = c("province", "geography type", "metro area"),

  column_in_D11 = c(any(str_detect(D11_columns, "province")), any(str_detect(D11_columns, "geo_type")), any(str_detect(D11_columns, "metro"))),

  column_in_selected_datasets = c(any(str_detect(selected_columns, "province")), any(str_detect(selected_columns, "geo_type")),
                                  any(str_detect(selected_columns, "metro")))
)
```

D11 shares only province and geography type with the selected datasets. It has no metro column, and its PSUs never match D01's (table above), so province and geography type are the most detailed levels at which D11 could be linked to our people.

**Treatment:** select only D01, D02, D03 and D08, and remove the other seven from memory.

- **D04, D05, D06, D07, D09 and D10** are household-level and share almost every household with D01, so they could technically be joined many-to-one. They are left out because of the topic, not because they cannot be linked. Dwelling conditions, water, energy and public transport are not links between digital access, education and employment. Household income (D07, D09) and salary income sources (D09) partly reveal whether someone in the household is employed, so they would leak the outcome. D09's only digital-access column duplicates D08's. D04, D07 and D09 also contain repeated household rows, which would duplicate people if joined without treatment.
- **D11** would be useful, because it records employment status and education for working-age people. It cannot be integrated at the person grain, however. It shares no households and not even a single PSU with D01, so it is a separate sample of different people, not extra information about ours. Its rows are person-quarters, not persons: most people appear in more than one quarter, with separate quarterly and annual weights. The only way to join it would be to match people on shared characteristics such as province, age and sex, which would attach one person's labour-market record to a different person. This is exactly the misalignment that invalid joins create.
- **Aggregation is not a useful alternative for D11 in this topic.** The only geography D11 shares with D01 is province and geography type. A D11 rate aggregated to province and attached to each person would be fully determined by `province_code`, which D01 already contains, so it adds no information at the person level. D11 also has no digital-access columns, so a parallel analysis could only repeat the education → employment link and could not test digital access or its moderating role.

```{r}
#| label: Remove Unselected Datasets

rm(D04, D05, D06, D07, D09, D10, D11, D11_columns, selected_columns) #Only needed for the checks above
```

**Effect / Verification:** the tables confirm the decision. All six household datasets share almost all of D01's households, so their exclusion is a topic decision. D09's computer column never differs from D08's, and D11 shares 0 households and 0 PSUs with D01. The rest of the pipeline cleans and integrates only D01, D02, D03 and D08 (section 9). The employment outcome comes from a single source (D03), which is recorded as a limitation.

# 2. Data Cleaning

## 2.1 Recover the missing household IDs in D01

**Problem:** `r sum(is.na(D01$household_id))` rows in D01 have a missing `household_id`. These people cannot be linked to their education, employment or household connectivity. Exactly the same number of people appear in D03 (and D02) but not in D01, which suggests these are the same people with their ID blanked out.

**Treatment:** match each D01 row with a missing ID to one of the people missing from D01, using `person_number`, `sex_code`, `province_code` and `person_weight`. The survey weight is identical for the same person across D01–D03 and has thousands of distinct values, so it is highly specific. The match is then checked with two attributes that were **not** used to match: the PSU (sampling cluster) must equal the first 11 digits of the recovered household ID, and the age must agree with D03.

```{r}
#| label: Identify Missing People

D01_key_duplicates_before <- get_dupes(D01, household_id, person_number) 

missing_from_D01 <- D03 |>
  distinct() |>
  anti_join(filter(D01, !is.na(household_id)), by = c("household_id", "person_number")) |> 
  select(household_id, person_number, sex_code, province_code, person_weight, age_d03 = age)

rows_missing_id <- D01 |>
  mutate(row_id = row_number()) |> 
  filter(is.na(household_id)) |>
  select(-household_id)

recovered <- rows_missing_id |>
  inner_join(missing_from_D01, by = c("person_number", "sex_code", "province_code", "person_weight"))
```

**Verification:** every D01 row with a missing ID matches exactly one person, every candidate is used exactly once, and both independent checks agree for every match.

```{r}
#| label: Verification of Missing People 

tibble(
  
  D01_rows_missing_id = nrow(rows_missing_id),
  
  people_missing_from_D01 = nrow(missing_from_D01),
  
  matches_found = nrow(recovered),
  
  each_d01_row_matched_once = n_distinct(recovered$row_id) == nrow(rows_missing_id),
  
  each_person_used_once = nrow(distinct(recovered, household_id, person_number)) == nrow(recovered),
  
  psu_agrees_with_id = sum(sprintf("%.0f", recovered$psu) == str_sub(recovered$household_id, 1, 11)), 
  
  age_agrees_with_D03 = sum(recovered$age == recovered$age_d03, na.rm = TRUE)
)
```

```{r}
#| label: Recover Missing People

  D01 <- D01 |> mutate(row_id = row_number()) |>
  left_join(select(recovered, row_id, recovered_id = household_id), by = "row_id") |> 
  
  mutate(household_id_flag = if_else(is.na(household_id) & !is.na(recovered_id), "recovered by matching to D03", NA_character_), household_id =  coalesce(household_id, recovered_id)) |> select(-row_id, -recovered_id)

tibble(
  
  missing_household_id_before = uncleaned_summary$missing_household_id[1], 

  missing_household_id_after = sum(is.na(D01$household_id)),
  
  rows_before = uncleaned_summary$rows[1],
  
  rows_after = nrow(D01),
  
  D01_keys_identical_to_D02 = setequal(paste(D01$household_id, D01$person_number), paste(D02$household_id, D02$person_number))
  
)
```

**Effect:** all `r nrow(recovered)` IDs are recovered with no rows added or removed, and D01 now contains exactly the same set of people as D02. Recovered rows are marked in `household_id_flag`.

## 2.2 Duplicates

**Problem:** the EDA found `r nrow(D01_key_duplicates_before)` (D01), `r nrow(get_dupes(D02, household_id, person_number))` (D02) and `r nrow(get_dupes(D03, household_id, person_number))` (D03) rows sharing a person key. These must be separated into genuine duplicate records (the same record entered twice) and legitimate repeated identifiers.

**Investigation:** every one of D01's key "duplicates" had a missing household ID. Rows with a missing ID and the same person number were grouped together, even though they are different people.

```{r}
#| label: False Duplicates in D01

D01_key_duplicates_before |> count(household_id_missing = is.na(household_id), name = "rows")
```

After the recovery in 2.1, check each dataset for exact copies (every column identical) and for keys that still repeat with different values (conflicts):

```{r}
#| label: Duplicate Check

#Rows that share their key with at least one other row
D01_key_duplicate_rows <- nrow(get_dupes(D01, household_id, person_number))
D02_key_duplicate_rows <- nrow(get_dupes(D02, household_id, person_number))
D03_key_duplicate_rows <- nrow(get_dupes(D03, household_id, person_number))
D08_key_duplicate_rows <- nrow(get_dupes(D08, household_id))

#Exact copies: rows where every column is identical to another row
D01_exact_duplicates <- nrow(D01) - nrow(distinct(D01))
D02_exact_duplicates <- nrow(D02) - nrow(distinct(D02))
D03_exact_duplicates <- nrow(D03) - nrow(distinct(D03))
D08_exact_duplicates <- nrow(D08) - nrow(distinct(D08))

#Keys that still repeat once exact copies are removed, meaning the same person with conflicting values
D01_conflicting_keys <- distinct(D01) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D02_conflicting_keys <- distinct(D02) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D03_conflicting_keys <- distinct(D03) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D08_conflicting_keys <- distinct(D08) |> count(household_id) |> filter(n > 1) |> nrow()

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows_before = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  rows_in_key_duplicate_groups = c(D01_key_duplicate_rows, D02_key_duplicate_rows, D03_key_duplicate_rows, D08_key_duplicate_rows),
  
  exact_duplicate_rows = c(D01_exact_duplicates, D02_exact_duplicates, D03_exact_duplicates, D08_exact_duplicates),
  
  conflicting_keys_after_removing_exact_copies = c(D01_conflicting_keys, D02_conflicting_keys, D03_conflicting_keys, D08_conflicting_keys)
)
```

**Treatment:** every key duplicate in D02 and D03 is an exact copy of another row, with no conflicting values, so they are genuine duplicate records. Exact copies are removed with `distinct()`. D01 and D08 contain none.

```{r}
#| label: Remove Duplicates

D01 <- distinct(D01)
D02 <- distinct(D02)
D03 <- distinct(D03)
D08 <- distinct(D08)

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows_before = uncleaned_summary$rows,
  
  rows_after = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  rows_removed = rows_before - rows_after,
  
  keys_unique_after = c(nrow(distinct(D01, household_id, person_number)) == nrow(D01), nrow(distinct(D02, household_id, person_number)) == nrow(D02), nrow(distinct(D03, household_id, person_number)) == nrow(D03), n_distinct(D08$household_id) == nrow(D08))
)
```

**Effect / Verification:** D02 loses `r D02_exact_duplicates` rows and D03 loses `r D03_exact_duplicates`. All three person datasets now contain exactly `r format(nrow(D01), big.mark = ",")` people, and every key is unique in every dataset. Repeated `household_id` values remain in the person files, because they are legitimate: one household, several people.

## 2.3 Text placeholders and data types

**Problem:** a code column should contain only numbers. Listing the text columns in each dataset shows which columns contain text where codes are expected (the `_flag` columns created in this pipeline are text by design and are left out):

```{r}
#| label: Text Columns Explanation

bind_rows(
  D01 = enframe(sapply(D01, class), name = "column", value = "type"),
  D02 = enframe(sapply(D02, class), name = "column", value = "type"),
  D03 = enframe(sapply(D03, class), name = "column", value = "type"),
  D08 = enframe(sapply(D08, class), name = "column", value = "type"),
  .id = "dataset"
) |>
  filter(type == "character", !str_ends(column, "_flag")) |> 
  group_by(column) |>
  summarise(datasets = paste(dataset, collapse = ", ")) |> 
  mutate(why_text = case_when(
    column %in% c("household_id", "person_number") ~ "Key, read as text on purpose in 1.1",
    column == "province" ~ "Province name (rebuilt from province_code, dropped in section 3)",
    TRUE ~ "Code column containing text placeholders"
  ))
```

Apart from IDs and the province name, only two columns are text: `education_attendance_code` (D02, kept) and `minutes_to_work` (D03, dropped as leakage in section 3, so it is not treated). `education_attendance_code` mixes numeric codes with three text placeholders, the same problem Milestone 1 found in D11's `hours_worked_raw`:

```{r}
#| label: Initial Student Attendance

D02 |> count(education_attendance_code)
```

None of the placeholders appears in the data dictionary. "not recorded" and "Unknown" are non-response. "N/A" cannot be assumed to mean code 8 ("not applicable"): comparing ages shows that code 8 is only used for children aged 0–4, while the placeholders occur at ages up to 72:

```{r}
#| label: Attendance Codes Placeholder Verification

attendance_text <- c("N/A", "not recorded", "Unknown")
D02 |>
  filter(education_attendance_code %in% c("8", attendance_text)) |>
  group_by(value = education_attendance_code) |>
  summarise(rows = n(), min_age = min(age), median_age = median(age), max_age = max(age))
```

**Treatment:** set the text placeholders to missing, record why in `education_attendance_flag`, and convert the column to integer.

```{r}
#| label: Fixing Broken Attendance Codes

D02 <- D02 |>
  mutate(
    education_attendance_flag = if_else(education_attendance_code %in% attendance_text, paste0("text placeholder '", education_attendance_code, "' set to missing"), NA_character_),
    
    education_attendance_code = as.integer(if_else(education_attendance_code %in% attendance_text, NA_character_, education_attendance_code))
  )

```

**Verification:**

```{r}
#| label: Fixed Attendance Codes

D02 |> count(education_attendance_code, education_attendance_flag)
class(D02$education_attendance_code)
```

**Effect:** `education_attendance_code` is now an integer column holding only dictionary codes (1, 2, 3, 8) or missing. The `r sum(!is.na(D02$education_attendance_flag))` former placeholders are traceable through the flag. Code 3 ("do not know") is handled in section 4 with the other special codes.

## 2.4 Repairs from other datasets

### Age (D01)

**Problem:** the data dictionary gives a valid age range of 0–110. D01 has `r sum(is.na(D01$age))` missing ages and `r sum(D01$age < 0 | D01$age > 110, na.rm = TRUE)` ages outside that range. D02 and D03 record the age of the same people.

**Investigation:** check that D02 and D03 agree with each other, and that D01 agrees with D03 wherever D01's age is valid:

```{r}
#| label: Ages Inspection

age_check <- D01 |>
  left_join(select(D03, household_id, person_number, age_d03 = age), by = c("household_id", "person_number")) |>
  left_join(select(D02, household_id, person_number, age_d02 = age), by = c("household_id", "person_number"))

age_check |>
  summarise(
    
    rows = n(),
    
    d02_vs_d03_mismatches = sum(age_d02 != age_d03),
    
    d03_ages_outside_0_110 = sum(age_d03 < 0 | age_d03 > 110),
    
    d01_valid_ages_agreeing_with_d03 = sum(age == age_d03 & between(age, 0, 110), na.rm = TRUE),
    
    d01_valid_ages_disagreeing_with_d03 = sum(age != age_d03 & between(age, 0, 110), na.rm = TRUE),
    
    d01_missing = sum(is.na(age)),
    
    d01_invalid = sum(age < 0 | age > 110, na.rm = TRUE)
    
  )

age_check |>
  filter(age < 0 | age > 110) |> select(household_id, person_number, age_d01 = age, age_d02, age_d03)
```

D02 and D03 agree on every person and contain no invalid ages, and D01 agrees with them on every valid age. The only disagreements are D01's invalid values, so D03's age is a reliable replacement. Deleting these people would lose valid education and employment records for no reason.

**Treatment:** replace D01's missing and invalid ages with D03's age and record which rows changed in `age_flag`.

```{r}
#| label: Fixed Ages

D01 <- age_check |>
  mutate(
    
    age_flag = case_when(is.na(age) ~ "missing - filled from D03",
      age < 0 | age > 110 ~ paste0("invalid (", age, ") - replaced from D03")
    ),
    
    age = if_else(is.na(age) | age < 0 | age > 110, age_d03, age)
  ) |>
  
  select(-age_d02, -age_d03)
```

**Verification:**

```{r}
#| label: Verification of Ages after Fix

tibble(
  rows = nrow(D01),
  
  missing_age = sum(is.na(D01$age)),
  
  min_age = min(D01$age),
  
  max_age = max(D01$age),
  
  ages_differing_from_d03 = sum(D01$age != age_check$age_d03)
)

D01 |> count(age_flag = if_else(is.na(age_flag), "unchanged", str_remove(age_flag, " \\(.*\\)")))
```

**Effect:** `r sum(!is.na(D01$age_flag))` ages are repaired and none are deleted. D01's age now lies between 0 and 110 with no missing values, and agrees with D02 and D03 for every person.

### Education level (D02)

**Problem:** `education_level_code` in D02 has `r sum(is.na(D02$education_level_code))` missing values. D03 records the same variable for the same people.

```{r}
#| label: Education Dataset Inspection

D02_education_missing_before <- sum(is.na(D02$education_level_code)) #Kept for the before/after check below

education_check <- D02 |>
  left_join(select(D03, household_id, person_number, education_d03 = education_level_code), by = c("household_id", "person_number"))

education_check |>
  summarise(
    
    rows = n(),
    
    missing_in_d02 = sum(is.na(education_level_code)),
    
    missing_in_d03 = sum(is.na(education_d03)),
    
    agree_where_both_present = sum(education_level_code == education_d03, na.rm = TRUE),
    
    disagree_where_both_present = sum(education_level_code != education_d03, na.rm = TRUE),
    
    d02_missing_but_d03_present = sum(is.na(education_level_code) & !is.na(education_d03))
    
  )
```

**Treatment:** D02 and D03 never disagree, and D03 has no missing values, so D02's gaps are filled from D03 and flagged in `education_level_flag`.

```{r}
#| label: Fixed Education Dataset 

D02 <- education_check |>
  mutate(
    education_level_flag = if_else(is.na(education_level_code), "missing - filled from D03", NA_character_),
    education_level_code = coalesce(education_level_code, education_d03)
  ) |>
  select(-education_d03)

tibble(

       missing_before = D02_education_missing_before,
       
       missing_after = sum(is.na(D02$education_level_code)),
       
       rows_flagged = sum(!is.na(D02$education_level_flag))
)
```

**Effect:** D02's education level is complete. Code 99, which appears in both D02 and D03 but not in the dictionary, is not a gap and cannot be filled from D03. It is treated with the other special codes in section 4.

# 3. Column Selection

**Problem:** the four datasets contain `r sum(uncleaned_summary$cols)` columns, most of which do not measure a link in the research pathway, duplicate another dataset, or leak the outcome. The full reasoning for each column is in the Task 1 write-up (Data Selection and Analytical Grain). The evidence below supports the main drop decisions.

## Evidence: columns that carry no information or duplicate another column

```{r}
#| label: Evidence of Province Mismatches

tibble(
  check = c("distinct survey_year values in D01", "distinct survey_year values in D02",
            "distinct survey_year values in D03", "distinct survey_year values in D08",
            "distinct province spellings in D01",
            "distinct province spellings after trimming and case-fixing",
            "distinct province_code values"),
  
  result = c(n_distinct(D01$survey_year), n_distinct(D02$survey_year),
             n_distinct(D03$survey_year), n_distinct(D08$survey_year),
             n_distinct(D01$province),
             n_distinct(str_to_title(str_squish(D01$province))),
             n_distinct(D01$province_code))
)
D01 |> distinct(province_code, province_clean = str_to_title(str_squish(province))) |> arrange(province_code)
```

```{r}
#| label: Evidence of Duplicate Columns

persons <- D01 |> inner_join(D02, by = c("household_id", "person_number"), suffix = c("", "_d02")) |>
  inner_join(D03, by = c("household_id", "person_number"), suffix = c("", "_d03"))


tibble(
  
  column = c("sex_code", "province_code", "geo_type_code", "person_weight", "age"),
  
  people_compared = nrow(persons),
  
  mismatches_D01_vs_D02 = 
c(sum(persons$sex_code != persons$sex_code_d02), sum(persons$province_code != persons$province_code_d02), sum(persons$geo_type_code != persons$geo_type_code_d02), sum(persons$person_weight != persons$person_weight_d02), sum(persons$age != persons$age_d02)),
  
  mismatches_D01_vs_D03 = 
c(sum(persons$sex_code != persons$sex_code_d03), sum(persons$province_code != persons$province_code_d03), sum(persons$geo_type_code != persons$geo_type_code_d03), sum(persons$person_weight != persons$person_weight_d03), sum(persons$age != persons$age_d03))
  
)

#D08 against D01 at household level
D01 |>
  distinct(household_id, province_code, geo_type_code) |>
  inner_join(D08, by = "household_id", suffix = c("", "_d08")) |>
  summarise(households_compared = n(),
  province_mismatch = sum(province_code != province_code_d08),
  geo_type_mismatch = sum(geo_type_code != geo_type_code_d08))


tibble(
  
      age_group_matches_age = sum(as.integer(cut(D01$age, c(-Inf, seq(4, 74, 5), Inf))) == D01$age_group_code), 
      
      rows = nrow(D01)
      
)

```

## Evidence: leakage (columns that reveal or define the employment outcome)

```{r}
#| label: Evidence of Leakage

D03 |> count(employment_status1code, employment_status2code) 

D03 |> count(wage_work_code, employment_status2code) 

D03 |> count(looking_for_work_code, employment_status2code) 

D03 |> count(sector_code, employment_status2code) 
```

## Evidence: unusable or redundant columns

```{r}
#| label: Evidence of Unusable/Redundant Columns

D03 |> count(salary_amount_raw) #Only a handful of distinct values, so not Rand amounts

D02 |>
  filter(between(age, 15, 64), education_attendance_code == 1, grade_code %in% 1:12, education_level_code %in% 0:12) |>
  count(grade_minus_education_level = grade_code - education_level_code) 

D02 |>
  filter(between(age, 15, 64)) |>
  summarise(
    across(
      c(public_private_institution_code, transport_mode_to_education_code, travel_time_to_education, bursary_code, total_education_fees, no_fees_school_code, school_food_programme_code, absence_code, absence_days), ~ round(mean(.x %in% c(8, 88)) * 100, 1))) |>
  pivot_longer(everything(), names_to = "column", values_to = "percent_not_applicable_age_15_64")
```

## Evidence: education columns kept or dropped for Link A (digital access → education)

```{r}
#| label: Linking Digital Access to Education

youth <- D02 |> filter(between(age, 15, 24))

youth |> filter(education_attendance_code == 1) |> count(institution_type_code) 
youth |> filter(education_attendance_code == 2) |> count(reason_not_attending_code, sort = TRUE) #Kept for descriptive use only

youth |>
  left_join(select(D08, household_id, study_internet_code), by = "household_id") |> 
  filter(education_attendance_code %in% 1:2, study_internet_code %in% 1:2) |>
  count(study_internet_code, education_attendance_code) 
```

**Treatment:** keep only the columns selected in the Task 1 write-up, plus the join keys and the traceability flags created during data cleaning (section 2).

```{r}
#| label: Selecting which columns to keep

#Column names before selection, kept to show which columns were dropped
D01_columns_before <- names(D01)
D02_columns_before <- names(D02)
D03_columns_before <- names(D03)
D08_columns_before <- names(D08)

D01 <- D01 |>
  select(household_id, person_number, psu, province_code, sex_code, age, population_group_code, geo_type_code, metro_indicator, person_weight, household_id_flag, age_flag)

D02 <- D02 |>
  select(household_id, person_number, education_level_code, education_attendance_code, institution_type_code, reason_not_attending_code, education_level_flag, education_attendance_flag)

D03 <- D03 |>
  select(household_id, person_number, employment_status2code, lab_salary_raw)

D08 <- D08 |>
  select(household_id, fixed_internet_code, mobile_internet_code, computer_asset_code, household_smartphones)

#Column types after selection, compared with the final types in section 8
D01_types_before <- sapply(D01, class)
D02_types_before <- sapply(D02, class)
D03_types_before <- sapply(D03, class)
D08_types_before <- sapply(D08, class)
```

**Verification:**

```{r}
#| label: Comparing Before and After Column Selection

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  columns_in_uncleaned_data = uncleaned_summary$cols,
  
  columns_before_selection = c(length(D01_columns_before), length(D02_columns_before), length(D03_columns_before), length(D08_columns_before)),
  
  columns_after_selection = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08))
  
)
```

Columns dropped from each dataset:

```{r}
#| label: Columns Dropped

#Dropped columns for each dataset, stacked into one table
columns_dropped <- bind_rows(
  tibble(dataset = "D01", column = setdiff(D01_columns_before, names(D01))),
  tibble(dataset = "D02", column = setdiff(D02_columns_before, names(D02))),
  tibble(dataset = "D03", column = setdiff(D03_columns_before, names(D03))),
  tibble(dataset = "D08", column = setdiff(D08_columns_before, names(D08)))
) |>
  
  #Detailed Table explainning why each column gets dropped
  mutate(
    reason = case_when(
    column == "survey_year" ~ "No information (always 2024)",
    
    column == "province" ~ "Messy text, rebuilt from province_code",
    
    column %in% c("province_code", "age", "sex_code", "geo_type_code", "person_weight") ~ "Duplicate of D01",
    
    column == "education_level_code" ~ "Duplicate of D02 (used to fill D02's gaps in 2.4)",
    
    column %in% c("age_group_code", "employment_status1code", "grade_code") ~ "Duplicate of another column",
    
    column %in% c("wage_work_code", "business_work_code", "volunteer_work_code", "retired_code", "looking_for_work_code", "could_start_work_code", "transport_to_work_code", "minutes_to_work", "sector_code", "work_internet_code", "study_internet_code") ~ "Leakage: reveals or defines an outcome",
    
    column %in% c("language_code", "relationship_to_head_code", "marital_status_code") ~ "Not needed for any hypothesis",
    
    column == "metro_code" ~ "Too detailed (disclosure risk)",
    
    column %in% c("public_private_institution_code", "transport_mode_to_education_code", "travel_time_to_education", "bursary_code", "total_education_fees", "no_fees_school_code", "school_food_programme_code", "absence_code", "absence_days") ~ "Only recorded for people already attending",
    
    column == "salary_amount_raw" ~ "Unusable (not Rand amounts)",
    
    column %in% c("salary_period_code", "salary_category_code") ~ "Salary detail not needed (lab_salary_raw kept)",
    
    column %in% c("telephone_code", "cellphone_code", "household_mobile_phones") ~ "Phones, not internet access",
    
    column %in% c("public_wi_fi_internet_code", "library_internet_code", "cafe_internet_code") ~ "Internet access outside the home",
    
    column == "household_weight" ~ "Household weight, not needed at person grain",
    
    TRUE ~ "NOT CLASSIFIED" #Should never appear; shows up if a dropped column has no reason
  )
  )

columns_dropped |>
  group_by(dataset, reason) |>
  summarise(columns_dropped = n(), columns = paste(column, collapse = ", "), .groups = "drop")
```

**Effect:** the four datasets go from `r sum(uncleaned_summary$cols)` uncleaned columns to `r ncol(D01) + ncol(D02) + ncol(D03) + ncol(D08)`, of which `r sum(str_ends(c(names(D01), names(D02), names(D03), names(D08)), "_flag"))` are the traceability flags from section 2. All `r nrow(columns_dropped)` dropped columns have a documented reason (`r sum(columns_dropped$reason == "NOT CLASSIFIED")` are "NOT CLASSIFIED"). `select()` removes columns only, so every person and household cleaned in section 2 is still present.

# 4. Special Values

Several kept columns contain values that are not ordinary answers: structural "not applicable" codes, non-response codes, codes missing from the data dictionary, and placeholder amounts in `lab_salary_raw`. Each is classified from evidence before anything is changed.

## 4.1 Special codes in the kept columns

**Problem:** the kept code columns mix valid categories with "not applicable" codes, non-response codes and codes that are not in the data dictionary. Treating them all as ordinary categories would put "do not know" next to real answers, while deleting the rows would lose the same people's other valid answers.

```{r}
#| label: Special Codes Check

D02 |> count(education_level_code, sort = TRUE) |> filter(education_level_code %in% c(98, 99, 29, 28))

D02 |> count(education_attendance_code)

D02 |> count(institution_type_code == 88, name = "rows")

D02 |> count(reason_not_attending_code == 88, name = "rows")

D08 |> count(fixed_internet_code)

D08 |> count(mobile_internet_code)

D08 |> count(computer_asset_code)
```

**Treatment:** classify each special code as a valid category, a "Not applicable" level or `NA`, as listed in the table below. The classification is applied in section 6 when the label columns are created: codes left out of `factor()`'s levels become `NA`.

- Codes 98 (No schooling) and 28 (Other) in `education_level_code` are real education levels and stay valid categories.
- Codes 29 and 3 mean "do not know", and `computer_asset_code` 9 means "Unspecified". These are non-response, not categories, so they become `NA`.
- Code 99 in `education_level_code` and code 9 in `fixed_internet_code` and `mobile_internet_code` are not in the data dictionary, so they cannot be given a meaning and become `NA`.
- Code 88 in `institution_type_code` and `reason_not_attending_code` means the question did not apply to the person (for example, a reason for not attending is only asked of people who do not attend). It becomes a "Not applicable" level, not missing.
- Attendance code 8 only occurs under age 15 and disappears after record selection (section 5).

```{r}
#| label: Special Code Classification

tibble(
  
  column = c("education_level_code", "education_level_code", "education_level_code", "education_level_code", "education_attendance_code", "education_attendance_code", "institution_type_code", "reason_not_attending_code", "fixed_internet_code", "mobile_internet_code", "computer_asset_code"),

  code = c(98, 28, 29, 99, 3, 8, 88, 88, 9, 9, 9),

  meaning = c("No schooling", "Other", "Do not know", "Not in the data dictionary", "Do not know", "Not applicable (only under age 15)", "Not applicable", "Not applicable", "Not in the data dictionary", "Not in the data dictionary", "Unspecified"),

  classification = c("Valid category", "Valid category", "NA (non-response)", "NA (undocumented code)", "NA (non-response)", "Removed by record selection", "'Not applicable' level", "'Not applicable' level", "NA (undocumented code)", "NA (undocumented code)", "NA (non-response)"),

  rows = c(sum(D02$education_level_code %in% 98), sum(D02$education_level_code %in% 28), sum(D02$education_level_code %in% 29), sum(D02$education_level_code %in% 99), sum(D02$education_attendance_code %in% 3), sum(D02$education_attendance_code %in% 8), sum(D02$institution_type_code %in% 88), sum(D02$reason_not_attending_code %in% 88), sum(D08$fixed_internet_code %in% 9), sum(D08$mobile_internet_code %in% 9), sum(D08$computer_asset_code %in% 9))
  
)
```

Two groups of values were already `NA` before this section. They stay `NA` because there is no dictionary code to assign them:

- The `r sum(is.na(D02$education_attendance_code))` missing values in `education_attendance_code` are not original gaps, because the uncleaned column had no missing values. They are the text placeholders ("N/A", "Unknown", "not recorded") set to missing and flagged in 2.3.
- The `r sum(is.na(D08$mobile_internet_code))` missing values in `mobile_internet_code` were already missing in the uncleaned data.

**Effect / Verification:** no values are changed in this section, and every `_code` column keeps its original codes. The classification takes effect in section 6. The "Codes Without a Label" table there verifies it: the rows left without a label are exactly the rows holding the codes classified as `NA` here, after record selection.

## 4.2 Salary placeholders

**Problem:** `lab_salary_raw` mixes real Rand values with two placeholder codes, 888888888 and 999999999. Left in place, they would be read as salaries of almost R1 billion. Their meaning also has to be checked against employment status before deciding how to treat them.

```{r}
#| label: Salary Placeholder Cross-Tab

D03 |>
  mutate(salary_type = case_when(
    lab_salary_raw == 888888888 ~ "888888888 (don't know/refused)",
    lab_salary_raw == 999999999 ~ "999999999 (not applicable)",
    TRUE ~ "Real salary value"
  )) |>
  count(employment_status2code, salary_type) |>
  pivot_wider(names_from = employment_status2code, values_from = n, values_fill = 0)

```

**Treatment:** the cross-table shows that 888888888 acts as "not applicable" for everyone without a job, and 999999999 as non-response among the employed. Both placeholders become `NA` in a new `salary` column, and a new `salary_status` column records why a salary is missing. It is derived from the value and `employment_status2code` together, not from the placeholder alone. `lab_salary_raw` is kept unchanged.

```{r}
#| label: Salary Reversal

D03_salary_missing_before <- sum(D03$lab_salary_raw %in% c(888888888, 999999999))

D03 <- D03 |>
  mutate(
    salary = if_else(lab_salary_raw %in% c(888888888, 999999999), NA_real_, lab_salary_raw),
    salary_status = case_when(
!is.na(salary) ~ "Reported", lab_salary_raw == 999999999 ~ "Non-response (employed)", lab_salary_raw == 888888888 &
employment_status2code == 1 ~ "Non-response (employed)", lab_salary_raw == 888888888 ~ "Not applicable (not employed)", TRUE ~ NA_character_
    )
  )

```

**Effect:** `r format(sum(!is.na(D03$salary) & D03$employment_status2code == 1), big.mark = ",")` of `r format(sum(D03$employment_status2code == 1), big.mark = ",")` employed people report a real salary (`r round(mean(!is.na(D03$salary[D03$employment_status2code == 1])) * 100)`%). This corrects Milestone 1's "74% unusable" figure, which counted people who cannot have a salary in the first place.

**Verification:**

```{r}
#| label: Salary Reversal Verification

D03 |> count(salary_status, has_salary = !is.na(salary))

tibble(
  
  placeholders_before = D03_salary_missing_before,
  placeholders_remaining_in_salary = sum(D03$salary %in% c(888888888, 999999999), na.rm = TRUE)
  
)
```

# 5. Record Selection (ages 15–64)

**Problem:** the analysis population is the working-age population, ages 15–64.

**Treatment:** filter D01 on `between(age, 15, 64)`, then keep only matching people in D02 and D03 with `semi_join()`. All D08 households are kept; Task 3 reports the households with no working-age member. The number of people in each household is counted before the filter, because the smartphone check in section 7 needs the full household size, not just the working-age members.

```{r}
#| label: Record Selection

D01_rows_before_selection <- nrow(D01)
D01_under_15 <- sum(D01$age < 15)
D01_65_plus <- sum(D01$age > 64)

#All members of each household, counted before the age filter (used in section 7)
household_size <- D01 |> count(household_id, name = "household_size")

D01 <- D01 |> filter(between(age, 15, 64))
D02 <- semi_join(D02, D01, by = c("household_id", "person_number"))
D03 <- semi_join(D03, D01, by = c("household_id", "person_number"))
```

**Verification:**

```{r}
#| label: Record Selection Verification

tibble(
  
  dataset = c("D01", "D02", "D03"),
  
  rows_before = D01_rows_before_selection,
  
  rows_after = c(nrow(D01), nrow(D02), nrow(D03)),
  
  removed_under_15 = D01_under_15,
  
  removed_65_plus = D01_65_plus,
  
  distinct_households = c(n_distinct(D01$household_id), n_distinct(D02$household_id), n_distinct(D03$household_id)),
  
  identical_keys_to_D01 = c(TRUE, setequal(paste(D01$household_id, D01$person_number), paste(D02$household_id, D02$person_number)), setequal(paste(D01$household_id, D01$person_number), paste(D03$household_id, D03$person_number)))
  
)
```

**Effect:** all three person files have `r format(nrow(D01), big.mark = ",")` rows and identical keys, covering `r format(n_distinct(D01$household_id), big.mark = ",")` households. `r format(D01_under_15, big.mark = ",")` people under 15 and `r format(D01_65_plus, big.mark = ",")` aged 65+ were removed.

# 6. Data Transformation: recoding into labelled factors

**Problem:** `_code` columns are not analytically meaningful or safe to treat as numbers. A model would treat province 9 as "more" than province 1, and a report reader cannot tell what code 5 means without the data dictionary.

**Treatment:** for every kept `_code` column, add a label column of the same name without the suffix, using `factor(..., levels =, labels =)` with the labels from the data dictionary. Codes left out of `levels =` become `NA`, applying the classification from section 4. `_code` columns are never modified, so every label can be traced back to its original code.

```{r}
#| label: Recode D01

D01 <- D01 |>
  mutate(
    province = factor(province_code, levels = 1:9, labels = c("Western Cape", "Eastern Cape", "Northern Cape", "Free State", "KwaZulu-Natal", "North West", "Gauteng", "Mpumalanga", "Limpopo")),
    
    sex = factor(sex_code, levels = 1:2, labels = c("Male", "Female")),
    
    population_group = factor(population_group_code, levels = 1:4, labels = c("Black African", "Coloured", "Indian/Asian", "White")),
    
    geo_type = factor(geo_type_code, levels = 1:3, labels = c("Urban", "Traditional", "Farms")),
    
    metro = factor(metro_indicator, levels = 1:2, labels = c("Metro", "Non-metro"))
    
  )
```

```{r}
#| label: Recode D02

D02 <- D02 |>
  mutate(
    education_level =
      factor(education_level_code, levels = c(98, 0:27, 28),
         labels = c(
             "No schooling",
             "Grade R/0", "Grade 1/Sub A/Class 1", "Grade 2/Sub B/Class 2",
             "Grade 3/Standard 1/ABET/AET 1", "Grade 4/Standard 2",
             "Grade 5/Standard 3/ABET/AET 2", "Grade 6/Standard 4",
             "Grade 7/Standard 5/ABET/AET 3", "Grade 8/Standard 6/Form 1",
             "Grade 9/Standard 7/Form 2/ABET/AET 4/NCV 1",
             "Grade 10/Standard 8/Form 3/NCV 2",
             "Grade 11/Standard 9/Form 4/NCV 3",
             "Grade 12/Standard 10/Form 5/NSC/Matric/NCV 4",
             "NTC I/N1/NQF 1", "NTC II/N2/NQF 2", "NTC III/N3/NQF 3",
             "N4/NTC 4/Occupational Cert NQF 5", "N5/NTC 5/Occupational Cert NQF 5",
             "N6/NTC 6/Occupational Cert NQF 5",
             "Certificate less than Grade 12", "Diploma less than Grade 12",
             "Higher/National/Advanced Certificate with Grade 12/NQF 5",
             "Diploma with Grade 12/Occupational Cert NQF 6",
             "Higher Diploma/B-Tech/Occupational Cert NQF 7",
             "Bachelor's Degree/Occupational Cert NQF 7",
             "Honours Degree/Postgrad Diploma/Occupational Cert NQF 8",
             "Post Higher Diploma (M-Tech/Master's) NQF 9",
             "Doctoral Degree (D-Tech/PhD) NQF 10",
             "Other")),
    
    education_attendance = factor(education_attendance_code, levels = 1:2, labels = c("Yes", "No")),
    
    institution_type = factor(institution_type_code, levels = c(1:9, 88),
         labels = c(
             "Pre-school (incl. ECD centre)",
             "School (Grade R-12)",
             "ABET/AET centre",
             "Literacy classes",
             "Higher education institution (university / university of technology)",
             "TVET college",
             "Other college",
             "Home-based education / home schooling",
             "Other educational institution",
             "Not applicable")),
    
    reason_not_attending = factor(reason_not_attending_code, levels = c(1:18, 88),
         labels = c(
             "Too old/young",
             "Completed education / satisfied / doesn't want to study",
             "Institution too far",
             "Difficulty getting there (transport)",
             "No money for fees",
             "Working (at home, business or job)",
             "Family commitment (e.g. child minding)",
             "Education useless / not interesting",
             "Unable to perform at school",
             "Illness",
             "Pregnancy",
             "Failed exams",
             "Got married",
             "Disability",
             "Violence in school",
             "Not accepted for enrolment",
             "No time / too busy",
             "Other reason",
             "Not applicable"))
    
  )
```

```{r}
#| label: Recode D03

D03 <- D03 |>
  mutate(
    employment_status = factor(employment_status2code, levels = 1:3, labels = c("Employed", "Unemployed", "Not economically active"))
  )

```

```{r}
#| label: Recode D08

D08 <- D08 |>
  mutate(
    fixed_internet = factor(fixed_internet_code, levels = 1:2, labels = c("Yes", "No")),
    mobile_internet = factor(mobile_internet_code, levels = 1:2, labels = c("Yes", "No")),
    computer_asset = factor(computer_asset_code, levels = 1:2, labels = c("Yes", "No"))
  )
```

**Verification:** every code maps to exactly one label, and only the codes classified as non-response in section 4 have no label.

```{r}
#| label: Recode Verification

D01 |> count(province_code, province)
D01 |> count(sex_code, sex)
D01 |> count(population_group_code, population_group)
D01 |> count(geo_type_code, geo_type)
D01 |> count(metro_indicator, metro)

D02 |> count(education_level_code, education_level)
D02 |> count(education_attendance_code, education_attendance)
D02 |> count(institution_type_code, institution_type)
D02 |> count(reason_not_attending_code, reason_not_attending)

D03 |> count(employment_status2code, employment_status)

D08 |> count(fixed_internet_code, fixed_internet)
D08 |> count(mobile_internet_code, mobile_internet)
D08 |> count(computer_asset_code, computer_asset)
```

```{r}
#| label: Codes Without a Label

#Rows with a code but no label, compared with the number of rows holding the non-response codes from section 4
tibble(
  
  label_column = c("education_level", "education_attendance", "fixed_internet", "mobile_internet", "computer_asset"),

  codes_set_to_NA = c("29, 99", "3", "9", "9", "9"),

  rows_with_these_codes = 
    c(sum(D02$education_level_code %in% c(29, 99)), sum(D02$education_attendance_code %in% 3), sum(D08$fixed_internet_code %in% 9), sum(D08$mobile_internet_code %in% 9), sum(D08$computer_asset_code %in% 9)),

  rows_with_code_but_no_label = 
    c(sum(!is.na(D02$education_level_code) & is.na(D02$education_level)), sum(!is.na(D02$education_attendance_code) & is.na(D02$education_attendance)), sum(!is.na(D08$fixed_internet_code) & is.na(D08$fixed_internet)), sum(!is.na(D08$mobile_internet_code) & is.na(D08$mobile_internet)), sum(!is.na(D08$computer_asset_code) & is.na(D08$computer_asset)))
  
)
```

**Effect:** 13 label columns are added (5 in D01, 4 in D02, 1 in D03 and 3 in D08) and no `_code` column is changed. The two columns in the table above always match, so the only codes left without a label are the non-response and undocumented codes classified in section 4: `r sum(D02$education_level_code %in% c(29, 99))` education levels, `r sum(D02$education_attendance_code %in% 3)` attendance values, `r sum(D08$fixed_internet_code %in% 9)` fixed internet, `r sum(D08$mobile_internet_code %in% 9)` mobile internet and `r sum(D08$computer_asset_code %in% 9)` computer asset values. `institution_type` and `reason_not_attending` now use the data dictionary's descriptions, and code 88 is kept as "Not applicable" rather than missing.

The levels of `education_level` follow the dictionary order, which is not a ranking: "No schooling" comes first and "Other" comes last. Education bands built for the analysis in Task 3 must therefore group explicit codes, not rely on the level order.

# 7. Outlier Review

`age`, `household_smartphones` and `salary` are the only kept columns that are real measurements rather than categories, so they are the only columns checked for outliers and impossible values. Each is shown as a histogram and a boxplot. The boxplot's whiskers end at the fences, 1.5 × the interquartile range (IQR) beyond the quartiles, and points outside the fences are drawn separately as statistical outliers. A statistical outlier is only a candidate: it is kept unless domain evidence shows the value is impossible.

## 7.1 Age

**Problem:** extreme ages could distort any analysis by age.

```{r}
#| label: Age Outlier Check

summary(D01$age)

ggplot(D01, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Age distribution after record selection", x = "Age", y = "Count")

ggplot(D01, aes(x = age)) +
  geom_boxplot() +
  labs(title = "Age after record selection (boxplot)", x = "Age", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

#Boxplot fences: 1.5 × IQR beyond the quartiles
age_boxplot <- tibble(

  q1 = quantile(D01$age, 0.25),

  median = median(D01$age),

  q3 = quantile(D01$age, 0.75),

  lower_fence = q1 - 1.5 * (q3 - q1),

  upper_fence = q3 + 1.5 * (q3 - q1),

  points_below_fence = sum(D01$age < lower_fence),

  points_above_fence = sum(D01$age > upper_fence)
)

age_boxplot
```

**Treatment:** none. Invalid ages were already repaired in 2.4, and record selection (section 5) limits age to 15–64.

**Effect / Verification:** the summary shows a minimum of `r min(D01$age)` and a maximum of `r max(D01$age)`. The histogram shows no isolated values, and the boxplot has `r age_boxplot$points_below_fence + age_boxplot$points_above_fence` points outside its fences (`r age_boxplot$lower_fence` to `r age_boxplot$upper_fence`), so there are no age outliers left to treat.

## 7.2 Household smartphones

**Problem:** a household cannot own a negative number of smartphones, but D08 contains `r sum(D08$household_smartphones < 0, na.rm = TRUE)` negative counts. Very high counts also need to be checked against the size of the household.

```{r}
#| label: Smartphones Outlier Check

summary(D08$household_smartphones)

ggplot(D08, aes(x = household_smartphones)) +
  geom_boxplot() +
  labs(title = "Household smartphones before treatment (boxplot)", x = "Smartphones in the household", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

#Boxplot fences: 1.5 × IQR beyond the quartiles
smartphones_boxplot <- tibble(

  q1 = quantile(D08$household_smartphones, 0.25, na.rm = TRUE),

  median = median(D08$household_smartphones, na.rm = TRUE),

  q3 = quantile(D08$household_smartphones, 0.75, na.rm = TRUE),

  lower_fence = q1 - 1.5 * (q3 - q1),

  upper_fence = q3 + 1.5 * (q3 - q1),

  points_below_fence = sum(D08$household_smartphones < lower_fence, na.rm = TRUE),

  points_above_fence = sum(D08$household_smartphones > upper_fence, na.rm = TRUE)
)

smartphones_boxplot

D08 |> filter(household_smartphones < 0) |> count(household_smartphones)

D08 |>
  left_join(household_size, by = "household_id") |>
  filter(household_smartphones >= 10) |>
  select(household_id, household_smartphones, household_size)
```

In the boxplot, the `r smartphones_boxplot$points_below_fence` points below the lower fence are exactly the negative counts. Most households have 1 or 2 smartphones, so the box is narrow (an IQR of `r smartphones_boxplot$q3 - smartphones_boxplot$q1`). As a result, the 1.5 × IQR rule also marks every household with `r floor(smartphones_boxplot$upper_fence) + 1` or more smartphones as an outlier: `r format(smartphones_boxplot$points_above_fence, big.mark = ",")` households. These high points are a result of the narrow box, not evidence of errors, so they are checked against household size instead of being removed.

The negative values cannot be repaired: no other kept dataset records smartphones, and removing the minus sign is not justified, because `r sum(D08$household_smartphones < -13, na.rm = TRUE)` of them would then exceed the dictionary maximum of 13. At the other end, every household with 10 or more smartphones has at least as many members as smartphones. `household_size` counts all members, including those outside the 15–64 range, because it was counted before record selection (section 5).

**Treatment:** set the negative counts to `NA` and record the original value in `household_smartphones_flag`. The high counts are plausible and kept unchanged. The household is kept, since its other connectivity columns are still valid.

```{r}
#| label: Fixing Negative Smartphone Counts

D08_negative_smartphones_before <- sum(D08$household_smartphones < 0, na.rm = TRUE)

D08 <- D08 |>
  mutate(
    
    household_smartphones_flag = if_else(household_smartphones < 0, paste0("invalid negative count (", household_smartphones, ") set to missing"), NA_character_),
    
    household_smartphones = if_else(household_smartphones < 0, NA_real_, household_smartphones)
    
  )

```

**Verification:**

```{r}
#| label: Smartphones Verification

D08 |>
  left_join(household_size, by = "household_id") |>
  summarise(
    
    negative_before = D08_negative_smartphones_before,

    negative_after = sum(household_smartphones < 0, na.rm = TRUE),

    missing_after = sum(is.na(household_smartphones)),

    rows_flagged = sum(!is.na(household_smartphones_flag)),

    min_after = min(household_smartphones, na.rm = TRUE),

    max_after = max(household_smartphones, na.rm = TRUE),

    households_with_more_smartphones_than_members = sum(household_smartphones > household_size, na.rm = TRUE)

  )

ggplot(D08, aes(x = household_smartphones)) +
  geom_boxplot() +
  labs(title = "Household smartphones after treatment (boxplot)", x = "Smartphones in the household", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

**Effect:** the `r D08_negative_smartphones_before` impossible counts are now `NA` and traceable through `household_smartphones_flag`, and no household is removed. After treatment, the boxplot has no points below the box. The remaining counts lie within the dictionary range of 0–13, and no household has more smartphones than members, including the `r format(smartphones_boxplot$points_above_fence, big.mark = ",")` households above the upper fence, which are kept unchanged. `household_size` is not joined onto D08 or saved; it is only used for this check.

## 7.3 Salary

**Problem:** reported salaries range from very small to very large values, and a few extreme values could distort averages.

```{r}
#| label: Salary Outlier Check

D03 |> filter(!is.na(salary)) |> summarise(
  min = min(salary), p1 = quantile(salary, 0.01), median = median(salary),
  p99 = quantile(salary, 0.99), max = max(salary)
)

D03 |>
  filter(!is.na(salary)) |>
  ggplot(aes(x = salary)) +
  geom_histogram(bins = 50) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Reported salary (log scale)", x = "Salary (Rand, log scale)", y = "Count")

#Drawn on the log scale like the histogram; on the Rand scale the box is squashed against zero
D03 |>
  filter(!is.na(salary)) |>
  ggplot(aes(x = salary)) +
  geom_boxplot() +
  geom_vline(xintercept = 100, linetype = "dashed") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Reported salary (boxplot, log scale)", subtitle = "Dashed line: R100 threshold used for salary_flag",
       x = "Salary (Rand, log scale)", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

#Boxplot fences on the log scale, converted back to Rand
salary_boxplot <- D03 |>
  filter(!is.na(salary)) |>
  summarise(

    log_q1 = quantile(log10(salary), 0.25),

    log_q3 = quantile(log10(salary), 0.75),

    lower_fence_rand = 10^(log_q1 - 1.5 * (log_q3 - log_q1)),

    upper_fence_rand = 10^(log_q3 + 1.5 * (log_q3 - log_q1)),

    points_below_fence = sum(salary < lower_fence_rand),

    points_above_fence = sum(salary > upper_fence_rand),

    of_which_under_R100 = sum(salary < 100)
  ) |>
  select(-log_q1, -log_q3)

salary_boxplot
```

On the log scale, the boxplot marks `r salary_boxplot$points_below_fence` salaries below about R`r format(round(salary_boxplot$lower_fence_rand), big.mark = ",")` and `r salary_boxplot$points_above_fence` above about R`r format(round(salary_boxplot$upper_fence_rand), big.mark = ",")` as outliers.

**Treatment:** the high outliers, up to R`r format(max(D03$salary, na.rm = TRUE), big.mark = ",")`, are genuine high earners and are kept. The low outliers are not all errors: values between R100 and about R`r format(round(salary_boxplot$lower_fence_rand), big.mark = ",")` are low but possible for casual or part-time work, and are kept unflagged. Only the `r salary_boxplot$of_which_under_R100` values under R100 cannot plausibly be a monthly salary. They are flagged in `salary_flag`, not deleted, since the pay period is not documented in the source data. The flag uses the fixed R100 threshold, not the boxplot fence, so it does not depend on the rest of the data.

```{r}
#| label: Flagging Low Salary Outliers

D03 <- D03 |>
  mutate(salary_flag = if_else(!is.na(salary) & salary < 100, "Implausibly low for a monthly salary", NA_character_))
```

**Verification:**

```{r}
#| label: Salary Flag Verification

D03 |>
  filter(!is.na(salary)) |>
  summarise(
    reported_salaries = n(),
    
    flagged_under_R100 = sum(!is.na(salary_flag)),
    
    largest_flagged_salary = max(salary[!is.na(salary_flag)]),
    
    smallest_unflagged_salary = min(salary[is.na(salary_flag)])
)

```

**Effect:** `r sum(!is.na(D03$salary_flag))` of `r format(sum(!is.na(D03$salary)), big.mark = ",")` reported salaries are flagged and every salary value is unchanged. Because the threshold is fixed, later analyses can include or exclude these rows with `salary < 100`, even without the flag column.

`person_weight` is not touched: survey weights must never be trimmed or scaled. No scaling or transformation is applied during preprocessing; that belongs in Task 4, after the train/test split, to avoid leakage. The log scale above is for plotting only.

# 8. Validation and Save of the Cleaned Datasets

## 8.1 Validation of the cleaned datasets

**Problem:** the cleaned datasets will be used in every later task, so before saving them it must be shown that the pipeline as a whole did what the earlier sections claim. People must only have been removed by record selection, keys must still be unique, no duplicates may remain, and every remaining missing value must have a reason. The rubric also requires before-and-after evidence for every major preprocessing decision.

**Treatment:** compare each cleaned dataset with the uncleaned summary recorded in 1.2, list the columns added or changed in type, and give a reason for every column that still has missing values.

**Verification:**

```{r}
#| label: Before vs After Preprocessing

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows_uncleaned = uncleaned_summary$rows,
  
  rows_cleaned = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  cols_uncleaned = uncleaned_summary$cols,
  
  cols_cleaned = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08)),
  
  missing_values_uncleaned = uncleaned_summary$missing_values,
  
  #_flag columns are left out: they are empty by design for every row that was not changed
  missing_values_cleaned = c(sum(is.na(select(D01, -ends_with("_flag")))), sum(is.na(select(D02, -ends_with("_flag")))), sum(is.na(select(D03, -ends_with("_flag")))), sum(is.na(select(D08, -ends_with("_flag"))))),
  
  distinct_keys_uncleaned = uncleaned_summary$distinct_keys,
  
  distinct_keys_cleaned = c(nrow(distinct(D01, household_id, person_number)), nrow(distinct(D02, household_id, person_number)), nrow(distinct(D03, household_id, person_number)),
                            
  n_distinct(D08$household_id)),
  
  exact_duplicates_uncleaned = uncleaned_summary$exact_duplicate_rows,
  
  exact_duplicates_cleaned = c(nrow(D01) - nrow(distinct(D01)), nrow(D02) - nrow(distinct(D02)), nrow(D03) - nrow(distinct(D03)), nrow(D08) - nrow(distinct(D08)))
  
)
```

**Note:** `missing_values_cleaned` leaves out the `_flag` columns, which are empty by design for every row that was not changed. D01 has no missing values left. The cleaned counts for D02, D03 and D08 are higher than the uncleaned ones, but this is not new data loss: placeholders, non-response codes and impossible values were stored as ordinary values in the uncleaned data, so the uncleaned count did not see them. They are now honest `NA`s. In D03, `salary` is `NA` for the placeholders reversed in 4.2, and nearly all of these belong to people who are not employed and cannot have a salary. In D02 and D08 the extra `NA`s come from the text placeholders set to missing in 2.3, the negative smartphone counts in 7.2, and the label columns from section 6, which repeat the missing values of their `_code` column and leave the non-response codes without a label. The Final Missingness table below gives the reason for every column.

Columns added or changed in type since column selection (section 3):

```{r}
#| label: Types Before and After Sections 4 to 7

bind_rows(
  D01 = enframe(D01_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D01, class), name = "column", value = "type_after"), by = "column"),
  
  D02 = enframe(D02_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D02, class), name = "column", value = "type_after"), by = "column"),
  
  D03 = enframe(D03_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D03, class), name = "column", value = "type_after"), by = "column"),
  
  D08 = enframe(D08_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D08, class), name = "column", value = "type_after"), by = "column"),
  .id = "dataset"
) |>
  filter(is.na(type_before) != is.na(type_after) | (!is.na(type_before) & !is.na(type_after) & type_before != type_after))
```

```{r}
#| label: Final Missingness

final_missingness <- bind_rows(
  D01 = enframe(colSums(is.na(D01)), name = "column", value = "missing"),
  D02 = enframe(colSums(is.na(D02)), name = "column", value = "missing"),
  D03 = enframe(colSums(is.na(D03)), name = "column", value = "missing"),
  D08 = enframe(colSums(is.na(D08)), name = "column", value = "missing"),
  .id = "dataset"
) |>
  filter(missing > 0) |>
  mutate(
    reason = case_when(
    column == "education_level" ~ "Codes 29/99 reclassified to NA in sections 4/6",
    
    column == "education_attendance_code" ~ "Text placeholders ('N/A', 'Unknown', 'not recorded') set to NA and flagged in 2.3",
    
    column == "education_attendance" ~ "Code 3 reclassified to NA in sections 4/6, plus the text placeholders set to NA in 2.3",
    
    column == "fixed_internet" ~ "Code 9 reclassified to NA in sections 4/6",
    
    column == "mobile_internet_code" ~ "Already missing in the uncleaned data; no dictionary code to assign (noted in section 4)",
    
    column == "mobile_internet" ~ "Code 9 reclassified to NA in sections 4/6, plus the missing mobile_internet_code values",
    
    column == "computer_asset" ~ "Code 9 ('Unspecified') reclassified to NA in sections 4/6",
    
    column == "household_smartphones" ~ "Negative counts set to NA and flagged in 7.2",
    
    str_detect(column, "_flag$") ~ "Flags only filled in for rows that changed (expected)",
    
    column == "salary" ~ "Placeholder codes reversed to NA in section 4",
    
    column == "salary_status" ~ "Could not be classified from value and employment status",
    
    TRUE ~ "NOT CLASSIFIED"
  ))

final_missingness
```

```{r}
#| label: Final Category Levels

list(
  
  province = levels(D01$province), sex = levels(D01$sex), population_group = levels(D01$population_group),
  
  geo_type = levels(D01$geo_type), metro = levels(D01$metro),
  
  education_level = levels(D02$education_level), education_attendance = levels(D02$education_attendance),
  
  institution_type = levels(D02$institution_type), reason_not_attending = levels(D02$reason_not_attending),
  
  employment_status = levels(D03$employment_status),
  
  fixed_internet = levels(D08$fixed_internet), mobile_internet = levels(D08$mobile_internet),
  
  computer_asset = levels(D08$computer_asset)
)
```

**Effect:** D01 goes from `r format(uncleaned_summary$rows[1], big.mark = ",")` to `r format(nrow(D01), big.mark = ",")` people. The difference is exactly the `r format(D01_under_15 + D01_65_plus, big.mark = ",")` people outside ages 15–64 removed in section 5, so no one was lost to cleaning. D02 and D03 match D01 after also losing their `r D02_exact_duplicates` and `r D03_exact_duplicates` exact duplicate rows (2.2), and D08 keeps all `r format(nrow(D08), big.mark = ",")` households. In every dataset each key appears exactly once and no exact duplicates remain. `r sum(final_missingness$reason == "NOT CLASSIFIED")` of the `r nrow(final_missingness)` columns with missing values lack a documented reason.

## 8.2 Save

**Problem:** the cleaned datasets must be saved so that later tasks can use them without rerunning this pipeline. The original files in `Datasets/` must not be changed.

**Treatment:** save each cleaned dataset as a new `.csv` (required deliverable) in `Datasets/Cleaned/`. The original files are only read, never written. Missing values are written as empty cells rather than the text "NA", so that Power BI reads them as blanks instead of turning numeric columns such as `salary` into text. A CSV does not store column types or the order of factor levels. Anyone reading the CSVs, including Power BI, must therefore read `household_id` and `person_number` as text, and re-apply the level order listed above where the order matters.

The `_flag` columns are left out of the saved files. Inside this pipeline they stay in the datasets, so every change remains traceable in the report and in the integration steps (section 9). They are removed from the analytical data in Task 4, because they are not predictors. The saved files contain only the analytical columns: the original codes, the cleaned values and the labels.

Saving at this point, before the train/test split in Task 4, does not leak information between the future training and test rows. Every change so far either applies a fixed rule to each row (dictionary codes, the 15–64 age range, the R100 salary threshold, negative counts) or copies a value from the same person's record in another dataset. Nothing is estimated from the data as a whole: no means, quantiles, imputation models or scaling are used to change a value. Steps that learn from the data belong after the split, where they are fitted on the training rows only.

```{r}
#| label: Save Cleaned Datasets

#na = "" writes missing values as empty cells, which both R and Power BI read as missing
#select(-ends_with("_flag")) leaves the traceability flags out of the saved files; they stay in the datasets in memory
write_csv(select(D01, -ends_with("_flag")), file.path(cleaned_directory, "D01_clean.csv"), na = "")

write_csv(select(D02, -ends_with("_flag")), file.path(cleaned_directory, "D02_clean.csv"), na = "")

write_csv(select(D03, -ends_with("_flag")), file.path(cleaned_directory, "D03_clean.csv"), na = "")

write_csv(select(D08, -ends_with("_flag")), file.path(cleaned_directory, "D08_clean.csv"), na = "")
```

**Verification:** read the saved files back with the IDs as text, the same way later tasks should read them. Compare them with the datasets in memory without their flag columns, and confirm that no flag column was saved.

```{r}
#| label: Saved Files Verification

D01_saved <- read_csv(file.path(cleaned_directory, "D01_clean.csv"), show_col_types = FALSE, col_types = cols(household_id = col_character(), person_number = col_character()))

D02_saved <- read_csv(file.path(cleaned_directory, "D02_clean.csv"), show_col_types = FALSE, col_types = cols(household_id = col_character(), person_number = col_character()))

D03_saved <- read_csv(file.path(cleaned_directory, "D03_clean.csv"), show_col_types = FALSE, col_types = cols(household_id = col_character(), person_number = col_character()))

D08_saved <- read_csv(file.path(cleaned_directory, "D08_clean.csv"), show_col_types = FALSE, col_types = cols(household_id = col_character()))

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),

  rows_in_memory = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),

  rows_in_saved_file = c(nrow(D01_saved), nrow(D02_saved), nrow(D03_saved), nrow(D08_saved)),

  flag_columns_in_saved_file = c(sum(str_ends(names(D01_saved), "_flag")), sum(str_ends(names(D02_saved), "_flag")),
                                 sum(str_ends(names(D03_saved), "_flag")), sum(str_ends(names(D08_saved), "_flag"))),

  same_columns_apart_from_flags = c(identical(names(D01_saved), names(select(D01, -ends_with("_flag")))),
                                    identical(names(D02_saved), names(select(D02, -ends_with("_flag")))),
                                    identical(names(D03_saved), names(select(D03, -ends_with("_flag")))),
                                    identical(names(D08_saved), names(select(D08, -ends_with("_flag"))))),

  ids_identical = c(identical(D01_saved$household_id, D01$household_id) & identical(D01_saved$person_number, D01$person_number), identical(D02_saved$household_id, D02$household_id) & identical(D02_saved$person_number, D02$person_number), identical(D03_saved$household_id, D03$household_id) & identical(D03_saved$person_number, D03$person_number), identical(D08_saved$household_id, D08$household_id)),

  missing_values_identical = c(sum(is.na(D01_saved)) == sum(is.na(select(D01, -ends_with("_flag")))),
                               sum(is.na(D02_saved)) == sum(is.na(select(D02, -ends_with("_flag")))),
                               sum(is.na(D03_saved)) == sum(is.na(select(D03, -ends_with("_flag")))),
                               sum(is.na(D08_saved)) == sum(is.na(select(D08, -ends_with("_flag")))))

)
```

**Effect:** four new files are written to `Datasets/Cleaned/`. These four `write_csv()` calls are the only place the pipeline writes data, so the original files in `Datasets/` are left unchanged. Read back with the IDs as text, no saved file contains a flag column. Apart from the flags, every saved file has the same rows, columns, IDs and missing values as the cleaned dataset in memory, so later tasks can start from these files instead of rerunning the pipeline.

# 9. Data Integration

The analytical grain chosen in the Task 1 write-up is one row per working-age person (15–64), identified by `household_id` + `person_number`. Section 1.4 explains why only D01, D02, D03 and D08 are used. They are joined here, after cleaning, because the keys are only reliable once the missing IDs are recovered (2.1) and the duplicates removed (2.2), and the population is only final after record selection (section 5). D01 is the base table, because it now defines exactly this population. This section confirms the key and cardinality of each join before it runs, and verifies the row counts and unmatched records before and after every join.

## 9.1 Keys and cardinality before joining

**Problem:** a join duplicates rows if the key is not unique on the side that should match only once, and it silently drops or misaligns people if keys are missing or do not match. The expected cardinality of every join must be confirmed before the join is run.

```{r}
#| label: Key Checks Before Joining

tibble(

  dataset = c("D01", "D02", "D03", "D08"),

  key = c("household_id + person_number", "household_id + person_number", "household_id + person_number", "household_id"),

  rows = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),

  distinct_keys = c(nrow(distinct(D01, household_id, person_number)), nrow(distinct(D02, household_id, person_number)),
                    nrow(distinct(D03, household_id, person_number)), n_distinct(D08$household_id)),

  missing_keys = c(sum(is.na(D01$household_id) | is.na(D01$person_number)), sum(is.na(D02$household_id) | is.na(D02$person_number)),
                   sum(is.na(D03$household_id) | is.na(D03$person_number)), sum(is.na(D08$household_id))),

  all_ids_18_digits = c(all(nchar(D01$household_id) == 18), all(nchar(D02$household_id) == 18),
                        all(nchar(D03$household_id) == 18), all(nchar(D08$household_id) == 18))
)
```

```{r}
#| label: Expected Cardinality and Unmatched Keys

#Households in D08 with no working-age member, and whether they had members before record selection (section 5)
D08_households_without_D01 <- anti_join(D08, D01, by = "household_id")

tibble(

  join = c("D01 + D02", "D01 + D03", "D01 + D08"),

  expected_cardinality = c("one-to-one", "one-to-one", "many-to-one"),

  most_D01_rows_per_key = c(max(count(D01, household_id, person_number)$n), max(count(D01, household_id, person_number)$n),
                            max(count(D01, household_id)$n)),

  most_rows_per_key_in_joined_dataset = c(max(count(D02, household_id, person_number)$n), max(count(D03, household_id, person_number)$n),
                                          max(count(D08, household_id)$n)),

  D01_rows_without_a_match = c(nrow(anti_join(D01, D02, by = c("household_id", "person_number"))),
                               nrow(anti_join(D01, D03, by = c("household_id", "person_number"))),
                               nrow(anti_join(D01, D08, by = "household_id"))),

  joined_dataset_rows_without_a_match = c(nrow(anti_join(D02, D01, by = c("household_id", "person_number"))),
                                          nrow(anti_join(D03, D01, by = c("household_id", "person_number"))),
                                          nrow(D08_households_without_D01))
)

tibble(
  D08_households_without_a_working_age_member = nrow(D08_households_without_D01),

  of_which_had_members_before_record_selection = sum(D08_households_without_D01$household_id %in% household_size$household_id)
)
```

**Treatment:** D01 is the left table, so every working-age person is kept. D02 and D03 are joined one-to-one on `household_id` + `person_number`. D08 is joined many-to-one on `household_id`, so each person receives their own household's connectivity values. Each join states its expected cardinality in dplyr's `relationship` argument, which stops the pipeline with an error if any key matches more rows than expected. The D02 and D03 joins also use `unmatched = "error"`, which stops the pipeline if any of their rows would be dropped. D08 cannot use `unmatched = "error"`, because its unmatched households are expected (see below).

**Effect / Verification:** every key is unique, complete and 18 digits long, and D01, D02 and D03 contain exactly the same people. Every person has a D08 household record, and each household appears once in D08, so no person can be duplicated. The only unmatched records are the `r format(nrow(D08_households_without_D01), big.mark = ",")` D08 households with no working-age member. All of them had members in D01 before record selection, so every one of their members is under 15 or over 64. They are outside the analysis population by design, not because of a key problem, and they remain available in `D08_clean.csv` for household-level description.

## 9.2 Joins

**Problem:** the brief requires row counts to be verified before and after each join, so the three joins are run one at a time.

**Treatment:** join D02, then D03, then D08 onto D01, recording the row count before each join.

```{r}
#| label: Join D01 and D02

integrated <- D01 |>
  left_join(D02, by = c("household_id", "person_number"), relationship = "one-to-one", unmatched = "error")

#education_level_code has no missing values in D02, so a missing value here would mean a person without a D02 match
tibble(
  join = "D01 + D02",

  rows_before = nrow(D01),

  rows_in_D02 = nrow(D02),

  rows_after = nrow(integrated),

  people_without_D02_match = sum(is.na(integrated$education_level_code)),

  columns_after = ncol(integrated),

  columns_expected = ncol(D01) + ncol(D02) - 2 #The two key columns appear once
)
```

```{r}
#| label: Join D03

integrated_rows_before_D03 <- nrow(integrated)

integrated <- integrated |>
  left_join(D03, by = c("household_id", "person_number"), relationship = "one-to-one", unmatched = "error")

#employment_status2code has no missing values in D03, so a missing value here would mean a person without a D03 match
tibble(
  join = "+ D03",

  rows_before = integrated_rows_before_D03,

  rows_in_D03 = nrow(D03),

  rows_after = nrow(integrated),

  people_without_D03_match = sum(is.na(integrated$employment_status2code)),

  columns_after = ncol(integrated),

  columns_expected = ncol(D01) + ncol(D02) - 2 + ncol(D03) - 2
)
```

```{r}
#| label: Join D08

integrated_rows_before_D08 <- nrow(integrated)

integrated <- integrated |>
  left_join(D08, by = "household_id", relationship = "many-to-one")

#computer_asset_code has no missing values in D08, so a missing value here would mean a person without a D08 match
tibble(
  join = "+ D08",

  rows_before = integrated_rows_before_D08,

  rows_in_D08 = nrow(D08),

  rows_after = nrow(integrated),

  people_without_D08_match = sum(is.na(integrated$computer_asset_code)),

  households_in_integrated_data = n_distinct(integrated$household_id),

  D08_households_not_joined = nrow(D08) - n_distinct(integrated$household_id),

  columns_after = ncol(integrated),

  columns_expected = ncol(D01) + ncol(D02) - 2 + ncol(D03) - 2 + ncol(D08) - 1
)
```

**Effect:** the row count stays at `r format(nrow(integrated), big.mark = ",")` through all three joins, so no person is duplicated or lost. Every person matches a D02, D03 and D08 record, and no columns are duplicated. The only records not carried into the person-level dataset are the `r format(nrow(D08) - n_distinct(integrated$household_id), big.mark = ",")` households with no working-age member.

## 9.3 Validation of the integrated dataset

**Problem:** matching row counts do not prove that values were attached to the right people. A join on corrupted keys (for example, IDs rounded to 15 digits as described in 1.1) could keep the row count while attaching one household's values to another.

**Treatment:** check the integrated dataset against its sources. Every person key must appear once, and no suffixed duplicate columns (`.x`/`.y`) may exist. Each household's members must share a single D08 record. The category counts must equal those in the source datasets: per person for D01, D02 and D03 columns, and per household for D08 columns.

**Verification:**

```{r}
#| label: Integrated Dataset Checks

tibble(
  rows = nrow(integrated),

  distinct_person_keys = nrow(distinct(integrated, household_id, person_number)),

  columns = ncol(integrated),

  suffixed_duplicate_columns = sum(str_detect(names(integrated), "\\.x$|\\.y$")),

  households = n_distinct(integrated$household_id),

  households_with_one_D08_record = nrow(distinct(integrated, household_id, fixed_internet_code, mobile_internet_code,
                                                 computer_asset_code, household_smartphones))
)
```

```{r}
#| label: Values Carried Through Unchanged

D08_matched_households <- semi_join(D08, D01, by = "household_id")

tibble(
  column = c("sex (D01, per person)", "age (D01, per person)", "education_level (D02, per person)",
             "education_attendance (D02, per person)", "employment_status (D03, per person)", "salary_status (D03, per person)",
             "fixed_internet (D08, per household)", "mobile_internet (D08, per household)",
             "computer_asset (D08, per household)", "household_smartphones (D08, per household)"),

  identical_to_source = c(identical(count(integrated, sex), count(D01, sex)),
                          identical(count(integrated, age), count(D01, age)),
                          identical(count(integrated, education_level), count(D02, education_level)),
                          identical(count(integrated, education_attendance), count(D02, education_attendance)),
                          identical(count(integrated, employment_status), count(D03, employment_status)),
                          identical(count(integrated, salary_status), count(D03, salary_status)),
                          identical(count(distinct(integrated, household_id, fixed_internet), fixed_internet), count(D08_matched_households, fixed_internet)),
                          identical(count(distinct(integrated, household_id, mobile_internet), mobile_internet), count(D08_matched_households, mobile_internet)),
                          identical(count(distinct(integrated, household_id, computer_asset), computer_asset), count(D08_matched_households, computer_asset)),
                          identical(count(distinct(integrated, household_id, household_smartphones), household_smartphones),
                                    count(D08_matched_households, household_smartphones)))
)
```

**Effect:** the integrated dataset has `r format(nrow(integrated), big.mark = ",")` rows, one per working-age person, and `r ncol(integrated)` columns from D01, D02, D03 and D08. It covers `r format(n_distinct(integrated$household_id), big.mark = ",")` households, and every category count matches its source, so no value was duplicated or lost. Every join matched on complete, 18-digit keys, so each value belongs to the person or household it came from. Household-level values are repeated for each working-age member of the household, which is correct at the person grain. `person_weight` is the matching survey weight at this grain; `household_weight` was dropped in section 3.
