---
title: "Data Preprocessing For Datasets"
author: "Group 5"
date: today
format:
  html:
    toc: true #Gives a table of contents
    code-fold: show
    df-print: kable
  pdf:
    toc: true
    df-print: kable
execute:
  warning: false #Makes warnings not appear to keep text clean in the file will appear in console when run
  message: false
---

## Loading the packages we need:

```{r}
#| label: Package Setup

library(tidyverse)
library(janitor)
```

## File path for the dataset folder:

```{r}
#| label: File-Path Declarations

directory <- "../Datasets/"
```

# 1.1 Import with explicit column types

**Problem:** `HouseholdID` is an 18-digit number. R stores numbers as doubles, which are only exact up to about 15–16 digits, so letting `read_csv()` guess the type rounds the IDs (Milestone 1, section 8.3). `PersonNumber` is stored as "01", "02", … and must have the same type in every file for joins to match. `EducationAttendanceCode` and `MinutesToWork` contain text placeholders mixed into numeric codes.

**Treatment:** read every file with explicit types: IDs as text (`col_character()`) in all four files, and the two contaminated columns as text so no value is silently lost. `clean_names()` converts all column names to snake_case. The datasets are named D01, D02, D03 and D08, as in the ERD and the EDA, and are cleaned step by step under the same names. A summary of each uncleaned dataset is recorded in 1.2, before anything changes, for the before/after comparisons.

```{r}
#| label: Import Datasets

D01 <- read_csv(file.path(directory, "D01_Population_Profile.csv"), show_col_types = FALSE,
                col_types = cols(HouseholdID = col_character(), PersonNumber = col_character())) |> clean_names()

D02 <- read_csv(file.path(directory, "D02_Learning_Profile.csv"), show_col_types = FALSE,
                col_types = cols(HouseholdID = col_character(), PersonNumber = col_character(), EducationAttendanceCode = col_character())) |> clean_names()

D03 <- read_csv(file.path(directory, "D03_Economic_Participation.csv"), show_col_types = FALSE,
                col_types = cols(HouseholdID = col_character(), PersonNumber = col_character(),MinutesToWork = col_character())) |> clean_names()

D08 <- read_csv(file.path(directory, "D08_Connectivity_Access.csv"), show_col_types = FALSE,
                col_types = cols(HouseholdID = col_character())) |> clean_names()
```


**Verification:** no parsing problems, IDs stored as text with all 18 digits intact.


```{r}
#| label: Verify Imports

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  parsing_problems = c(nrow(problems(D01)), nrow(problems(D02)), nrow(problems(D03)), nrow(problems(D08))),
  
  household_id_type = c(class(D01$household_id), class(D02$household_id), class(D03$household_id), class(D08$household_id)),
  
  shortest_id = c(min(nchar(D01$household_id), na.rm = TRUE), min(nchar(D02$household_id)),
                  min(nchar(D03$household_id)), min(nchar(D08$household_id))),
  
  longest_id = c(max(nchar(D01$household_id), na.rm = TRUE), max(nchar(D02$household_id)),
                 max(nchar(D03$household_id)), max(nchar(D08$household_id))),
  
  distinct_households = c(n_distinct(D01$household_id, na.rm = TRUE), n_distinct(D02$household_id),
                          n_distinct(D03$household_id), n_distinct(D08$household_id)),
  
  distinct_if_stored_as_number = c(n_distinct(as.numeric(D01$household_id), na.rm = TRUE), n_distinct(as.numeric(D02$household_id)),
                                   n_distinct(as.numeric(D03$household_id)), n_distinct(as.numeric(D08$household_id)))
)
```

**Effect:** the last column shows what would have happened without this treatment. Stored as numbers, D08's `r format(n_distinct(D08$household_id), big.mark = ",")` distinct households would collapse to `r format(n_distinct(as.numeric(D08$household_id)), big.mark = ",")`, merging different households together. This is exactly the count behind the EDA's 98.77% household match, so that figure was caused by the rounding problem rather than by missing households. As text, every ID keeps all 18 digits.


# 1.2 Uncleaned "before" summary

**Problem:** the rubric requires before-and-after evidence for every major preprocessing decision.

**Treatment:** record the size, missing values, key uniqueness and duplicates of each uncleaned dataset in `uncleaned_summary`, so Part 1 and Part 2 can compare the cleaned data against the uncleaned data.

```{r}
#| label: Uncleaned Datasets Summary

uncleaned_summary <- tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  cols = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08)),
  
  missing_values = c(sum(is.na(D01)), sum(is.na(D02)), sum(is.na(D03)), sum(is.na(D08))),
  
  missing_household_id = c(sum(is.na(D01$household_id)), sum(is.na(D02$household_id)),
                           sum(is.na(D03$household_id)), sum(is.na(D08$household_id))),
  
  distinct_keys = c(nrow(distinct(D01, household_id, person_number)), nrow(distinct(D02, household_id, person_number)),
                    nrow(distinct(D03, household_id, person_number)), n_distinct(D08$household_id)),
  
  exact_duplicate_rows = c(nrow(D01) - nrow(distinct(D01)), nrow(D02) - nrow(distinct(D02)),
                           nrow(D03) - nrow(distinct(D03)), nrow(D08) - nrow(distinct(D08)))
)

uncleaned_summary
```

Columns with missing values in each uncleaned dataset:

```{r}
#| label: Missing Values in Datasets


bind_rows(
  D01 = enframe(colSums(is.na(D01)), name = "column", value = "missing"),
  D02 = enframe(colSums(is.na(D02)), name = "column", value = "missing"),
  D03 = enframe(colSums(is.na(D03)), name = "column", value = "missing"),
  D08 = enframe(colSums(is.na(D08)), name = "column", value = "missing"),
  .id = "dataset"
) |>
  filter(missing > 0) |> #Only show columns that have missing values
  left_join(select(uncleaned_summary, dataset, rows), by = "dataset") |>
  mutate(percent_of_rows = round(missing / rows * 100, 2))
```

# 1.3 Grain and key checks

**Problem:** before any join, the unit of observation (grain) and the key of every dataset must be confirmed, and it must be shown that the datasets can legitimately be linked.

**Treatment:** compare row counts with distinct keys, check how many households each dataset shares with D01, and confirm that D11 cannot be linked.

```{r}
#| label: Grain Check/Verification

D01_repeated_keys <- D01 |> filter(!is.na(household_id)) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D02_repeated_keys <- D02 |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D03_repeated_keys <- D03 |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D08_repeated_keys <- D08 |> count(household_id) |> filter(n > 1) |> nrow()

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  
  grain = c("person", "person", "person", "household"),
  
  key = c("household_id + person_number", "household_id + person_number", "household_id + person_number", "household_id"),
  
  rows = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  distinct_keys = c(nrow(distinct(filter(D01, !is.na(household_id)), household_id, person_number)),
                    nrow(distinct(D02, household_id, person_number)),
                    nrow(distinct(D03, household_id, person_number)),
                    n_distinct(D08$household_id)),
  
  keys_appearing_more_than_once = c(D01_repeated_keys, D02_repeated_keys, D03_repeated_keys, D08_repeated_keys),
  
  rows_missing_household_id = c(sum(is.na(D01$household_id)), sum(is.na(D02$household_id)),
                                sum(is.na(D03$household_id)), sum(is.na(D08$household_id)))
)
```


```{r}
#| label: Shared Households Check

D11_ids <- read_csv(file.path(directory, "D11_Labour_Market.csv"), show_col_types = FALSE,
                    col_select = HouseholdID, col_types = cols(.default = col_character())) |> clean_names()

D01_households <- unique(na.omit(D01$household_id)) #Distinct household IDs in D01

tibble(
  compared_with = c("D02", "D03", "D08", "D11"),
  
  households_in_dataset = c(n_distinct(D02$household_id), n_distinct(D03$household_id),
                            n_distinct(D08$household_id), n_distinct(D11_ids$household_id, na.rm = TRUE)),
  
  shared_with_D01 = c(length(intersect(D01_households, D02$household_id)),
                      length(intersect(D01_households, D03$household_id)),
                      length(intersect(D01_households, D08$household_id)),
                      length(intersect(D01_households, D11_ids$household_id))),
  
  percent_of_D01 = round(shared_with_D01 / length(D01_households) * 100, 2)
)
```

```{r}
#| label: People per Household Check

D01 |>
  filter(!is.na(household_id)) |>
  count(household_id, name = "persons") |>
  summarise(households = n(), mean_persons = round(mean(persons), 2), median_persons = median(persons), max_persons = max(persons))
```

**Effect / Verification:** D01, D02 and D03 are one row per person and D08 is exactly one row per household (`r format(nrow(D08), big.mark = ",")` rows, `r format(n_distinct(D08$household_id), big.mark = ",")` distinct IDs). All `r format(length(D01_households), big.mark = ",")` D01 households appear in D02, D03 and D08 (100%). This corrects the 98.77% reported in the EDA: its count of `r format(n_distinct(as.numeric(D08$household_id)), big.mark = ",")` is the number of distinct IDs left after rounding (step 1.1). D11 shares **0** households with D01, which confirms its exclusion. Household IDs repeat in the person files because a household contains several people. These are legitimate repeats in a one-to-many relationship, not duplicates. The keys that appear more than once in D02 and D03, and the missing IDs in D01, are treated in steps 1.4 and 1.5.

# 1.4 Recover the missing household IDs in D01

**Problem:** `r sum(is.na(D01$household_id))` rows in D01 have a missing `household_id`. These people cannot be linked to their education, employment or household connectivity. Exactly the same number of people appear in D03 (and D02) but not in D01, which suggests these are the same people with their ID blanked out.

**Treatment:** match each D01 row with a missing ID to one of the people missing from D01, using `person_number`, `sex_code`, `province_code` and `person_weight`. The survey weight is identical for the same person across D01–D03 and has thousands of distinct values, so it is highly specific. The match is then checked with two attributes that were **not** used to match: the PSU (sampling cluster) must equal the first 11 digits of the recovered household ID, and the age must agree with D03.

```{r}
#| label: Identify Missing People

D01_key_duplicates_before <- get_dupes(D01, household_id, person_number) 

missing_from_D01 <- D03 |>
  distinct() |> #
  anti_join(filter(D01, !is.na(household_id)), by = c("household_id", "person_number")) |> 
  select(household_id, person_number, sex_code, province_code, person_weight, age_d03 = age)

rows_missing_id <- D01 |>
  mutate(row_id = row_number()) |> 
  filter(is.na(household_id)) |>
  select(-household_id)

recovered <- rows_missing_id |>
  inner_join(missing_from_D01, by = c("person_number", "sex_code", "province_code", "person_weight"))
```

**Verification:** every D01 row with a missing ID matches exactly one person, every candidate is used exactly once, and both independent checks agree for every match.

```{r}
#| label: Verifification of Missing People 

tibble(
  
  D01_rows_missing_id = nrow(rows_missing_id),
  
  people_missing_from_D01 = nrow(missing_from_D01),
  
  matches_found = nrow(recovered),
  
  each_d01_row_matched_once = n_distinct(recovered$row_id) == nrow(rows_missing_id),
  
  each_person_used_once = nrow(distinct(recovered, household_id, person_number)) == nrow(recovered),
  
  psu_agrees_with_id = sum(sprintf("%.0f", recovered$psu) == str_sub(recovered$household_id, 1, 11)), 
  
  age_agrees_with_D03 = sum(recovered$age == recovered$age_d03, na.rm = TRUE)
)
```

```{r}
#| label: Recover Missing People

D01 <- D01 |>
  mutate(row_id = row_number()) |>
  
  left_join(select(recovered, row_id, recovered_id = household_id), by = "row_id") |>
  
  mutate(
    household_id_flag = if_else(is.na(household_id) & !is.na(recovered_id), "recovered by matching to D03", NA_character_),
    household_id = coalesce(household_id, recovered_id)
  ) |>
 
  select(-row_id, -recovered_id)

tibble(
  missing_household_id_before = uncleaned_summary$missing_household_id[1], 

  missing_household_id_after = sum(is.na(D01$household_id)),
  
  rows_before = uncleaned_summary$rows[1],
  
  rows_after = nrow(D01),
  
  D01_keys_identical_to_D02 = setequal(paste(D01$household_id, D01$person_number),
                                       paste(D02$household_id, D02$person_number))
)
```

**Effect:** all `r nrow(recovered)` IDs are recovered with no rows added or removed, and D01 now contains exactly the same set of people as D02. Recovered rows are marked in `household_id_flag`.


# 1.5 Duplicates

**Problem:** the EDA found `r nrow(D01_key_duplicates_before)` (D01), `r nrow(get_dupes(D02, household_id, person_number))` (D02) and `r nrow(get_dupes(D03, household_id, person_number))` (D03) rows sharing a person key. These must be separated into genuine duplicate records (the same record entered twice) and legitimate repeated identifiers.

**Investigation:** every one of D01's key "duplicates" had a missing household ID. Rows with a missing ID and the same person number were grouped together, even though they are different people.

```{r}
#| label: False Duplicates in D01

D01_key_duplicates_before |> count(household_id_missing = is.na(household_id), name = "rows")
```

After the recovery in 1.4, check each dataset for exact copies (every column identical) and for keys that still repeat with different values (conflicts):

```{r}
#| label: Duplicate Check

#Rows that share their key with at least one other row
D01_key_duplicate_rows <- nrow(get_dupes(D01, household_id, person_number))
D02_key_duplicate_rows <- nrow(get_dupes(D02, household_id, person_number))
D03_key_duplicate_rows <- nrow(get_dupes(D03, household_id, person_number))
D08_key_duplicate_rows <- nrow(get_dupes(D08, household_id))

#Exact copies: rows where every column is identical to another row
D01_exact_duplicates <- nrow(D01) - nrow(distinct(D01))
D02_exact_duplicates <- nrow(D02) - nrow(distinct(D02))
D03_exact_duplicates <- nrow(D03) - nrow(distinct(D03))
D08_exact_duplicates <- nrow(D08) - nrow(distinct(D08))

#Keys that still repeat once exact copies are removed, meaning the same person with conflicting values
D01_conflicting_keys <- distinct(D01) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D02_conflicting_keys <- distinct(D02) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D03_conflicting_keys <- distinct(D03) |> count(household_id, person_number) |> filter(n > 1) |> nrow()
D08_conflicting_keys <- distinct(D08) |> count(household_id) |> filter(n > 1) |> nrow()

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows_before = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  rows_in_key_duplicate_groups = c(D01_key_duplicate_rows, D02_key_duplicate_rows, D03_key_duplicate_rows, D08_key_duplicate_rows),
  
  exact_duplicate_rows = c(D01_exact_duplicates, D02_exact_duplicates, D03_exact_duplicates, D08_exact_duplicates),
  
  conflicting_keys_after_removing_exact_copies = c(D01_conflicting_keys, D02_conflicting_keys, D03_conflicting_keys, D08_conflicting_keys)
)
```

**Treatment:** every key duplicate in D02 and D03 is an exact copy of another row, with no conflicting values, so they are genuine duplicate records. Exact copies are removed with `distinct()`. D01 and D08 contain none.

```{r}
#| label: Remove Duplicates

D01 <- distinct(D01)
D02 <- distinct(D02)
D03 <- distinct(D03)
D08 <- distinct(D08)

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  
  rows_before = uncleaned_summary$rows,
  
  rows_after = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  
  rows_removed = rows_before - rows_after,
  
  keys_unique_after = c(nrow(distinct(D01, household_id, person_number)) == nrow(D01),
                        nrow(distinct(D02, household_id, person_number)) == nrow(D02),
                        nrow(distinct(D03, household_id, person_number)) == nrow(D03),
                        n_distinct(D08$household_id) == nrow(D08))
)
```

**Effect / Verification:** D02 loses `r D02_exact_duplicates` rows and D03 loses `r D03_exact_duplicates`. All three person datasets now contain exactly `r format(nrow(D01), big.mark = ",")` people, and every key is unique in every dataset. Repeated `household_id` values remain in the person files, because they are legitimate: one household, several people.

# 1.6 Text placeholders and data types

**Problem:** a code column should contain only numbers. Listing the text columns in each dataset shows which columns contain text where codes are expected (the `_flag` columns created in this pipeline are text by design and are left out):

```{r}
#| label: Text Columns Explanation

bind_rows(
  D01 = enframe(sapply(D01, class), name = "column", value = "type"),
  D02 = enframe(sapply(D02, class), name = "column", value = "type"),
  D03 = enframe(sapply(D03, class), name = "column", value = "type"),
  D08 = enframe(sapply(D08, class), name = "column", value = "type"),
  .id = "dataset"
) |>
  filter(type == "character", !str_ends(column, "_flag")) |> 
  group_by(column) |>
  summarise(datasets = paste(dataset, collapse = ", ")) |> 
  mutate(why_text = case_when(
    column %in% c("household_id", "person_number") ~ "Key, read as text on purpose in 1.1",
    column == "province" ~ "Province name (rebuilt from province_code, dropped in 1.8)",
    TRUE ~ "Code column containing text placeholders"
  ))
```

Apart from IDs and the province name, only two columns are text: `education_attendance_code` (D02, kept) and `minutes_to_work` (D03, dropped as leakage in 1.8, so it is not treated). `education_attendance_code` mixes numeric codes with three text placeholders, the same problem Milestone 1 found in D11's `hours_worked_raw`:

```{r}
#| label: Initial Student Attendance

D02 |> count(education_attendance_code)
```

None of the placeholders appears in the data dictionary. "not recorded" and "Unknown" are non-response. "N/A" cannot be assumed to mean code 8 ("not applicable"): comparing ages shows that code 8 is only used for children aged 0–4, while the placeholders occur at ages up to 72:

```{r}
#| label: Attendance Codes Placeholder Verification

attendance_text <- c("N/A", "not recorded", "Unknown")
D02 |>
  filter(education_attendance_code %in% c("8", attendance_text)) |>
  group_by(value = education_attendance_code) |>
  summarise(rows = n(), min_age = min(age), median_age = median(age), max_age = max(age))
```

**Treatment:** set the text placeholders to missing, record why in `education_attendance_flag`, and convert the column to integer.

```{r}
#| label: Fixing Broken Attendance Codes

D02 <- D02 |>
  mutate(
    education_attendance_flag = if_else(education_attendance_code %in% attendance_text,
                                        paste0("text placeholder '", education_attendance_code, "' set to missing"), NA_character_),
    
    education_attendance_code = as.integer(if_else(education_attendance_code %in% attendance_text, NA_character_, education_attendance_code))
  )
```

**Verification:**

```{r}
#| label: Fixed Attendance Codes

D02 |> count(education_attendance_code, education_attendance_flag)
class(D02$education_attendance_code)
```

**Effect:** `education_attendance_code` is now an integer column holding only dictionary codes (1, 2, 3, 8) or missing. The `r sum(!is.na(D02$education_attendance_flag))` former placeholders are traceable through the flag. Code 3 ("do not know") is handled in Part 2 with the other special codes.

# 1.7 Repairs from other datasets

## Age (D01)

**Problem:** the data dictionary gives a valid age range of 0–110. D01 has `r sum(is.na(D01$age))` missing ages and `r sum(D01$age < 0 | D01$age > 110, na.rm = TRUE)` ages outside that range. D02 and D03 record the age of the same people.

**Investigation:** check that D02 and D03 agree with each other, and that D01 agrees with D03 wherever D01's age is valid:

```{r}
#| label: Ages Inspection

age_check <- D01 |>
  left_join(select(D03, household_id, person_number, age_d03 = age), by = c("household_id", "person_number")) |>
  left_join(select(D02, household_id, person_number, age_d02 = age), by = c("household_id", "person_number"))

age_check |>
  summarise(
    rows = n(),
    
    d02_vs_d03_mismatches = sum(age_d02 != age_d03),
    
    d03_ages_outside_0_110 = sum(age_d03 < 0 | age_d03 > 110),
    
    d01_valid_ages_agreeing_with_d03 = sum(age == age_d03 & between(age, 0, 110), na.rm = TRUE),
    
    d01_valid_ages_disagreeing_with_d03 = sum(age != age_d03 & between(age, 0, 110), na.rm = TRUE),
    
    d01_missing = sum(is.na(age)),
    
    d01_invalid = sum(age < 0 | age > 110, na.rm = TRUE)
  )

age_check |>
  filter(age < 0 | age > 110) |>
  select(household_id, person_number, age_d01 = age, age_d02, age_d03)
```

D02 and D03 agree on every person and contain no invalid ages, and D01 agrees with them on every valid age. The only disagreements are D01's invalid values, so D03's age is a reliable replacement. Deleting these people would lose valid education and employment records for no reason.

**Treatment:** replace D01's missing and invalid ages with D03's age and record which rows changed in `age_flag`.

```{r}
#| label: Fixed Ages

D01 <- age_check |>
  mutate(
    
    age_flag = case_when(is.na(age) ~ "missing - filled from D03",
      age < 0 | age > 110 ~ paste0("invalid (", age, ") - replaced from D03")
    ),
    
    age = if_else(is.na(age) | age < 0 | age > 110, age_d03, age)
  ) |>
  
  select(-age_d02, -age_d03)
```

**Verification:**

```{r}
#| label: Verification of Ages after Fix

tibble(
  rows = nrow(D01),
  
  missing_age = sum(is.na(D01$age)),
  
  min_age = min(D01$age),
  
  max_age = max(D01$age),
  
  ages_differing_from_d03 = sum(D01$age != age_check$age_d03)
)

D01 |> count(age_flag = if_else(is.na(age_flag), "unchanged", str_remove(age_flag, " \\(.*\\)")))
```

**Effect:** `r sum(!is.na(D01$age_flag))` ages are repaired and none are deleted. D01's age now lies between 0 and 110 with no missing values, and agrees with D02 and D03 for every person.

## Education level (D02)

**Problem:** `education_level_code` in D02 has `r sum(is.na(D02$education_level_code))` missing values. D03 records the same variable for the same people.

```{r}
#| label: Education Dataset Inspection

D02_education_missing_before <- sum(is.na(D02$education_level_code)) #Kept for the before/after check below

education_check <- D02 |>
  left_join(select(D03, household_id, person_number, education_d03 = education_level_code), by = c("household_id", "person_number"))

education_check |>
  summarise(
    rows = n(),
    
    missing_in_d02 = sum(is.na(education_level_code)),
    
    missing_in_d03 = sum(is.na(education_d03)),
    
    agree_where_both_present = sum(education_level_code == education_d03, na.rm = TRUE),
    
    disagree_where_both_present = sum(education_level_code != education_d03, na.rm = TRUE),
    
    d02_missing_but_d03_present = sum(is.na(education_level_code) & !is.na(education_d03))
  )
```

**Treatment:** D02 and D03 never disagree, and D03 has no missing values, so D02's gaps are filled from D03 and flagged in `education_level_flag`.

```{r}
#| label: Fixed Education Dataset 

D02 <- education_check |>
  mutate(
    education_level_flag = if_else(is.na(education_level_code), "missing - filled from D03", NA_character_),
    education_level_code = coalesce(education_level_code, education_d03)
  ) |>
  select(-education_d03)

tibble(
       missing_before = D02_education_missing_before,
       
       missing_after = sum(is.na(D02$education_level_code)),
       
       rows_flagged = sum(!is.na(D02$education_level_flag))
       )
```

**Effect:** D02's education level is complete. Code 99, which appears in both D02 and D03 but not in the dictionary, is not a gap and cannot be filled from D03. It is treated with the other special codes in Part 2.

# 1.8 Column selection

**Problem:** the four datasets contain 83 columns, most of which do not measure a link in the research pathway, duplicate another dataset, or leak the outcome. The full reasoning for each column is in the Task 1 write-up (Data Selection and Analytical Grain). The evidence below supports the main drop decisions.

## Evidence: columns that carry no information or duplicate another column

```{r}
#| label: Evidence of Province Mismatches

tibble(
  check = c("distinct survey_year values in D01", "distinct survey_year values in D02",
            "distinct survey_year values in D03", "distinct survey_year values in D08",
            "distinct province spellings in D01",
            "distinct province spellings after trimming and case-fixing",
            "distinct province_code values"),
  
  result = c(n_distinct(D01$survey_year), n_distinct(D02$survey_year),
             n_distinct(D03$survey_year), n_distinct(D08$survey_year),
             n_distinct(D01$province),
             n_distinct(str_to_title(str_squish(D01$province))),
             n_distinct(D01$province_code))
)
D01 |> distinct(province_code, province_clean = str_to_title(str_squish(province))) |> arrange(province_code)
```

```{r}
#| label: Evidence of Duplicate Columns

#One row per person with the D01, D02 and D03 versions of each shared column side by side
persons <- D01 |>
  inner_join(D02, by = c("household_id", "person_number"), suffix = c("", "_d02")) |>
  inner_join(D03, by = c("household_id", "person_number"), suffix = c("", "_d03"))

#How many people have a different value in D02 or D03 than in D01
tibble(
  column = c("sex_code", "province_code", "geo_type_code", "person_weight", "age"),
  
  people_compared = nrow(persons),
  
  mismatches_D01_vs_D02 = c(sum(persons$sex_code != persons$sex_code_d02),
                            sum(persons$province_code != persons$province_code_d02),
                            sum(persons$geo_type_code != persons$geo_type_code_d02),
                            sum(persons$person_weight != persons$person_weight_d02),
                            sum(persons$age != persons$age_d02)),
  
  mismatches_D01_vs_D03 = c(sum(persons$sex_code != persons$sex_code_d03),
                            sum(persons$province_code != persons$province_code_d03),
                            sum(persons$geo_type_code != persons$geo_type_code_d03),
                            sum(persons$person_weight != persons$person_weight_d03),
                            sum(persons$age != persons$age_d03))
)

#D08 against D01 at household level
D01 |>
  distinct(household_id, province_code, geo_type_code) |>
  inner_join(D08, by = "household_id", suffix = c("", "_d08")) |>
  summarise(households_compared = n(),
  province_mismatch = sum(province_code != province_code_d08),
  geo_type_mismatch = sum(geo_type_code != geo_type_code_d08))


tibble(
      age_group_matches_age = sum(as.integer(cut(D01$age, c(-Inf, seq(4, 74, 5), Inf))) == D01$age_group_code), 
      
      rows = nrow(D01)
)

```

## Evidence: leakage (columns that reveal or define the employment outcome)

```{r}
#| label: Evidence of Leakage

D03 |> count(employment_status1code, employment_status2code) 

D03 |> count(wage_work_code, employment_status2code) 

D03 |> count(looking_for_work_code, employment_status2code) 

D03 |> count(sector_code, employment_status2code) 
```

## Evidence: unusable or redundant columns

```{r}
#| label: Evidence of Unusable/Redundant Columns

D03 |> count(salary_amount_raw) #Only a handful of distinct values, so not Rand amounts

D02 |>
  filter(between(age, 15, 64), education_attendance_code == 1, grade_code %in% 1:12, education_level_code %in% 0:12) |>
  count(grade_minus_education_level = grade_code - education_level_code) 

D02 |>
  filter(between(age, 15, 64)) |>
  summarise(
    across(c(public_private_institution_code, transport_mode_to_education_code, travel_time_to_education,
                     bursary_code, total_education_fees, no_fees_school_code, school_food_programme_code,
                     absence_code, absence_days),
                   ~ round(mean(.x %in% c(8, 88)) * 100, 1))) |>
  pivot_longer(everything(), names_to = "column", values_to = "percent_not_applicable_age_15_64")
```

## Evidence: education columns kept or dropped for Link A (digital access → education)

```{r}
#| label: Linking Digital Access to Education

youth <- D02 |> filter(between(age, 15, 24))

youth |> filter(education_attendance_code == 1) |> count(institution_type_code) 
youth |> filter(education_attendance_code == 2) |> count(reason_not_attending_code, sort = TRUE) #Kept for descriptive use only

youth |>
  left_join(select(D08, household_id, study_internet_code), by = "household_id") |> 
  filter(education_attendance_code %in% 1:2, study_internet_code %in% 1:2) |>
  count(study_internet_code, education_attendance_code) 
```

**Treatment:** keep only the 20 selected columns plus the join keys and the traceability flags created in Part 1.

```{r}
#| label: Slecting which columns to keep

#Column names before selection, kept to show which columns were dropped
D01_columns_before <- names(D01)
D02_columns_before <- names(D02)
D03_columns_before <- names(D03)
D08_columns_before <- names(D08)

D01 <- D01 |>
  select(household_id, person_number, psu, province_code, sex_code, age, population_group_code,
         geo_type_code, metro_indicator, person_weight, household_id_flag, age_flag)

D02 <- D02 |>
  select(household_id, person_number, education_level_code, education_attendance_code, institution_type_code,
         reason_not_attending_code, education_level_flag, education_attendance_flag)

D03 <- D03 |>
  select(household_id, person_number, employment_status2code, lab_salary_raw)

D08 <- D08 |>
  select(household_id, fixed_internet_code, mobile_internet_code, computer_asset_code, household_smartphones)
```

**Verification:**

```{r}
#| label: Comparing Before and After Column Selection

tibble(
  
  dataset = c("D01", "D02", "D03", "D08"),
  
  columns_in_uncleaned_data = uncleaned_summary$cols,
  
  columns_before_selection = c(length(D01_columns_before), length(D02_columns_before), length(D03_columns_before), length(D08_columns_before)),
  
  columns_after_selection = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08))
)
```

Columns dropped from each dataset:

```{r}
#| label: Columns Dropped

#Dropped columns for each dataset, stacked into one table
bind_rows(
  tibble(dataset = "D01", column = setdiff(D01_columns_before, names(D01))),
  tibble(dataset = "D02", column = setdiff(D02_columns_before, names(D02))),
  tibble(dataset = "D03", column = setdiff(D03_columns_before, names(D03))),
  tibble(dataset = "D08", column = setdiff(D08_columns_before, names(D08)))
) |>
  
  #Reason for dropping each column, matching the Task 1 write-up
  mutate(reason = case_when(
    column == "survey_year" ~ "No information (always 2024)",
    column == "province" ~ "Messy text, rebuilt from province_code",
    column %in% c("province_code", "age", "sex_code", "geo_type_code", "person_weight") ~ "Duplicate of D01",
    column == "education_level_code" ~ "Duplicate of D02 (used to fill D02's gaps in 1.7)",
    column %in% c("age_group_code", "employment_status1code", "grade_code") ~ "Duplicate of another column",
    column %in% c("wage_work_code", "business_work_code", "volunteer_work_code", "retired_code",
                  "looking_for_work_code", "could_start_work_code", "transport_to_work_code",
                  "minutes_to_work", "sector_code", "work_internet_code", "study_internet_code") ~ "Leakage: reveals or defines an outcome",
    column %in% c("language_code", "relationship_to_head_code", "marital_status_code") ~ "Not needed for any hypothesis",
    column == "metro_code" ~ "Too detailed (disclosure risk)",
    column %in% c("public_private_institution_code", "transport_mode_to_education_code", "travel_time_to_education",
                  "bursary_code", "total_education_fees", "no_fees_school_code", "school_food_programme_code",
                  "absence_code", "absence_days") ~ "Only recorded for people already attending",
    column == "salary_amount_raw" ~ "Unusable (not Rand amounts)",
    column %in% c("salary_period_code", "salary_category_code") ~ "Salary detail not needed (lab_salary_raw kept)",
    column %in% c("telephone_code", "cellphone_code", "household_mobile_phones") ~ "Phones, not internet access",
    column %in% c("public_wi_fi_internet_code", "library_internet_code", "cafe_internet_code") ~ "Internet access outside the home",
    column == "household_weight" ~ "Household weight, not needed at person grain",
    TRUE ~ "NOT CLASSIFIED" #Should never appear; shows up if a dropped column has no reason
  )) |>
  group_by(dataset, reason) |>
  summarise(columns_dropped = n(), columns = paste(column, collapse = ", "), .groups = "drop")
```

#####################################################################################################################


# Part 2: Validity, Recoding and Final Validation

```{r}
#| label: Output Folder

cleaned_directory <- "../Datasets/Cleaned/" #Final cleaned datasets for Task 3 and Milestone 3
dir.create(cleaned_directory, showWarnings = FALSE)
```

# 2.1 Special codes and salary reversal

**Problem:** several coded columns mix valid categories, structural "not applicable" codes and non-response codes. `lab_salary_raw` mixes real Rand values with two placeholder codes.

**Treatment:** classify each special code as valid, "Not applicable", or `NA`, applied via `factor()` levels in 2.3. For salary, cross-tabulate the placeholders against `employment_status2code` before deciding.

```{r}
#| label: Special Codes Check

D02 |> count(education_level_code, sort = TRUE) |> filter(education_level_code %in% c(98, 99, 29, 28))
D02 |> count(education_attendance_code)
D02 |> count(institution_type_code == 88, name = "rows")
D02 |> count(reason_not_attending_code == 88, name = "rows")
D08 |> count(fixed_internet_code)
D08 |> count(mobile_internet_code)
D08 |> count(computer_asset_code)
```

Code 98 (No schooling) and 28 (Other) in `education_level_code` are valid categories. Codes 99 (not in dictionary) and 29 (do not know) become `NA`. Code 3 in `education_attendance_code` becomes `NA`; code 8 only occurs under age 15 and disappears after 2.2. `education_attendance_code` also has `r sum(is.na(D02$education_attendance_code))` genuinely missing values (present since the raw data, distinct from the text placeholders resolved in 1.6); these remain `NA` in the label column since there is no dictionary code to assign them. Code 88 in `institution_type_code`/`reason_not_attending_code` becomes "Not applicable", not missing. `fixed_internet_code` has `r sum(D08$fixed_internet_code == 9, na.rm = TRUE)` rows with code 9, which is not in the dictionary and becomes `NA`. `mobile_internet_code` has `r sum(is.na(D08$mobile_internet_code))` genuinely missing values, present since the raw data; these remain `NA` since there is no dictionary code to assign them. `computer_asset_code` 9 ("Unspecified") is a valid dictionary code and is kept as its own level.

```{r}
#| label: Salary Placeholder Cross-Tab

D03 |>
  mutate(salary_type = case_when(
    lab_salary_raw == 888888888 ~ "888888888 (dont know/refused)",
    lab_salary_raw == 999999999 ~ "999999999 (not applicable)",
    TRUE ~ "Real salary value"
  )) |>
  count(employment_status2code, salary_type) |>
  pivot_wider(names_from = employment_status2code, values_from = n, values_fill = 0)
```

888888888 acts as "not applicable" for everyone without a job, and 999999999 as non-response among the employed. `salary` and `salary_status` are derived from the value and `employment_status2code` together, not from the placeholder alone.

```{r}
#| label: Salary Reversal

D03_salary_missing_before <- sum(D03$lab_salary_raw %in% c(888888888, 999999999))

D03 <- D03 |>
  mutate(
    salary = if_else(lab_salary_raw %in% c(888888888, 999999999), NA_real_, lab_salary_raw),
    salary_status = case_when(
      !is.na(salary) ~ "Reported",
      lab_salary_raw == 999999999 ~ "Non-response (employed)",
      lab_salary_raw == 888888888 & employment_status2code == 1 ~ "Non-response (employed)",
      lab_salary_raw == 888888888 ~ "Not applicable (not employed)",
      TRUE ~ NA_character_
    )
  )
```

**Effect:** `r format(sum(!is.na(D03$salary) & D03$employment_status2code == 1), big.mark = ",")` of `r format(sum(D03$employment_status2code == 1), big.mark = ",")` employed people report a real salary (`r round(mean(!is.na(D03$salary[D03$employment_status2code == 1])) * 100)`%). This corrects Milestone 1's "74% unusable" figure, which counted people who cannot have a salary in the first place.

**Verification:**

```{r}
#| label: Salary Reversal Verification

D03 |> count(salary_status, has_salary = !is.na(salary))

tibble(
  placeholders_before = D03_salary_missing_before,
  placeholders_remaining_in_salary = sum(D03$salary %in% c(888888888, 999999999), na.rm = TRUE)
)
```

# 2.2 Record selection (ages 15–64)

**Problem:** the analysis population is the working-age population, ages 15–64.

**Treatment:** filter D01 on `between(age, 15, 64)`, then keep only matching people in D02 and D03 with `semi_join()`. All D08 households are kept; Task 3 reports the households with no working-age member.

```{r}
#| label: Record Selection

D01_rows_before_selection <- nrow(D01)
D01_under_15 <- sum(D01$age < 15)
D01_65_plus <- sum(D01$age > 64)

D01 <- D01 |> filter(between(age, 15, 64))
D02 <- semi_join(D02, D01, by = c("household_id", "person_number"))
D03 <- semi_join(D03, D01, by = c("household_id", "person_number"))
```

**Verification:**

```{r}
#| label: Record Selection Verification

tibble(
  dataset = c("D01", "D02", "D03"),
  rows_before = D01_rows_before_selection,
  rows_after = c(nrow(D01), nrow(D02), nrow(D03)),
  removed_under_15 = D01_under_15,
  removed_65_plus = D01_65_plus,
  distinct_households = c(n_distinct(D01$household_id), n_distinct(D02$household_id), n_distinct(D03$household_id)),
  identical_keys_to_D01 = c(TRUE,
                            setequal(paste(D01$household_id, D01$person_number), paste(D02$household_id, D02$person_number)),
                            setequal(paste(D01$household_id, D01$person_number), paste(D03$household_id, D03$person_number)))
)
```

**Effect:** all three person files have `r format(nrow(D01), big.mark = ",")` rows and identical keys, covering `r format(n_distinct(D01$household_id), big.mark = ",")` households. `r format(D01_under_15, big.mark = ",")` people under 15 and `r format(D01_65_plus, big.mark = ",")` aged 65+ were removed.

# 2.3 Recode into labelled factors

**Problem:** `_code` columns are not analytically meaningful or safe to treat as numbers.

**Treatment:** for every kept `_code` column, add a label column of the same name without the suffix, using `factor(..., levels =, labels =)`. Codes left out of `levels =` become `NA`, applying the 2.1 classification. `_code` columns are never modified.

```{r}
#| label: Recode D01

D01_types_before <- sapply(D01, class)

D01 <- D01 |>
  mutate(
    province = factor(province_code, levels = 1:9,
                      labels = c("Western Cape", "Eastern Cape", "Northern Cape", "Free State",
                                 "KwaZulu-Natal", "North West", "Gauteng", "Mpumalanga", "Limpopo")),
    sex = factor(sex_code, levels = 1:2, labels = c("Male", "Female")),
    population_group = factor(population_group_code, levels = 1:4,
                              labels = c("Black African", "Coloured", "Indian/Asian", "White")),
    geo_type = factor(geo_type_code, levels = 1:3, labels = c("Urban", "Traditional", "Farms")),
    metro = factor(metro_indicator, levels = 1:2, labels = c("Metro", "Non-metro"))
  )
```

```{r}
#| label: Recode D02

D02_types_before <- sapply(D02, class)

D02 <- D02 |>
  mutate(
    education_level = factor(education_level_code,
                             levels = c(98, 0:27, 28),
                             labels = c(
                               "No schooling",
                               "Grade R/0", "Grade 1/Sub A/Class 1", "Grade 2/Sub B/Class 2",
                               "Grade 3/Standard 1/ABET/AET 1", "Grade 4/Standard 2",
                               "Grade 5/Standard 3/ABET/AET 2", "Grade 6/Standard 4",
                               "Grade 7/Standard 5/ABET/AET 3", "Grade 8/Standard 6/Form 1",
                               "Grade 9/Standard 7/Form 2/ABET/AET 4/NCV 1",
                               "Grade 10/Standard 8/Form 3/NCV 2",
                               "Grade 11/Standard 9/Form 4/NCV 3",
                               "Grade 12/Standard 10/Form 5/NSC/Matric/NCV 4",
                               "NTC I/N1/NQF 1", "NTC II/N2/NQF 2", "NTC III/N3/NQF 3",
                               "N4/NTC 4/Occupational Cert NQF 5", "N5/NTC 5/Occupational Cert NQF 5",
                               "N6/NTC 6/Occupational Cert NQF 5",
                               "Certificate less than Grade 12", "Diploma less than Grade 12",
                               "Higher/National/Advanced Certificate with Grade 12/NQF 5",
                               "Diploma with Grade 12/Occupational Cert NQF 6",
                               "Higher Diploma/B-Tech/Occupational Cert NQF 7",
                               "Bachelor's Degree/Occupational Cert NQF 7",
                               "Honours Degree/Postgrad Diploma/Occupational Cert NQF 8",
                               "Post Higher Diploma (M-Tech/Master's) NQF 9",
                               "Doctoral Degree (D-Tech/PhD) NQF 10",
                               "Other"
                             )),
    education_attendance = factor(education_attendance_code, levels = 1:2, labels = c("Yes", "No")),
    institution_type = factor(institution_type_code, levels = c(1:9, 88),
                              labels = c(paste0("Institution type ", 1:9), "Not applicable")),
    reason_not_attending = factor(reason_not_attending_code, levels = c(1:18, 88),
                                  labels = c(paste0("Reason ", 1:18), "Not applicable"))
  )
```

```{r}
#| label: Recode D03

D03_types_before <- sapply(D03, class)

D03 <- D03 |>
  mutate(
    employment_status = factor(employment_status2code, levels = 1:3,
                               labels = c("Employed", "Unemployed", "Not economically active"))
  )
```

```{r}
#| label: Recode D08

D08_types_before <- sapply(D08, class)

D08 <- D08 |>
  mutate(
    fixed_internet = factor(fixed_internet_code, levels = 1:2, labels = c("Yes", "No")),
    mobile_internet = factor(mobile_internet_code, levels = 1:2, labels = c("Yes", "No")),
    computer_asset = factor(computer_asset_code, levels = c(1, 2, 9), labels = c("Yes", "No", "Unspecified"))
  )
```

**Verification:** every code maps to exactly one label.

```{r}
#| label: Recode Verification

D01 |> count(province_code, province)
D01 |> count(sex_code, sex)
D01 |> count(population_group_code, population_group)
D01 |> count(geo_type_code, geo_type)
D01 |> count(metro_indicator, metro)

D02 |> count(education_level_code, education_level)
D02 |> count(education_attendance_code, education_attendance)
D02 |> count(institution_type_code, institution_type)
D02 |> count(reason_not_attending_code, reason_not_attending)

D03 |> count(employment_status2code, employment_status)

D08 |> count(fixed_internet_code, fixed_internet)
D08 |> count(mobile_internet_code, mobile_internet)
D08 |> count(computer_asset_code, computer_asset)
```

# 2.4 Outlier review

**Problem:** `age`, `household_smartphones` and `salary` are the only kept columns that are real measurements rather than categories.

```{r}
#| label: Age Outlier Check

summary(D01$age)

ggplot(D01, aes(x = age)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Age distribution after record selection", x = "Age", y = "Count")
```

**Decision:** age is bounded at 15–64 by 2.2. No further treatment needed.

```{r}
#| label: Smartphones Outlier Check

summary(D08$household_smartphones)

household_size <- D01 |> count(household_id, name = "household_size")

D08 |>
  left_join(household_size, by = "household_id") |>
  filter(household_smartphones >= 10) |>
  select(household_id, household_smartphones, household_size)
```

**Decision:** the range (0–13) matches the dictionary maximum. The `r sum(D08$household_smartphones >= 10, na.rm = TRUE)` households with 10+ smartphones are plausible against household size, so kept unchanged. `household_size` is not joined onto `D08` or saved; it is only used here for this check.

```{r}
#| label: Salary Outlier Check

D03 |> filter(!is.na(salary)) |> summarise(
  min = min(salary), p1 = quantile(salary, 0.01), median = median(salary),
  p99 = quantile(salary, 0.99), max = max(salary)
)

D03 |>
  filter(!is.na(salary)) |>
  ggplot(aes(x = salary)) +
  geom_histogram(bins = 50) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Reported salary (log scale)", x = "Salary (Rand, log scale)", y = "Count")
```

**Decision:** the top tail (up to R1.4 million) contains genuine high earners and is kept. Values under R100 cannot plausibly be a monthly salary and are flagged, not deleted, since the pay period is not documented in the source data.

```{r}
#| label: Flagging Low Salary Outliers

D03 <- D03 |>
  mutate(salary_flag = if_else(!is.na(salary) & salary < 100, "Implausibly low for a monthly salary", NA_character_))

D03 |> filter(!is.na(salary_flag)) |> count(salary_flag)
```

`person_weight` is not touched: survey weights must never be trimmed or scaled. No scaling or transformation is applied in Part 2; that belongs in Task 4, after the train/test split, to avoid leakage. The log scale above is for plotting only.

# 2.5 Final validation and save

**Problem:** the rubric requires before-and-after evidence for every major preprocessing decision.

```{r}
#| label: Before vs After Part 2

tibble(
  dataset = c("D01", "D02", "D03", "D08"),
  rows_raw = uncleaned_summary$rows,
  rows_clean = c(nrow(D01), nrow(D02), nrow(D03), nrow(D08)),
  cols_raw = uncleaned_summary$cols,
  cols_clean = c(ncol(D01), ncol(D02), ncol(D03), ncol(D08)),
  missing_values_raw = uncleaned_summary$missing_values,
  missing_values_clean = c(sum(is.na(D01)), sum(is.na(D02)), sum(is.na(D03)), sum(is.na(D08))),
  distinct_keys_raw = uncleaned_summary$distinct_keys,
  distinct_keys_clean = c(nrow(distinct(D01, household_id, person_number)),
                          nrow(distinct(D02, household_id, person_number)),
                          nrow(distinct(D03, household_id, person_number)),
                          n_distinct(D08$household_id)),
  exact_duplicates_raw = uncleaned_summary$exact_duplicate_rows,
  exact_duplicates_clean = c(nrow(D01) - nrow(distinct(D01)), nrow(D02) - nrow(distinct(D02)),
                             nrow(D03) - nrow(distinct(D03)), nrow(D08) - nrow(distinct(D08)))
)
```

**Note:** `missing_values_clean` is higher than `missing_values_raw` for D01–D03 because the clean datasets have more columns than the raw ones (new label and flag columns from 2.3 and Part 1), most of which are legitimately `NA` for rows that were never changed or that fall outside a structural category. This is not new data loss; the row-level missingness for the original analytical columns is covered in the Final Missingness table below.

```{r}
#| label: Types Before and After Part 2

bind_rows(
  D01 = enframe(D01_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D01, class), name = "column", value = "type_after"), by = "column"),
  D02 = enframe(D02_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D02, class), name = "column", value = "type_after"), by = "column"),
  D03 = enframe(D03_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D03, class), name = "column", value = "type_after"), by = "column"),
  D08 = enframe(D08_types_before, name = "column", value = "type_before") |>
    full_join(enframe(sapply(D08, class), name = "column", value = "type_after"), by = "column"),
  .id = "dataset"
) |>
  filter(is.na(type_before) != is.na(type_after) | (!is.na(type_before) & !is.na(type_after) & type_before != type_after))
```

```{r}
#| label: Final Missingness

bind_rows(
  D01 = enframe(colSums(is.na(D01)), name = "column", value = "missing"),
  D02 = enframe(colSums(is.na(D02)), name = "column", value = "missing"),
  D03 = enframe(colSums(is.na(D03)), name = "column", value = "missing"),
  D08 = enframe(colSums(is.na(D08)), name = "column", value = "missing"),
  .id = "dataset"
) |>
  filter(missing > 0) |>
  mutate(reason = case_when(
    column == "education_level" ~ "Codes 29/99 reclassified to NA in 2.1/2.3",
    column == "education_attendance_code" ~ "Genuinely missing since raw data; no dictionary code to assign (noted in 2.1)",
    column == "education_attendance" ~ "Code 3 reclassified to NA in 2.1/2.3, plus the pre-existing missing education_attendance_code values",
    column == "fixed_internet" ~ "Code 9 reclassified to NA in 2.1/2.3",
    column == "mobile_internet_code" ~ "Genuinely missing since raw data; no dictionary code to assign (noted in 2.1)",
    column == "mobile_internet" ~ "Code 9 reclassified to NA in 2.1/2.3, plus the pre-existing missing mobile_internet_code values",
    str_detect(column, "_flag$") ~ "Flags only filled in for rows that changed (expected)",
    column == "salary" ~ "Placeholder codes reversed to NA in 2.1",
    column == "salary_status" ~ "Could not be classified from value and employment status",
    TRUE ~ "NOT CLASSIFIED"
  ))
```

```{r}
#| label: Final Category Levels

list(
  province = levels(D01$province), sex = levels(D01$sex), population_group = levels(D01$population_group),
  geo_type = levels(D01$geo_type), metro = levels(D01$metro),
  education_level = levels(D02$education_level), education_attendance = levels(D02$education_attendance),
  institution_type = levels(D02$institution_type), reason_not_attending = levels(D02$reason_not_attending),
  employment_status = levels(D03$employment_status),
  fixed_internet = levels(D08$fixed_internet), mobile_internet = levels(D08$mobile_internet),
  computer_asset = levels(D08$computer_asset)
)
```

**Save:** both `.rds` (keeps factor levels and ID types) and `.csv` (required deliverable). Anyone reading the CSVs, including Power BI, must read `household_id` and `person_number` as text.

```{r}
#| label: Save Cleaned Datasets

saveRDS(D01, file.path(cleaned_directory, "D01_clean.rds"))
saveRDS(D02, file.path(cleaned_directory, "D02_clean.rds"))
saveRDS(D03, file.path(cleaned_directory, "D03_clean.rds"))
saveRDS(D08, file.path(cleaned_directory, "D08_clean.rds"))

write_csv(D01, file.path(cleaned_directory, "D01_clean.csv"))
write_csv(D02, file.path(cleaned_directory, "D02_clean.csv"))
write_csv(D03, file.path(cleaned_directory, "D03_clean.csv"))
write_csv(D08, file.path(cleaned_directory, "D08_clean.csv"))
```
