---
title: "EDA For Datasets"
author: "Group 5"
date: today
format:
  html: 
    toc: true #Gives a table of contents
    code-fold: show #Code chunks visible by default but can be collapsed
execute:
  warning: false #Makes warnings not appear to keep text clean in the file will appear in console when run
  message: false
---

## Install packages if they are not installed:

```{r}
#| label: install-packages
#install.packages(c("tidyverse", "skimr", "janitor", "visdat"))
```

## Loading the packages we need:

```{r}
#| label: package-setup
library(tidyverse)
library(skimr)
library(janitor)
library(visdat)
```

### This is the file path variable for the data set folder:

```{r}
#| label: data-directory
directory <- "../Datasets/"
```

## Creating the data frames for all the data sets:

```{r}
#| label: data-frame-creation
D01 <- read_csv(file.path(directory, "D01_Population_Profile.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names() #clean_names() is used to change all column names into a consistent form for example Household ID becomes household_id
D02 <- read_csv(file.path(directory, "D02_Learning_Profile.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D03 <- read_csv(file.path(directory, "D03_Economic_Participation.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D04 <- read_csv(file.path(directory, "D04_Dwelling_Conditions.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D05 <- read_csv(file.path(directory, "D05_Basic_Services.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D06 <- read_csv(file.path(directory, "D06_Household_Energy.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D07 <- read_csv(file.path(directory, "D07_Household_Wellbeing.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D08 <- read_csv(file.path(directory, "D08_Connectivity_Access.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D09 <- read_csv(file.path(directory, "D09_Household_Resources.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D10 <- read_csv(file.path(directory, "D10_Mobility_Access.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
D11 <- read_csv(file.path(directory, "D11_Labour_Market.csv"), show_col_types = FALSE, col_types = cols(HouseholdID = col_character())) |> clean_names()
```

### Finding the one or more parsing issues:

```{r}
#| label: find-issues
problems(D01)
problems(D02)
problems(D03)
problems(D04)
problems(D05)
problems(D06)
problems(D07)
problems(D08)
problems(D09)
problems(D10)
problems(D11)
```

#### We see that Data set 11 has a few problems in column 30, lets inspect what column it is:

```{r}
#| label: column-names
names(D11)[30] #Gives column name of col 30
nrow(problems(D11)) #Shows how many problems there are
```

> This tells us that 45 rows in the hours_worked_raw column has issues in data set 11 that need to be resolved in the pre-processing

## Quick summary about column and row counts for all datasets

```{r}
#| label: tibble-summary
#Tibble creates a small dataframe from scratch modern way of doing data.frame
tibble(
  dataset = c("D01","D02","D03","D04","D05","D06","D07","D08","D09","D10","D11"),
  rows = c(nrow(D01),nrow(D02),nrow(D03),nrow(D04),nrow(D05),nrow(D06),nrow(D07),nrow(D08),nrow(D09),nrow(D10),nrow(D11)),
  cols = c(ncol(D01),ncol(D02),ncol(D03),ncol(D04),ncol(D05),ncol(D06),ncol(D07),ncol(D08),ncol(D09),ncol(D10),ncol(D11))
)
```

> From this quick tibble summary we can see that D04 - D10 (All are around 20,000) are Household based, where D01 - D03 is person based (Around 70,000) and then we have D11 with around 170,000 which is also likely to do with people

## Exploring the data:

```{r}
#| label: exploring-D01
#View(D01)
head(D01,5)
skim(D01)
```

```{r}
#| label: exploring-D02
head(D02,5)
skim(D02)
```

```{r}
#| label: exploring-D03
head(D03,5)
skim(D03)
```

```{r}
#| label: exploring-D04
head(D04,5)
skim(D04)
```

```{r}
#| label: exploring-D05
head(D05,5)
skim(D05)
```

```{r}
#| label: exploring-D06
head(D06,5)
skim(D06)
```

```{r}
#| label: exploring-D07
head(D07,5)
skim(D07)
```

```{r}
#| label: exploring-D08
head(D08,5)
skim(D08)
```

```{r}
#| label: exploring-D09
head(D09,5)
skim(D09)
```

```{r}
#| label: exploring-D10
head(D10,5)
skim(D10)
```

```{r}
#| label: exploring-D11
head(D11,5)
skim(D11)
```

# Check For missing Data:

### Total missing values for each dataset

```{r}
#| label: total-missing-values
paste0("D01: ", sum(is.na(D01))) #paste0 combines the text with 0 spaces in between
paste0("D02: ", sum(is.na(D02)))
paste0("D03: ", sum(is.na(D03)))
paste0("D04: ", sum(is.na(D04)))
paste0("D05: ", sum(is.na(D05)))
paste0("D06: ", sum(is.na(D06)))
paste0("D07: ", sum(is.na(D07)))
paste0("D08: ", sum(is.na(D08)))
paste0("D09: ", sum(is.na(D09)))
paste0("D10: ", sum(is.na(D10)))
paste0("D11: ", sum(is.na(D11)))
```

> Dataset 11 has the most missing values with 205, with the person level datasets D01-D03 having around 100 and the rest of the datasets that are household level (D04-D10) all have around the 50-60 range.

### Diving deeper into which columns have the missing values in each dataset

```{r}
#| label: missing-Values-D01
data.frame(missing = colSums(is.na(D01)))
```

> There are 2 columns in Dataset 1 with missing values, (household_id = 25 missing values and age = 90 missing values)

```{r}
#| label: missing-Values-D02
data.frame(missing=colSums(is.na(D02)))
```

> There is 1 column in Dataset 2 with missing values (education_level_code = 120 missing values)

```{r}
#| label: missing-Values-D03
data.frame(missing=colSums(is.na(D03)))
```

> There is 1 column in Dataset 3 with missing values (employment_status1code = 100 missing values)

```{r}
#| label: missing-Values-D04
data.frame(missing = colSums(is.na(D04)))
```

> There is 1 column in Dataset 4 with missing values (wall_material_code = 55 missing values)

```{r}
#| label: missing-Values-D05
data.frame(missing=colSums(is.na(D05)))
```

> There is 1 column in Dataset 5 with missing values (drinking_water_source_code = 60 missing values)

```{r}
#| label: missing-Values-D06
data.frame(missing=colSums(is.na(D06)))
```

> There is 1 column in Dataset 6 with missing values (electricity_access_code = 50 missing values)

```{r}
#| label: missing-Values-D07
data.frame(missing=colSums(is.na(D07)))
```

> There is 1 column in Dataset 7 with missing values (adult_hunger_code = 65 missing values)

```{r}
#| label: missing-Values-D08
data.frame(missing=colSums(is.na(D08)))
```

> There is 1 column in Dataset 8 with missing values (mobile_internet_code = 60 missing values)

```{r}
#| label: missing-Values-D09
data.frame(missing=colSums(is.na(D09)))
```

> There is 1 column in Dataset 9 with missing values (main_income_source_code = 65 missing values)

```{r}
#| label: missing-Values-D10
data.frame(missing=colSums(is.na(D10)))
```

> There is 1 column in Dataset 10 with missing values (transport_to_health_facility_code = 55 missing values)

```{r}
#| label: missing-Values-D11
data.frame(missing=colSums(is.na(D11)))
```

> There are 2 columns in Dataset 11 with missing values (education_level_code = 160 missing values and hours_worked_raw = 45 missing values)

# Checking for duplicate Rows:

### Total duplicated values for each dataset

```{r}
#| label: total-duplicated-values
paste0("D01: ", D01 |>
  get_dupes(household_id,person_number) |>
  nrow())
paste0("D02: ",D02 |>
  get_dupes(household_id,person_number) |>
  nrow())
paste0("D03: ",D03 |>
  get_dupes(household_id,person_number) |>
  nrow())
paste0("D04: ",D04 |>
  get_dupes(household_id) |>
  nrow())
paste0("D05: ",D05 |>
  get_dupes(household_id) |>
  nrow())
paste0("D06: ",D06 |>
  get_dupes(household_id) |>
  nrow())
paste0("D07: ",D07 |>
  get_dupes(household_id) |>
  nrow())
paste0("D08: ",D08 |>
  get_dupes(household_id) |>
  nrow())
paste0("D09: ",D09 |>
  get_dupes(household_id) |>
  nrow())
paste0("D10: ",D10 |>
  get_dupes(household_id) |>
  nrow())
paste0("D11: ",D11 |>
  get_dupes() |>
  nrow())
```

> When we look at the dulicated values we see they are all below 100 which is good with most of them even below 50, the only above 50 is Dataset 11 and it has over 170000 rows which means that the data is relatively clean on duplicates.

### Diving deeper into the duplicated values in each dataset

```{r}
#| label: duplicate-Values-D01
D01 |> 
  get_dupes(household_id, person_number) #This is a compositie key we use on D01,D02,D03 since they have person number and household_id
```

> Data relatively clean with duplicates for Dataset 1 only duplicates occur due to NA for household_id

```{r}
#| label: duplicate-Values-D02
D02 |> 
  get_dupes(household_id, person_number)
```

> Dataset 2 has 44 rows of duplicate values

```{r}
#| label: duplicate-Values-D03
D03 |> 
  get_dupes(household_id, person_number)
```

> Dataset 3 has 40 rows of duplicate values

```{r}
#| label: duplicate-Values-D04
D04 |> 
  get_dupes(household_id)
```

> Dataset 4 has 32 rows of duplicate values

```{r}
#| label: duplicate-Values-D05
D05 |> 
  get_dupes(household_id)
```

> Dataset 5 has no duplicate values

```{r}
#| label: duplicate-Values-D06
D06 |> 
  get_dupes(household_id)
```

> Dataset 6 has no duplicate values

```{r}
#| label: duplicate-Values-D07
D07 |> 
  get_dupes(household_id)
```

> Dataset 6 has no duplicate values

```{r}
#| label: duplicate-Values-D08
D08 |> 
  get_dupes(household_id)
```

> No duplicate rows in Dataset 8

```{r}
#| label: duplicate-Values-D09
D09 |> 
  get_dupes(household_id)
```

> Dataset 9 has 30 rows of duplicate data

```{r}
#| label: duplicate-Values-D10
D10 |> 
  get_dupes(household_id)
```

> No duplicate rows in Dataset 10

```{r}
#| label: duplicate-Values-D11
D11 |> 
  get_dupes()
```

> Dataset 11 has 60 rows of duplicate data

## Which Datasets we choose:

> D01 - This dataset data about population information about people, for example (age,sex,province,population group) helps us link the other datasets to each other.
>
> D02 - This dataset provides data about education level, this is required to see the level of education attained when we compare it with connectivity level.
>
> D03 - This dataset covers employment data, which is required to see how connectivity compares with job. Chosen over D11 since the row count is similar with consistent id and person number.
>
> D08 - This dataset shows connectivity, this is required to see which households have internet or not it is the core way to determine if there is a link between education, employment and internet connectivity.

## Which Datasets we dont choose:

> D09 - Household resources
>
> D04 - Dwelling conditions
>
> D05 - Basic services
>
> D06 - Household energy
>
> D07 - Household wellbeing
>
> D10 - Mobility access
>
> D11 - Labour market
>
> D09 has an affect on whether a household has internet, but we are looking at how internet has an effect being able to get a job or education so for now data set 9 will be skipped but it will be reviewed later if we need more data.
>
> D04,D05,D06,D07,D10,D11 will not be selected for this scenario since they do not greatly affect whether a household has internet, and they also do not help us realize if internet has an affect on education or acquiring a job.

## Check intersect comparison to see if the datasets are compatible with each other:

> Now we need to confirm that the datasets we chose are related to each other and can be linked

```{r}
#| label: intersection-code
intersect(D01$household_id, D03$household_id) |> length()
intersect(D01$household_id, D02$household_id) |> length()
intersect(D01$household_id, D08$household_id) |> length()
intersect(D01$household_id, D11$household_id) |> length()
```

> We see that for household level, that D01,D02,D03,D08 are very closely related meaning they have the same household ids so they can be used for analysis and comparison

### Check the household ID matching %

```{r}
#| label: percent-of-household-ids-that-match
20684 / n_distinct(D01$household_id) * 100
```

> This tells us that 98.77% of the household ids have a match. This means that out of the total number of household ids in dataset 1, 98.77% of them match with the household ids in data set 2,3 and 8.

#### For person level checks we need to use composite keys so we first merge the household_id and person_id into a new id variable and then compare those values across datasets:

```{r}
#| label: person-level-ID-comparison
D01_id <- paste(D01$household_id, D01$person_number)
D02_id <- paste(D02$household_id, D02$person_number)
D03_id <- paste(D03$household_id, D03$person_number)

intersect(D01_id, D02_id) |> length()
intersect(D01_id, D03_id) |> length()
```

> We see that the values are very close to the total length of the datasets (70000 and change) meaning alot of the composite keys match so we can use this data for comparisons and analysis
>
> D08 is house hold level so we can not compare it at the person level with the others.

# Analysis of our chosen datasets:

## Check for Empty string columns

```{r}
#| label: check-for-empty-strings
D01 |>
  summarise(across(where(is.character), ~sum(.x == "", na.rm=TRUE)))
D02 |>
  summarise(across(where(is.character), ~sum(.x == "", na.rm=TRUE)))
D03 |>
  summarise(across(where(is.character), ~sum(.x == "", na.rm=TRUE)))
D08 |>
  summarise(across(where(is.character), ~sum(.x == "", na.rm=TRUE)))
```

> We see that there are no empty strings in any of our selected datasets

## Check for Negative values in age

```{r}
#| label: check-negative-ages
paste0("D01: ", sum(D01$age < 0, na.rm = TRUE))#na.rm means should NA values be counted or not TRUE means skip we dont want it messing with our values so we remove it
paste0("D02: ", sum(D02$age < 0, na.rm = TRUE))
paste0("D03: ", sum(D03$age < 0, na.rm = TRUE))
```

```{r}
#| label: which-rownr-negative-ages
which(D01$age < 0) #Which is used to see which row numbers are below 0 for age
```

```{r}
#| label: filter dataset-1-age
D01 |>
  filter(age < 0)
```

## Check for invalid values in the columns

### Dataset 1:

```{r}
#| label: dataset-1-valid-values-check
paste0("province_code: ", sum(!D01$province_code %in% c(1,2,3,4,5,6,7,8,9), na.rm = TRUE))

paste0("sex_code: ", sum(!D01$sex_code %in% c(1,2), na.rm = TRUE))

paste0("age_group_code: ", sum(!D01$age_group_code %in% c(01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16), na.rm =TRUE))

paste0("population_group_code: ", sum(!D01$population_group_code %in% c(1,2,3,4,5), na.rm = TRUE))

paste0("language_code: ", sum(!D01$language_code %in% c(01,02,03,04,05,06,07,08,09,10,11,12,13,14), na.rm = TRUE))

paste0("relationship_to_head_code: ", sum(!D01$relationship_to_head_code %in% c(1,2,3,4,5,6,7,8,9,99), na.rm = TRUE))

paste0("marital_status_code: ", sum(!D01$marital_status_code %in% c(1,2,3,4,5,6,7,8,9), na.rm = TRUE))


paste0("geo_type_code: ", sum(!D01$geo_type_code %in% c(1,2,3), na.rm = TRUE))

paste0("metro_code: ", sum(!D01$metro_code %in% c(01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17), na.rm = TRUE))

paste0("metro_indicator: ", sum(!D01$metro_indicator %in% c(1,2), na.rm = TRUE))
```

### Dataset 2:

```{r}
#| label: dataset-2-valid-values-check
paste0("province_code: ", sum(!D02$province_code %in% c(1,2,3,4,5,6,7,8,9), na.rm = TRUE))

paste0("sex_code: ", sum(!D02$sex_code %in% c(1,2), na.rm = TRUE))

paste0("education_level_code: ", sum(!D02$education_level_code %in% c(98,00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29), na.rm = TRUE))

paste0("education_attendance_code: ", sum(!D02$education_attendance_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("reason_not_attending_code: ", sum(!D02$reason_not_attending_code %in% c(01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17,18,88), na.rm = TRUE))

paste0("institution_type_code: ", sum(!D02$institution_type_code %in% c(1,2,3,4,5,6,7,8,9,88), na.rm = TRUE))

paste0("public_private_institution_code: ", sum(!D02$public_private_institution_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("transport_mode_to_education_code: ", sum(!D02$transport_mode_to_education_code %in% c(1,2,3,4,5,6,7,8,9,10,88), na.rm = TRUE))

paste0("travel_time_to_education: ", sum(!D02$travel_time_to_education %in% c(1,2,3,4,5,6,7,8), na.rm = TRUE))

paste0("total_education_fees: ", sum(!D02$total_education_fees %in% c(00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,88), na.rm = TRUE))

paste0("no_fees_school_code: ", sum(!D02$no_fees_school_code %in% c(1,2,3,4,5,6,7,8), na.rm = TRUE))

paste0("grade_code: ", sum(!D02$grade_code %in% c(00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,88), na.rm = TRUE))

paste0("school_food_programme_code: ", sum(!D02$school_food_programme_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("absence_code: ", sum(!D02$absence_code %in% c(1,2,3,4,8), na.rm = TRUE))

paste0("absence_days: ", sum(!D02$absence_days %in% c(1,2,3,4,5,8), na.rm = TRUE))

paste0("geo_type_code: ", sum(!D02$geo_type_code %in% c(1,2,3), na.rm = TRUE))
```

#### Exploring the invalid values we have in columns

```{r}
#| label: dataset-2-explore-invalid-values
D02 |> 
  filter(!D02$education_level_code %in% c(98,00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29))
D02 |> 
  filter(!D02$education_attendance_code %in% c(1,2,3,8))
```

### Dataset 3:

```{r}
#| label: dataset-3-valid-values-check
paste0("province_code: ", sum(!D03$province_code %in% c(1,2,3,4,5,6,7,8,9), na.rm = TRUE))

paste0("sex_code: ", sum(!D03$sex_code %in% c(1,2), na.rm = TRUE))

paste0("education_level_code: ", sum(!D03$education_level_code %in% c(98,00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29), na.rm = TRUE))

paste0("wage_work_code: ", sum(!D03$wage_work_code %in% c(1,2,8), na.rm = TRUE))

paste0("business_work_code: ", sum(!D03$business_work_code %in% c(1,2,8), na.rm = TRUE))

paste0("volunteer_work_code: ", sum(!D03$volunteer_work_code %in% c(1,2,8), na.rm = TRUE))

paste0("retired_code: ", sum(!D03$retired_code %in% c(1,2,8), na.rm = TRUE))

paste0("salary_amount_raw: ", sum(D03$salary_amount_raw < 0, na.rm = TRUE))

paste0("salary_period_code: ", sum(!D03$salary_period_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("salary_category_code: ", sum(!D03$salary_category_code %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,88), na.rm = TRUE))

paste0("transport_to_work_code: ", sum(!D03$transport_to_work_code %in% c(1,2,3,4,5,6,7,8,9,10,11,88), na.rm = TRUE))

paste0("minutes_to_work: ", sum(!D03$minutes_to_work %in% c(1,2,3,4,5,6,8), na.rm = TRUE))

paste0("sector_code: ", sum(!D03$sector_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("looking_for_work_code: ", sum(!D03$looking_for_work_code %in% c(1,2,8), na.rm = TRUE))

paste0("could_start_work_code: ", sum(!D03$could_start_work_code %in% c(1,2,3,8), na.rm = TRUE))

paste0("employment_status1code: ", sum(!D03$employment_status1code %in% c(1,2,8,9), na.rm = TRUE))

paste0("employment_status2code: ", sum(!D03$employment_status2code %in% c(1,2,3,8), na.rm = TRUE))

paste0("lab_salary_raw: ", sum(D03$lab_salary_raw < 0, na.rm = TRUE))
```

```{r}
#| label: dataset-3-explore-invalid-values
D03 |> 
  filter(!D03$education_level_code %in% c(98,00,01,02,03,04,05,06,07,08,09,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29))
D03 |>
  filter(!D03$wage_work_code %in% c(1,2,8))
D03 |>
  filter(!D03$business_work_code %in% c(1,2,8))
D03 |>
  filter(!D03$volunteer_work_code %in% c(1,2,8))
D03 |>
  filter(!D03$minutes_to_work %in% c(1,2,3,4,5,6,8))
D03 |>
  filter(!D03$employment_status1code %in% c(1,2,8,9))
```

### Dataset 8:

```{r}
#| label: dataset-8-valid-values-check
paste0("province_code: ", sum(!D08$province_code %in% c(1,2,3,4,5,6,7,8,9), na.rm = TRUE))

paste0("telephone_code: ", sum(!D08$telephone_code %in% c(1,2), na.rm = TRUE))

paste0("cellphone_code: ", sum(!D08$cellphone_code %in% c(1,2), na.rm = TRUE))

paste0("fixed_internet_code: ", sum(!D08$fixed_internet_code %in% c(1,2), na.rm = TRUE))

paste0("mobile_internet_code: ", sum(!D08$mobile_internet_code %in% c(1,2), na.rm = TRUE))

paste0("public_wi_fi_internet_code: ", sum(!D08$public_wi_fi_internet_code %in% c(1,2), na.rm = TRUE))

paste0("study_internet_code: ", sum(!D08$study_internet_code %in% c(1,2), na.rm = TRUE))

paste0("work_internet_code: ", sum(!D08$work_internet_code %in% c(1,2), na.rm = TRUE))

paste0("library_internet_code: ", sum(!D08$library_internet_code %in% c(1,2), na.rm = TRUE))

paste0("cafe_internet_code: ", sum(!D08$cafe_internet_code %in% c(1,2), na.rm = TRUE))

paste0("household_mobile_phones: ", sum(!D08$household_mobile_phones %in% c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15), na.rm = TRUE))

paste0("household_smartphones: ", sum(!D08$household_smartphones %in% c(0,1,2,3,4,5,6,7,8,9,10,11,12,13), na.rm = TRUE))

paste0("computer_asset_code: ", sum(!D08$computer_asset_code %in% c(1,2,9), na.rm = TRUE))

paste0("geo_type_code: ", sum(!D08$geo_type_code %in% c(1,2,3), na.rm = TRUE))
```

```{r}
#| label: dataset-8-explore-invalid-values
D08 |>
  filter(!D08$telephone_code %in% c(1,2))

D08 |>
  filter(!D08$fixed_internet_code %in% c(1,2))

D08 |>
  filter(!D08$mobile_internet_code %in% c(1,2))

D08 |>
  filter(!D08$public_wi_fi_internet_code %in% c(1,2))

D08 |>
  filter(!D08$study_internet_code %in% c(1,2))

D08 |>
  filter(!D08$work_internet_code %in% c(1,2))

D08 |>
  filter(!D08$library_internet_code %in% c(1,2))

D08 |>
  filter(!D08$cafe_internet_code %in% c(1,2))

D08 |>
  filter(!D08$household_smartphones %in% c(0,1,2,3,4,5,6,7,8,9,10,11,12,13))
```

# Graphs for our chosen datasets:

## Dataset 1 Graphs:

```{r}
#| label: dataset-1-graphs
boxplot(D01$age)
hist(D01$age)
```

For age we can see there are a couple of outliers at about 90, nothing to major though, we can also see that most people are around the age of 20, we also see some values below 0 that need to be dropped alongside the outliers

```{r}
summary(D01$age)
```

After a bit further investigation we see that the average age is 30 and that the majority of people are aged 28

```{r}
D01 |>
  ggplot(aes(x=province)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Province Distribution",
       x = "Province",
       y = "Count") +
  theme(axis.text.x = element_text(angle=45, hjust = 1))
```

This graph tells us that the spelling for all the different provinces are different.

```{r}
unique(D01$province)
```

By running unique province we prove this and we see there is a naming convention problem

```{r}
D01 <- D01 |> mutate(province = str_to_title(province))
```

To fix it we mutate the province with str_to_title to normalize the capitalization of all the words

```{r}
unique(D01$province)
```

```{r}
D01 |>
  ggplot(aes(x=province)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Province Distribution",
       x = "Province",
       y = "Count") +
  theme(axis.text.x = element_text(angle=45, hjust = 1))
```

Our graph now looks nice, and we can see majority of the people live in Gauteng.

## Dataset 2 Graphs:

```{r}
D02 |>
  ggplot(aes(D02$education_level_code)) + 
  geom_histogram(bins = 50, binwidth = 1, fill = "steelblue", color="white") +
  labs(title = "Education Distribution",
       x = "Education Code",
       y = "count")
```

The rows the graphs has removed are the NA values that need to be dropped in the preprocessing stage.

From this graph however we can see that majority lies at around Code 10

```{r}
education_counts <- table(D02$education_level_code)
paste0("Most common code: ", names(which.max(education_counts)))
paste0("Frequency of the code: ", max(education_counts))
```

```{r}
data.frame(education_counts) |>
  arrange(desc(Freq))
```

After further exploration we see that code 12 is most common education which is Grade 12, with code 98 close behind it with 9378 records which means no schooling

## Dataset 3 Graphs:

```{r}

```

## Dataset 4 Graphs:

```{r}

```

# What to handle in preprocessing:

# Initial Hypotheses:
