---
title: "EDA For Datasets"
author: "Group 5"
date: today
format:
  html: 
    toc: true #Gives a table of contents
    code-fold: show #Code chunks visible by default but can be collapsed
execute:
  warning: false #Makes warnings not appear to keep text clean in the file will appear in console when run
  message: false
---

## Install packages if they are not installed:

```{r}
#| label: Install Packages
#install.packages(c("tidyverse", "skimr", "janitor", "visdat"))
```

## Loading the packages we need:

```{r}
#| label: package setup
library(tidyverse)
library(skimr)
library(janitor)
library(visdat)
```

### This is the file path variable for the data set folder:

```{r}
#| label: data Directory
directory <- "../Datasets/"
```

## Creating the data frames for all the data sets:

```{r}
#| label: Data Frame Creation
D01 <- read_csv(file.path(directory,"D01_Population_Profile.csv"),show_col_types = FALSE) |> clean_names() #clean_names() is used to change all column names into a consistent form for example Household ID becomes household_id
D02 <- read_csv(file.path(directory,"D02_Learning_Profile.csv"),show_col_types = FALSE) |> clean_names()
D03 <- read_csv(file.path(directory,"D03_Economic_Participation.csv"),show_col_types = FALSE) |> clean_names()
D04 <- read_csv(file.path(directory,"D04_Dwelling_Conditions.csv"),show_col_types = FALSE) |> clean_names()
D05 <- read_csv(file.path(directory,"D05_Basic_Services.csv"),show_col_types = FALSE) |> clean_names()
D06 <- read_csv(file.path(directory,"D06_Household_Energy.csv"),show_col_types = FALSE) |> clean_names()
D07 <- read_csv(file.path(directory,"D07_Household_Wellbeing.csv"),show_col_types = FALSE) |> clean_names()
D08 <- read_csv(file.path(directory,"D08_Connectivity_Access.csv"),show_col_types = FALSE) |> clean_names()
D09 <- read_csv(file.path(directory,"D09_Household_Resources.csv"),show_col_types = FALSE) |> clean_names()
D10 <- read_csv(file.path(directory,"D10_Mobility_Access.csv"),show_col_types = FALSE) |> clean_names()
D11 <- read_csv(file.path(directory,"D11_Labour_Market.csv"),show_col_types = FALSE) |> clean_names()
```

### Finding the one or more parsing issues:

```{r}
#| label: find issues
problems(D01)
problems(D02)
problems(D03)
problems(D04)
problems(D05)
problems(D06)
problems(D07)
problems(D08)
problems(D09)
problems(D10)
problems(D11)
```

#### We see that Data set 11 has a few problems in column 30, lets inspect what column it is:

```{r}
#| label: column names
names(D11)[30] #Gives column name of col 30
nrow(problems(D11)) #Shows how many problems there are
```

> This tells us that 45 rows in the hours_worked_raw column has issues in data set 11 that need to be resolved in the pre-processing

## Quick summary about column and row counts for all datasets

```{r}
#| label: tibble summary
#Tibble creates a small dataframe from scratch from 3 vectors that are manually defined, used to show the dataframe name, col count and row count
tibble(
  dataset = c("D01","D02","D03","D04","D05","D06","D07","D08","D09","D10","D11"),
  rows = c(nrow(D01),nrow(D02),nrow(D03),nrow(D04),nrow(D05),nrow(D06),nrow(D07),nrow(D08),nrow(D09),nrow(D10),nrow(D11)),
  cols = c(ncol(D01),ncol(D02),ncol(D03),ncol(D04),ncol(D05),ncol(D06),ncol(D07),ncol(D08),ncol(D09),ncol(D10),ncol(D11))
)
```

> From this quick tibble summary we can see that D04 - D10 (All are around 20,000) are Household based, where D01 - D03 is person based (Around 70,000) and then we have D11 with around 170,000 which is also likely to do with people

## Exploring the data:

```{r}
#| label: exploring D01
#View(D01)
head(D01,5)
skim(D01)
```

```{r}
#| label: exploring D02
head(D02,5)
skim(D02)
```

```{r}
#| label: exploring D03
head(D03,5)
skim(D03)
```

```{r}
#| label: exploring D04
head(D04,5)
skim(D04)
```

```{r}
#| label: exploring D05
head(D05,5)
skim(D05)
```

```{r}
#| label: exploring D06
head(D06,5)
skim(D06)
```

```{r}
#| label: exploring D07
head(D07,5)
skim(D07)
```

```{r}
#| label: exploring D08
head(D08,5)
skim(D08)
```

```{r}
#| label: exploring D09
head(D09,5)
skim(D09)
```

```{r}
#| label: exploring D10
head(D10,5)
skim(D10)
```

```{r}
#| label: exploring D11
head(D11,5)
skim(D11)
```

## Which Datasets we choose:

D01 - This dataset data about population information about people, for example (age,sex,province,population group) helps us link the other datasets to each other.

D02 - This dataset provides data about education level, this is required to see the level of education attained when we compare it with connectivity level.

D03 - This dataset covers employment data, which is required to see how connectivity compares with job. Chosen over D11 since the row count is similar with consistent id and person number.

D08 - This dataset shows connectivity, this is required to see which households have internet or not it is the core way to determine if there is a link between education, employment and internet connectivity.

## Which Datasets we dont choose: (Still needs descriptions as to why not chosen)

D04 - Dwelling conditions (Could affect whether they have internet or not, maybe we need it?)

D05 - Basic services

D06 - Household energy

D07 - Household wellbeing

D09 - Household resources

D10 - Mobility access

D11 - Labour market

### 

## Check intersect comparison to see if the datasets are compatible with each other:

#### Now we need to confirm that the datasets we chose are related to each other and can be linked

```{r}
intersect(D01$household_id, D03$household_id) |> length()
intersect(D01$household_id, D02$household_id) |> length()
intersect(D01$household_id, D08$household_id) |> length()
intersect(D01$household_id, D11$household_id) |> length()
```

We see that for household level, that D01,D02,D03,D08 are very closely related meaning they have the same household ids so they can be used for analysis and comparison

```{r}
20684 / n_distinct(D01$household_id) * 100
```

This tells us that 99% of the household ids have a match

#### For person level checks we need to use composite keys so we first merge the household_id and person_id into a new id variable and then compare those values across datasets:

```{r}
D01_id <- paste(D01$household_id, D01$person_number)
D02_id <- paste(D02$household_id, D02$person_number)
D03_id <- paste(D03$household_id, D03$person_number)

intersect(D01_id, D02_id) |> length()
intersect(D01_id, D03_id) |> length()
```

We see that the values are very close to the total length of the datasets (70000 and change) meaning alot of the composite keys match so we can use this data for comparisons and analysis

D08 is house hold level so we can not compare it at the person level with the others.
