---
title: "Train/Test Split"
author: "Group 5"
date: today
format:
  html:
    toc: true
    code-fold: show
    df-print: kable
    embed-resources: true
  pdf:
    toc: true
    df-print: kable
execute:
  warning: false
  message: false
---


## Purpose

This covers the Milestone 2 rubric item "Modelling preparation & leakage control": defining the modelling target, designing a train/test split, and preventing leakage. The target, predictors and leakage exclusions are already decided and documented in `Data Preprocessing.qmd` (sections 10.4–10.6) and `Cleaned Datasets Data Dictionary.md` (sections 11–13). This file only adds the train/test split for `model_data.csv` and `income_data.csv`.

`analytical_dataset.csv` is not split here: it is the Power BI/reporting file, described in the data dictionary as containing every working-age person for description rather than modelling.

## Loading the packages we need:

```{r}
#| label: Package Setup

library(readr)
library(tidyverse)
```

## File paths:

```{r}
#| label: File-Path Declarations

analytical_directory <- "../Datasets/Analytical/"
```

# 1. Read the analytical datasets

**Problem:** the split must start from the saved files, not from objects left in memory from `Data Preprocessing.qmd`, so this file can be run on its own.

```{r}
#| label: Read Analytical Datasets

model_data <- read_csv(file.path(analytical_directory, "model_data.csv"), show_col_types = FALSE,
                       col_types = cols(household_id = col_character()))

income_data <- read_csv(file.path(analytical_directory, "income_data.csv"), show_col_types = FALSE,
                        col_types = cols(household_id = col_character()))
```

**Re-apply the reference levels** lost when the factors were written to CSV (per `Cleaned Datasets Data Dictionary.md`, section 2.5):

```{r}
#| label: Re-apply Reference Levels

model_data <- model_data |>
  mutate(
    employed = factor(employed, levels = c("Unemployed", "Employed")),
    education_band = fct_relevel(education_band, "Some secondary"),
    province = fct_relevel(province, "Gauteng"),
    population_group = fct_relevel(population_group, "Black African"),
    settlement_type = fct_relevel(settlement_type, "Metro urban"),
    edu_digital_group = fct_relevel(edu_digital_group, "Below matric, no internet"),
    smartphone_access = fct_relevel(smartphone_access, "None"),
    other_member_matric = fct_relevel(other_member_matric, "No")
  )

income_data <- income_data |>
  mutate(
    education_band = factor(education_band),
    population_group = factor(population_group),
    province = factor(province),
    settlement_type = factor(settlement_type)
  )
```

# 2. Target and class balance

**Problem:** the rubric asks for the modelling target to be defined and class imbalance assessed before splitting.

`model_data.csv`'s target is `employed` (Employed vs Unemployed); `income_data.csv`'s outcome is `log_salary` (continuous, no class balance to check).

```{r}
#| label: Target Definition and Class Balance

model_data |> count(employed) |> mutate(share = n / sum(n))
```

**Effect:** `r sum(model_data$employed == "Employed")` employed (`r round(mean(model_data$employed == "Employed") * 100, 1)`%) vs `r sum(model_data$employed == "Unemployed")` unemployed. This is a mild imbalance (roughly 2:1), not severe enough to need resampling (e.g. SMOTE or down-sampling); `glm()`, `rpart()` and `randomForest()` can all fit this directly. Milestone 3 can still compare a class-weighted model against an unweighted one if the minority (Unemployed) class needs better recall.

# 3. Household-grouped train/test split

**Problem:** rows in both files are people, and people in the same household share household-level predictors (`any_home_internet`, `digital_access_score`, `has_fixed_internet`, and so on, per section 10 of `Data Preprocessing.qmd`). A row-level random split can put two members of the same household in both the training and the test set. The model would then see that household's digital-access profile during training and be tested on a near-duplicate of it, which overstates test performance. Splitting by `household_id` instead keeps every member of a household on the same side of the split.

**Treatment:** sample 80% of the *distinct households* in each file for training; every person in a household follows their household's assignment.

```{r}
#| label: Household-Grouped Split

set.seed(67) #Fixed seed so the split is reproducible

split_by_household <- function(data, train_share = 0.8) {
  households <- unique(data$household_id)
  train_households <- sample(households, size = round(train_share * length(households)))
  list(
    train = data |> filter(household_id %in% train_households),
    test = data |> filter(!household_id %in% train_households)
  )
}

model_split <- split_by_household(model_data)
model_train <- model_split$train
model_test <- model_split$test

income_split <- split_by_household(income_data)
income_train <- income_split$train
income_test <- income_split$test
```

**Verification:** every household is on exactly one side, the two sides don't overlap, and the target's class balance is close to the full dataset in both the training and test rows (a household-grouped split can shift the balance slightly, since it doesn't sample people directly).

```{r}
#| label: Split Verification

tibble(
  dataset = c("model_data", "income_data"),

  households_total = c(n_distinct(model_data$household_id), n_distinct(income_data$household_id)),

  households_train = c(n_distinct(model_train$household_id), n_distinct(income_train$household_id)),

  households_test = c(n_distinct(model_test$household_id), n_distinct(income_test$household_id)),

  household_overlap = c(length(intersect(model_train$household_id, model_test$household_id)),
                        length(intersect(income_train$household_id, income_test$household_id))),

  rows_train = c(nrow(model_train), nrow(income_train)),

  rows_test = c(nrow(model_test), nrow(income_test)),

  rows_total_matches = c(nrow(model_train) + nrow(model_test) == nrow(model_data),
                         nrow(income_train) + nrow(income_test) == nrow(income_data))
)

model_train |> count(employed) |> mutate(share = round(n / sum(n), 3))
model_test |> count(employed) |> mutate(share = round(n / sum(n), 3))
```

**Effect:** `r n_distinct(model_train$household_id)` households (`r nrow(model_train)` people) in the `model_data` training set and `r n_distinct(model_test$household_id)` households (`r nrow(model_test)` people) in the test set, with zero household overlap. The employed/unemployed split stays close to the full dataset's `r round(mean(model_data$employed == "Employed") * 100, 1)`% in both the training and test rows.

# 4. Leakage check on the split itself

**Problem:** the rubric specifically asks for leakage prevention to be demonstrated, not just stated. The column-level leakage exclusions (salary, salary-derived and reason-not-attending columns) are already handled in `Data Preprocessing.qmd` section 10.4; this section checks that the *split* introduces no further leakage.

```{r}
#| label: Leakage Check

tibble(
  check = c("No household appears in both model train and test", "No household appears in both income train and test",
           "model_data columns unchanged by the split", "income_data columns unchanged by the split"),

  result = c(length(intersect(model_train$household_id, model_test$household_id)) == 0,
            length(intersect(income_train$household_id, income_test$household_id)) == 0,
            identical(names(model_train), names(model_data)) && identical(names(model_test), names(model_data)),
            identical(names(income_train), names(income_data)) && identical(names(income_test), names(income_data)))
)
```

**Effect:** no household appears on both sides of either split, and splitting adds no new columns and removes none — `person_weight` and `household_id` stay in place as design columns (not predictors, per section 11.1/12 of the data dictionary), so Milestone 3 can still fit the survey weights and identify each row's household without either being used as a predictor.

No further transformation is applied here: scaling, encoding and the fixed centring of age were already done in `Data Preprocessing.qmd` before this split using only fixed rules or constants (section 10.4), not statistics estimated from the data, so applying them before or after this split makes no difference. Any future step that does estimate something from the data (for example, imputing `matric_plus`'s 497 missing values, or scaling a numeric predictor by its mean and standard deviation) belongs in Milestone 3, fitted on `model_train`/`income_train` only and then applied unchanged to the corresponding test set.

# 5. Save the split datasets

```{r}
#| label: Save Split Datasets

write_csv(model_train, file.path(analytical_directory, "model_train.csv"), na = "")
write_csv(model_test, file.path(analytical_directory, "model_test.csv"), na = "")

write_csv(income_train, file.path(analytical_directory, "income_train.csv"), na = "")
write_csv(income_test, file.path(analytical_directory, "income_test.csv"), na = "")
```

**Verification:**

```{r}
#| label: Saved Split Verification

model_train_saved <- read_csv(file.path(analytical_directory, "model_train.csv"), show_col_types = FALSE,
                              col_types = cols(household_id = col_character()))
model_test_saved <- read_csv(file.path(analytical_directory, "model_test.csv"), show_col_types = FALSE,
                             col_types = cols(household_id = col_character()))
income_train_saved <- read_csv(file.path(analytical_directory, "income_train.csv"), show_col_types = FALSE,
                               col_types = cols(household_id = col_character()))
income_test_saved <- read_csv(file.path(analytical_directory, "income_test.csv"), show_col_types = FALSE,
                              col_types = cols(household_id = col_character()))

tibble(
  file = c("model_train", "model_test", "income_train", "income_test"),
  rows_in_memory = c(nrow(model_train), nrow(model_test), nrow(income_train), nrow(income_test)),
  rows_in_saved_file = c(nrow(model_train_saved), nrow(model_test_saved), nrow(income_train_saved), nrow(income_test_saved))
)
```

**Effect:** four new files in `Datasets/Analytical/`: `model_train.csv`, `model_test.csv`, `income_train.csv`, `income_test.csv`. `model_data.csv`, `income_data.csv` and `analytical_dataset.csv` are unchanged. As with the other saved files, re-apply the reference levels in section 1 after reading these back in Milestone 3.

